---
title: "Untitled"
output: html_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(stringr)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```

## Payouts

First determine payouts for given inputs (year, grid, interval, coverage level, productivity factor, grazing/haying, )

Assumptions made:
- constant productivity factor of 1
- rates are constant across years
- rates are consistent despite changing temporal baseline or spatial resolution
- acreage proportionally split among cells in county
- liability, premia, payouts calculated based on acreage for each interval

Equations used:
Dollar Amount of Insurance Per Acre = County-Base Value * Productivity Factor (also = Liability per Acre?)
Liability = Dollar Amount * # Acres
Premium per acre = Liability per acre * rate 
Payment factor = (trigger_threshold - RI)/trigger_threshold
Indemnity = Liability * payment factor 

Columns to add
CHIRPS data set
- rates, 
- payment factor for each interval x year


## load rates data, join to chirps_data

```{r}

load(file = paste0(here("data/"), "chirps_data_TX_region.rda"))

years <- 2000:2023
suffixes <- 625:635

# Build pattern
year_pattern <- paste0(years, collapse = "|")
suffix_pattern <- paste0(suffixes, collapse = "|")
regex_pattern <- paste0("^cpc_index_(", year_pattern, ")_(", suffix_pattern, ")$")

cpc_data <- chirps_data %>% select(GRIDCODE, lon, lat, region, matches(regex_pattern))



```




## load grids
```{r}

## for CPC grids
grids <- st_read(here("data/rainfall_index_grids/official_RMA_RI_grid.shp"))
grids$lon <- rowMeans(cbind(grids$X_MIN, grids$X_MAX))
grids$lat <- rowMeans(cbind(grids$Y_MIN, grids$Y_MAX))

grids <- cpc_data %>%
  left_join(grids %>% select(-c(lon, lat)), by = "GRIDCODE") %>%
  st_as_sf()

rm(cpc_data, chirps_data)

```

Unzip all USDM .zip files

```{r}
unzipped <- TRUE

if(unzipped == FALSE){
    # Set root directory where all the year folders are located
  root_dir <- here("data/USDM_GIS/")
  
  # Get all subfolders matching the pattern
  folders <- list.dirs(root_dir, recursive = FALSE, full.names = TRUE)
  target_folders <- folders[grepl("_USDM_M", basename(folders))]
  print(target_folders)
  # Loop through each folder
  for (folder in target_folders) {
    # List all zip files in the folder
    zip_files <- list.files(folder, pattern = "\\.zip$", full.names = TRUE)
    
    # Extract each zip file to the same folder
    for (zip in zip_files) {
      unzip(zip, exdir = folder)
    }
  }
}


```




Load individual USDM shapefiles, extract values to TX grids
```{r}

sf::sf_use_s2(FALSE)

grids_pt <- grids %>% st_drop_geometry() %>% st_as_sf(., coords = c("lon", "lat"), crs = 4326)



for (year in years) {
  print(paste0("processing year ", year))
  # Get list of all shapefiles for the year
  year_shps <- list.files(here(paste0("data/USDM_GIS/", year, "_USDM_M")),
                         pattern = "\\.shp$", full.names = TRUE)
  
  for (shp in year_shps) {
    # Extract date from file name using regex (expects format like USDM_YYYYMMDD.shp)
    usdm_date <- str_extract(basename(shp), "\\d{8}")
    print(usdm_date)
    # Read shapefile
    a <- st_read(shp, quiet = TRUE)
    
    # Prepare grids points (assuming grids is available in environment)
    grids_pt <- grids %>% 
      st_drop_geometry() %>% 
      st_as_sf(coords = c("lon", "lat"), crs = 4326)
    
    # Spatial join and keep only DM column renamed with date
    grids_joined <- st_join(grids_pt, a, left = TRUE) %>%
      mutate(DM = coalesce(DM, -1)) %>%
      select(DM) %>%
      rename(!!paste0("DM_", usdm_date) := DM)
    
    grids[[paste0("DM_", usdm_date)]]  <- grids_joined[[paste0("DM_", usdm_date)]]
  }
}

save(grids, file = here("data/grids_usdm"))

rm(grids_joined, grids_pt)

```

## calculate derived USDM stats

```{r}

load(file = here("data/grids_usdm"))

## long drought >= 3 for 8+ weeks

min_severity <- 3
min_duration <- 8


# Step 1: Identify and sort all DM columns
dm_cols <- names(grids)[str_detect(names(grids), "^DM_\\d{8}$")]
dm_dates <- str_remove(dm_cols, "DM_")
dm_cols_sorted <- dm_cols[order(dm_dates)]

# Step 2: Subset the DM matrix
dm_matrix <- as.matrix(grids[, dm_cols_sorted] %>% st_drop_geometry())

# Step 3: Create empty matrix for LD results
ld_matrix <- matrix(0, nrow = nrow(dm_matrix), ncol = ncol(dm_matrix))

# Step 4: Identify sequences of 8+ consecutive weeks where DM >= 3
for (i in seq_len(nrow(dm_matrix))) {
  row_vals <- dm_matrix[i, ]
  above_thresh <- row_vals >= min_severity
  rle_res <- rle(above_thresh)
  
  # Get positions in the row that are part of valid sequences
  pos <- rep(rle_res$lengths >= min_duration & rle_res$values, rle_res$lengths)
  
  ld_matrix[i, pos] <- 1
}

num_gt0 <- sum(ld_matrix > 0, na.rm = TRUE)              # number of cells > 0
pct_gt0 <- mean(ld_matrix > 0, na.rm = TRUE) * 100  

# Step 5: Name columns like LD_YYYYMMDD
ld_colnames <- str_replace(dm_cols_sorted, "DM_", "LD3_")
colnames(ld_matrix) <- ld_colnames

print(paste0("# of dates with at least one grid >= ", min_severity, " for ", min_duration, "+ weeks: ", sum(colSums(ld_matrix) > 0), " out of ", dim(ld_matrix)[2], " total weeks"))

# Step 6: Bind to grids
grids <- bind_cols(grids, as.data.frame(ld_matrix))

rm(ld_matrix, dm_matrix, rle_res, row_vals)


```


```{r}

## long drought >= 4 for 8+ weeks

min_severity <- 4
min_duration <- 8

# Step 1: Identify and sort all DM columns
dm_cols <- names(grids)[str_detect(names(grids), "^DM_\\d{8}$")]
dm_dates <- str_remove(dm_cols, "DM_")
dm_cols_sorted <- dm_cols[order(dm_dates)]

# Step 2: Subset the DM matrix
dm_matrix <- as.matrix(grids[, dm_cols_sorted] %>% st_drop_geometry())

# Step 3: Create empty matrix for LD results
ld_matrix <- matrix(0, nrow = nrow(dm_matrix), ncol = ncol(dm_matrix))

# Step 4: Identify sequences of 8+ consecutive weeks where DM >= 3
for (i in seq_len(nrow(dm_matrix))) {
  row_vals <- dm_matrix[i, ]
  above_thresh <- row_vals >= min_severity
  rle_res <- rle(above_thresh)
  
  # Get positions in the row that are part of valid sequences
  pos <- rep(rle_res$lengths >= min_duration & rle_res$values, rle_res$lengths)
  
  ld_matrix[i, pos] <- 1
}

num_gt0 <- sum(ld_matrix > 0, na.rm = TRUE)              # number of cells > 0
pct_gt0 <- mean(ld_matrix > 0, na.rm = TRUE) * 100  

# Step 5: Name columns like LD_YYYYMMDD
ld_colnames <- str_replace(dm_cols_sorted, "DM_", "LD4_")
colnames(ld_matrix) <- ld_colnames

print(paste0("# of dates with at least one grid >= ", min_severity, " for ", min_duration, "+ weeks: ", sum(colSums(ld_matrix) > 0), " out of ", dim(ld_matrix)[2], " total weeks"))

# Step 6: Bind to grids
grids <- bind_cols(grids, as.data.frame(ld_matrix))

rm(ld_matrix, dm_matrix, rle_res, row_vals)


```




## calculate per-interval stats 
```{r}
# 
# ## per interval max, median
# 
# library(dplyr)
# library(lubridate)
# library(stringr)
# 
# years <- 2000:2023
# intervals <- 625:635
# interval_start_months <- 1:11
# names(interval_start_months) <- 625:635
# 
# interval_end_months <- 2:12
# names(interval_end_months) <- 625:635
# 
# interval_df <- expand.grid("interval" = intervals, "year" = years)
# 
# interval_df$start_month <- interval_start_months[as.character(interval_df$interval)]
# interval_df$end_month <- interval_end_months[as.character(interval_df$interval)]
# 
# interval_df$start_date <- paste0(interval_df$year, "-", str_pad(interval_df$start_month, 2, "left", "0"), "-01")
# interval_df$end_date <- paste0(interval_df$year, "-", str_pad(interval_df$end_month, 2, "left", "0"), "-31")
# 
# library(stringr)
# library(dplyr)
# 
# for(k in 1:nrow(interval_df)){
#   current_year <- interval_df$year[k]
#   current_interval <- interval_df$interval[k]
#   current_start_month <- str_pad(interval_df$start_month[k], 2, "left", "0")
#   current_end_month <- str_pad(interval_df$end_month[k], 2, "left", "0")
#   
#   print(paste0("processing ", current_year, " ", current_interval))
# 
#   # Select all columns starting with year + start month or year + end month
#   dm_columns <- c(
#     names(grids)[startsWith(names(grids), paste0("DM_", current_year, current_start_month))],
#     names(grids)[startsWith(names(grids), paste0("DM_", current_year, current_end_month))]
#   ) %>% unique()  # remove duplicates if any
# 
#   median_column_name <- paste0("DM_median_", current_year, "_", current_interval)
#   max_column_name <- paste0("DM_max_", current_year, "_", current_interval)
# 
#   grids <- grids %>%
#     rowwise() %>%
#     mutate(
#       dm_values = list(c_across(all_of(dm_columns))),
#       # create columns with dynamic names using :=
#       !!max_column_name := max(dm_values, na.rm = TRUE),
#       !!median_column_name := median(dm_values, na.rm = TRUE)
#     ) %>%
#     ungroup() %>%
#     select(-dm_values)  # remove temp column if not needed
# }
# 
# 
# for(k in 1:nrow(interval_df)){
#   current_year <- interval_df$year[k]
#   current_interval <- interval_df$interval[k]
#   current_start_month <- str_pad(interval_df$start_month[k], 2, "left", 0)
#   current_end_month <- str_pad(interval_df$end_month[k], 2, "left", 0)
#   
#   dm_columns <- c(names(grids)[names(grids) %>% startsWith(paste0("DM_", current_year, current_start_month))],
#                   names(grids)[names(grids) %>% startsWith(paste0("DM_", current_year, current_end_month))])
#   
#   median_column_name <- paste0("DM_median_", current_year, "_", current_interval)
#   max_column_name <- paste0("DM_max_", current_year, "_", current_interval)
# 
#                   
#   grids <- grids %>%
#     rowwise() %>%
#     mutate(
#       dm_values = list(c_across(dm_columns)),  # or any custom column set
#     #  n_non_na = sum(!is.na(dm_values),
#       `max_column_name` = max(dm_values, na.rm = TRUE),
#       `median_column_name` = median(dm_values, na.rm = TRUE),
#     ) %>%
#     ungroup()
#   
# }
# 
# 
# 
# 
# for(k)
# 
# # Example: get all DM columns
# dm_cols <- grep("^DM_\\d{8}$", names(grids), value = TRUE)
# 
# 
# 
# 
# # Create a data frame of date info
# dm_dates <- data.frame(
#   colname = dm_cols,
#   date = ymd(str_remove(dm_cols, "DM_"))
# )
# 
# # Assign each date to a two-month interval ID (e.g., Jan-Feb = 625)
# dm_dates <- dm_dates %>%
#   mutate(
#     year = year(date),
#     month = month(date),
#     interval_id = case_when(
#       month == 1 ~ 625,
#       month == 2 ~ 625,
#       month == 3 ~ 626,
#       month == 4 ~ 626,
#       month == 5 ~ 627,
#       month == 6 ~ 627,
#       month == 7 ~ 628,
#       month == 8 ~ 628,
#       month == 9 ~ 629,
#       month == 10 ~ 629,
#       month == 11 ~ 630,
#       month == 12 ~ 630
#     ),
#     interval_label = paste0("DM_", year, "_", interval_id)
#   )
# 
# # Now, for each year-interval combo, calculate max and median
# for (intvl in unique(dm_dates$interval_label)) {
#   cols_to_use <- dm_dates %>%
#     filter(interval_label == intvl) %>%
#     pull(colname)
#   
#   if (length(cols_to_use) > 0) {
#     grids[[paste0(intvl, "_max")]] <- apply(grids[cols_to_use], 1, max, na.rm = TRUE)
#     grids[[paste0(intvl, "_median")]] <- ceiling(apply(grids[cols_to_use], 1, median, na.rm = TRUE))
#   }
# }
# 
# 

```


```{r}
# library(dplyr)
# library(stringr)
# library(purrr)
# 
# # 1. Extract year and month from DM_ columns
# dm_cols <- names(grids)[startsWith(names(grids), "DM_")]
# 
# # Extract YYYYMMDD from col names: "DM_20000104"
# dates <- str_sub(dm_cols, 4, 11)  # "20000104"
# 
# years <- as.integer(str_sub(dates, 1, 4))
# months <- as.integer(str_sub(dates, 5, 6))
# 
# # 2. Define your intervals, e.g. Jan-Feb = 625, Feb-Mar = 626, etc.
# intervals <- 625:635
# interval_start_months <- 1:11
# names(interval_start_months) <- intervals
# interval_end_months <- 2:12
# names(interval_end_months) <- intervals
# 
# # Create a lookup to assign each month to interval(s)
# # For simplicity, assign each month to all intervals it belongs to
# 
# # But your intervals overlap, so a month can belong to multiple intervals.
# # We want to assign columns to intervals if their month falls between start and end month.
# 
# # Build a tibble of intervals:
# interval_df <- tibble(
#   interval = intervals,
#   start_month = interval_start_months[as.character(intervals)],
#   end_month = interval_end_months[as.character(intervals)]
# )
# 
# # 3. For each column, find which intervals it belongs to
# 
# col_intervals <- map(months, function(m) {
#   interval_df$interval[interval_df$start_month <= m & interval_df$end_month >= m]
# })
# 
# # Now, col_intervals is a list, each element = vector of intervals that column month belongs to
# 
# # 4. Create a mapping: interval -> columns
# 
# interval_to_cols <- setNames(vector("list", length(intervals)), intervals)
# 
# for(i in seq_along(dm_cols)) {
#   for(interval in col_intervals[[i]]) {
#     interval_to_cols[[as.character(interval)]] <- c(interval_to_cols[[as.character(interval)]], dm_cols[i])
#   }
# }
# 
# # 5. Now calculate max and median for each interval across its assigned columns in one big mutate
# 
# for(interval in intervals) {
#   cols_for_int <- unique(interval_to_cols[[as.character(interval)]])
#   max_colname <- paste0("DM_max_", interval)
#   median_colname <- paste0("DM_median_", interval)
# 
#   grids <- grids %>%
#     rowwise() %>%
#     mutate(
#       !!max_colname := max(c_across(all_of(cols_for_int)), na.rm = TRUE),
#       !!median_colname := median(c_across(all_of(cols_for_int)), na.rm = TRUE)
#     ) %>%
#     ungroup()
# }

```


```{r}
years <- 2000:2023
intervals <- 625:635
interval_start_months <- 1:11
names(interval_start_months) <- 625:635

interval_end_months <- 2:12
names(interval_end_months) <- 625:635

interval_df <- expand.grid("interval" = intervals, "year" = years)

interval_df$start_month <- interval_start_months[as.character(interval_df$interval)]
interval_df$end_month <- interval_end_months[as.character(interval_df$interval)]

interval_df$start_date <- paste0(interval_df$year, "-", str_pad(interval_df$start_month, 2, "left", "0"), "-01")
interval_df$end_date <- paste0(interval_df$year, "-", str_pad(interval_df$end_month, 2, "left", "0"), "-31")



for(k in 1:nrow(interval_df)){
  current_year <- interval_df$year[k]
  current_interval <- interval_df$interval[k]
  current_start_month <- str_pad(interval_df$start_month[k], 2, "left", "0")
  current_end_month <- str_pad(interval_df$end_month[k], 2, "left", "0")
  
  print(paste0("processing ", current_year, " ", current_interval))

  # Select all columns starting with year + start month or year + end month
  dm_columns <- c(
    names(grids)[startsWith(names(grids), paste0("DM_", current_year, current_start_month))],
    names(grids)[startsWith(names(grids), paste0("DM_", current_year, current_end_month))]
  ) %>% unique()  # remove duplicates if any

  median_column_name <- paste0("DM_median_", current_year, "_", current_interval)
  max_column_name <- paste0("DM_max_", current_year, "_", current_interval)

  grids <- grids %>%
    rowwise() %>%
    mutate(
      dm_values = list(c_across(all_of(dm_columns))),
      # create columns with dynamic names using :=
      !!max_column_name := max(dm_values, na.rm = TRUE),
      !!median_column_name := median(dm_values, na.rm = TRUE)
    ) %>%
    ungroup() %>%
    select(-dm_values)  # remove temp column if not needed
}

save(grids, file = here("data/grids_usdm_stats.rda"))


```





