---
title: "Untitled"
output: html_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(stringr)
library(tidyr)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```



## initial plots

## summary stats
###mean RI by year

```{r}
load("data/grids_usdm_stats.rda")

ri_threshold <- 0.9
```



## sample grid maps
```{r}

# 1. Recode values
grids <- grids %>%
  mutate(DM_class = ifelse(DM_max_2022_631 == 4, "D4", "Other")) %>% 
  mutate(RI_class = ifelse(cpc_index_2022_631 <= 0.90, "Payout", "No Payout")) %>% 
  mutate(case_dm_4 = ifelse(DM_max_2022_631 == 4, TRUE, FALSE)) %>% 
  mutate(case_ri_90 = ifelse(cpc_index_2022_631 <= 0.90, TRUE, FALSE))



# 2. Plot with two classes
ggplot(grids) +
  geom_sf(aes(fill = DM_class), color = NA) +
  scale_fill_manual(
    values = c(
      "D4" = "darkred",
      "Other" = "grey"
    ),
    name = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "Max USDM July–August 2022")

ggplot(grids) +
  geom_sf(aes(fill = RI_class), color = NA) +
  scale_fill_manual(
    values = c(
      "Payout" = "darkred",
      "No Payout" = "grey"
    ),
    name = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "PRF Payouts July–August 2022")


```

```{r}

grids_collapsed <- grids %>%
  group_by(GRIDCODE) %>%
  summarise(across(everything(), ~ first(.)), .groups = "drop")

```

## crosstab map

```{r}

grids_collapsed$combo <- with(grids_collapsed,
  factor(
    paste(case_dm_4, case_ri_90, sep = "_"),
    levels = c("FALSE_FALSE", "FALSE_TRUE", "TRUE_FALSE", "TRUE_TRUE"),
    labels = c("No drought without payout", 
               "No drought with payout",
               "Drought without payout", "Drought with payout")
  )
)

combo_colors <- c(
  "No drought without payout" = "lightgrey",
  "No drought with payout" = "skyblue",
  "Drought without payout" = "red",
  "Drought with payout" = "orange"
)

ggplot(data = grids_collapsed) +
  geom_sf(aes(fill = combo), color = NA) +
  scale_fill_manual(
    values = combo_colors,
    name = "Drought / Payout Classification"
  ) +
  labs(title = "Crosstab Map: DM = 4 vs RI ≤ 0.9") +
  theme_minimal() +
  theme(
    legend.position = "right",         # or "bottom", "top", etc.
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )




print(table(grids_collapsed$combo))


```

## crosstab table

```{r}

years <- 2007:2023
intervals <- 625:635

for (year in years) {
  for (interval in intervals) {
    # Construct column names dynamically
    dm_col <- paste0("DM_max_", year, "_", interval)
    ri_col <- paste0("cpc_index_", year, "_", interval)
    
    case_dm_col <- paste0("case_dm_4_", year, "_", interval)
    case_ri_col <- paste0("case_ri_90_", year, "_", interval)
    combo_col   <- paste0("combo_", year, "_", interval)
    
    # Only create if both source columns exist
    if (all(c(dm_col, ri_col) %in% names(grids_collapsed))) {
      
      # Boolean columns
      grids_collapsed[[case_dm_col]] <- grids_collapsed[[dm_col]] == 4
      grids_collapsed[[case_ri_col]] <- grids_collapsed[[ri_col]] <= 0.9
      
      # Combo factor column
      grids_collapsed[[combo_col]] <- with(grids_collapsed,
        factor(
          paste(grids_collapsed[[case_dm_col]], grids_collapsed[[case_ri_col]], sep = "_"),
          levels = c("FALSE_FALSE", "FALSE_TRUE", "TRUE_FALSE", "TRUE_TRUE"),
          labels = c("Neither", "RI only", "DM only", "Both")
        )
      )
    }
  }
}

# Initialize count vectors
dm_only     <- integer(nrow(grids_collapsed))
ri_only     <- integer(nrow(grids_collapsed))
both        <- integer(nrow(grids_collapsed))
neither     <- integer(nrow(grids_collapsed))
total_valid <- integer(nrow(grids_collapsed))  # count of total comparisons per row

# Loop through each year and interval
for (year in years) {
  for (interval in intervals) {
    dm_col <- paste0("case_dm_4_", year, "_", interval)
    ri_col <- paste0("case_ri_90_", year, "_", interval)
    
    # Only proceed if both columns exist
    if (all(c(dm_col, ri_col) %in% names(grids_collapsed))) {
      dm_vals <- grids_collapsed[[dm_col]]
      ri_vals <- grids_collapsed[[ri_col]]
      
      valid <- !is.na(dm_vals) & !is.na(ri_vals)  # handle missing data safely
      total_valid <- total_valid + valid
      
      dm_only     <- dm_only     + (dm_vals & !ri_vals)
      ri_only     <- ri_only     + (!dm_vals & ri_vals)
      both        <- both        + (dm_vals & ri_vals)
      neither     <- neither     + (!dm_vals & !ri_vals)
    }
  }
}

# Add raw count columns
grids_collapsed$dm_only_count  <- dm_only
grids_collapsed$ri_only_count  <- ri_only
grids_collapsed$both_count     <- both
grids_collapsed$neither_count  <- neither
grids_collapsed$total_valid    <- total_valid

grids_collapsed$dm_true_count <- grids_collapsed$dm_only_count + grids_collapsed$both_count

grids_collapsed$missed_payout_pct <- 100*(grids_collapsed$dm_only_count/grids_collapsed$dm_true_count)

# Calculate percentages for each class
grids_collapsed$dm_only_pct   <- 100 * grids_collapsed$dm_only_count / grids_collapsed$total_valid
grids_collapsed$ri_only_pct   <- 100 * grids_collapsed$ri_only_count / grids_collapsed$total_valid
grids_collapsed$both_pct      <- 100 * grids_collapsed$both_count    / grids_collapsed$total_valid
grids_collapsed$neither_pct   <- 100 * grids_collapsed$neither_count / grids_collapsed$total_valid

summary_table <- grids_collapsed %>%
  st_drop_geometry() %>%  # remove spatial data for printing
  select(dm_only_pct, ri_only_pct, both_pct, neither_pct) %>%
  summarise(across(everything(), ~ round(mean(.x, na.rm = TRUE), 1)))

print(summary_table)

```


## crosstab table
```{r}

# Sum total counts across all rows
dm_only_total   <- sum(grids_collapsed$dm_only_count, na.rm = TRUE)
ri_only_total   <- sum(grids_collapsed$ri_only_count, na.rm = TRUE)
both_total      <- sum(grids_collapsed$both_count, na.rm = TRUE)
neither_total   <- sum(grids_collapsed$neither_count, na.rm = TRUE)

# Total number of valid combinations evaluated
total_combos <- dm_only_total + ri_only_total + both_total + neither_total

# Percentages
dm_only_pct   <- 100 * dm_only_total / total_combos
ri_only_pct   <- 100 * ri_only_total / total_combos
both_pct      <- 100 * both_total / total_combos
neither_pct   <- 100 * neither_total / total_combos

# Print summary
summary_table <- data.frame(
  category = c("DM only", "RI only", "Both", "Neither"),
  count    = c(dm_only_total, ri_only_total, both_total, neither_total),
  percent  = round(c(dm_only_pct, ri_only_pct, both_pct, neither_pct), 1)
)

print(summary_table)

```



## absent payout %
```{r}

library(RColorBrewer)


# Plot: Continuous scale with viridis "C" reversed
ggplot(data = grids_collapsed) +
  geom_sf(aes(fill = missed_payout_pct), color = NA) +
  scale_fill_viridis_c(
    name = "% Absent Payout",
    option = "C",
    direction = -1,
    limits = c(0, 100),
    na.value = "grey90"  # NA values in light grey
  ) +
  labs(title = "Absent Payout % by Grid Cell") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

```

## absent payout count

```{r}

grids_collapsed$dm_only_count[is.na(grids_collapsed$missed_payout_pct)] <- NA

# Plot: Continuous scale with viridis "C" reversed
ggplot(data = grids_collapsed) +
  geom_sf(aes(fill = dm_only_count), color = NA) +
  scale_fill_viridis_c(
    name = "# Absent Payouts",
    option = "C",
    direction = -1,
    na.value = "grey90"  # NA values in light grey
  ) +
  labs(title = "Absent Payout Count 2007-2023") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )


```



```{r}


# 1. Gather DM_max columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )


# 4. Summarize: for each year, calculate % of intervals with DM ≥ 3
dm_summary <- dm_long %>%
  group_by(year) %>%
  summarise(
    mean_ri = mean(RI, na.rm = TRUE)
  )

# 5. Plot
ggplot(dm_summary, aes(x = year, y = mean_ri)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = "Mean RI",
    title = "Mean RI by Year"
  ) +
  ylim(0,2) +
  theme_minimal(base_size = 14)






```



### % obs for different RI thresholds

```{r}

# 1. Gather DM_max columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )


# 4. Summarize: for each year, calculate % of intervals with RI < thresholds
for(threshold in seq(0.7, 0.9, 0.05)){
  print(threshold)
  pct <- 100*(sum(dm_long$RI < threshold)/nrow(dm_long))
  print(paste0(round(pct, 1), "%"))
}



```

### summary stats, DM category
```{r}

# 1. Gather DM_max columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_max_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "DM_value"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "DM_max_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )


# 4. Summarize: for each year, calculate % of intervals with RI < thresholds
for(threshold in seq(0, 4, 1)){
  print(threshold)
  pct <- 100*(sum(dm_long$DM_value >= threshold)/nrow(dm_long))
  print(paste0(round(pct, 1), "%"))
}


```


## by year, >= 3, 4

```{r}



# 1. Gather DM_max columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_max_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "dm_value"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "DM_max_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )

# 3. For each row (site × interval), define if DM ≥ 3
dm_long <- dm_long %>%
  mutate(high_dm = dm_value >= 3)

# 4. Summarize: for each year, calculate % of intervals with DM ≥ 3
dm_summary <- dm_long %>%
  group_by(year) %>%
  summarise(
    pct_high_dm = mean(high_dm, na.rm = TRUE) * 100
  )

# 5. Plot
ggplot(dm_summary, aes(x = year, y = pct_high_dm)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = "% intervals where max DM ≥ 3",
    title = "High Drought Frequency by Year"
  ) +
  ylim(0,100) +
  theme_minimal(base_size = 14)


# 3. For each row (site × interval), define if DM ≥ 4
dm_long <- dm_long %>%
  mutate(high_dm = dm_value >= 4)

# 4. Summarize: for each year, calculate % of intervals with DM ≥ 4
dm_summary <- dm_long %>%
  group_by(year) %>%
  summarise(
    pct_high_dm = mean(high_dm, na.rm = TRUE) * 100
  )

# 5. Plot
ggplot(dm_summary, aes(x = year, y = pct_high_dm)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = "% intervals where max DM ≥ 4",
    title = "High Drought Frequency by Year"
  ) +
  ylim(0,100) +
  theme_minimal(base_size = 14)



```


## intervals >=3, >= 4

```{r}


# 1. Make sure you have the correct long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_max_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_max_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(dm_value)) %>%
  mutate(high_dm = dm_value >= 3)

# 2. Summarize: % of rows × years where DM_max ≥ 3 by interval
interval_summary <- dm_long %>%
  group_by(interval) %>%
  summarise(
    pct_high_dm = mean(high_dm) * 100,
    .groups = "drop"
  )

# 3. Create readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

# 4. Join labels
interval_summary <- interval_summary %>%
  left_join(interval_labels, by = "interval")

# 5. Plot
ggplot(interval_summary, aes(x = factor(label, levels = label), y = pct_high_dm)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Interval",
    y = "% obs with DM ≥ 3",
    title = "High Drought Frequency by Calendar Interval"
  ) +
  theme_minimal(base_size = 14) +  
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20),         # major breaks every 20
    minor_breaks = seq(0, 100, 10)    # minor breaks every 10
  ) 



# 1. Make sure you have the correct long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_max_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_max_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(dm_value)) %>%
  mutate(high_dm = dm_value >= 4)

# 2. Summarize: % of rows × years where DM_max ≥ 4 by interval
interval_summary <- dm_long %>%
  group_by(interval) %>%
  summarise(
    pct_high_dm = mean(high_dm) * 100,
    .groups = "drop"
  )

# 3. Create readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

# 4. Join labels
interval_summary <- interval_summary %>%
  left_join(interval_labels, by = "interval")

# 5. Plot
ggplot(interval_summary, aes(x = factor(label, levels = label), y = pct_high_dm)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Interval",
    y = "% obs with DM ≥ 4",
    title = "High Drought Frequency by Calendar Interval"
  ) +
  theme_minimal(base_size = 14) +  
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20),         # major breaks every 20
    minor_breaks = seq(0, 100, 10)    # minor breaks every 10
  ) 

```


## heatmap DM >=3, >=4

```{r}

# Step 1: Long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_max_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_max_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(dm_value)) %>%
  mutate(high_dm = dm_value >= 3)

# Step 2: Summarize by year × interval
heatmap_data <- dm_long %>%
  group_by(year, interval) %>%
  summarise(
    pct_high_dm = mean(high_dm) * 100,
    .groups = "drop"
  )

# Step 3: Add readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

heatmap_data <- heatmap_data %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))  # Preserve order

# Step 4: Plot heatmap
ggplot(heatmap_data, aes(x = label, y = factor(year), fill = pct_high_dm)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "% DM ≥ 3", option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = "Heatmap of High Drought Frequency (DM_max ≥ 3)",
    fill = "% High DM"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )


# Step 1: Long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_max_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_max_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(dm_value)) %>%
  mutate(high_dm = dm_value >= 4)

# Step 2: Summarize by year × interval
heatmap_data <- dm_long %>%
  group_by(year, interval) %>%
  summarise(
    pct_high_dm = mean(high_dm) * 100,
    .groups = "drop"
  )

# Step 3: Add readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

heatmap_data <- heatmap_data %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))  # Preserve order

# Step 4: Plot heatmap
ggplot(heatmap_data, aes(x = label, y = factor(year), fill = pct_high_dm)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "% DM = 4", option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = "Heatmap of High Drought Frequency (DM Max = 4)",
    fill = "% High DM"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )


```


## similar plots for RI
### year

```{r}
 
# 1. Gather DM_max columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )

# 3. For each row (site × interval), define if DM ≥ 3
dm_long <- dm_long %>%
  mutate(low_ri = RI < ri_threshold)

# 4. Summarize: for each year, calculate % of intervals with DM ≥ 3
dm_summary <- dm_long %>%
  group_by(year) %>%
  summarise(
    pct_low_ri = mean(low_ri, na.rm = TRUE) * 100
  )

# 5. Plot
ggplot(dm_summary, aes(x = year, y = pct_low_ri)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = paste0("% obs where RI ≤ ", ri_threshold), 
    title = "Low RI Frequency by Year"
  ) +
  ylim(0,100) +
  theme_minimal(base_size = 14)

# # 3. For each row (site × interval), define if DM ≥ 3
# dm_long <- dm_long %>%
#   mutate(low_ri = RI < 0.9)
# 
# # 4. Summarize: for each year, calculate % of intervals with DM ≥ 3
# dm_summary <- dm_long %>%
#   group_by(year) %>%
#   summarise(
#     pct_low_ri = mean(low_ri, na.rm = TRUE) * 100
#   )
# 
# # 5. Plot
# ggplot(dm_summary, aes(x = year, y = pct_low_ri)) +
#   geom_line(color = "firebrick", size = 1.2) +
#   geom_point(color = "black") +
#   scale_y_continuous(labels = scales::percent_format(scale = 1)) +
#   labs(
#     x = "Year",
#     y = "% obs where RI ≤ 0.9", 
#     title = "High Drought Frequency by Year"
#   ) +
#   ylim(0,100) +
#   theme_minimal(base_size = 14)
# 


```



## interval

```{r}


# 1. Make sure you have the correct long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  ) %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(RI)) %>%
  mutate(low_ri = RI < ri_threshold)

# 2. Summarize: % of rows × years where DM_max ≥ 3 by interval
interval_summary <- dm_long %>%
  group_by(interval) %>%
  summarise(
    pct_low_ri = mean(low_ri) * 100,
    .groups = "drop"
  )

# 3. Create readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

# 4. Join labels
interval_summary <- interval_summary %>%
  left_join(interval_labels, by = "interval")

# 5. Plot
ggplot(interval_summary, aes(x = factor(label, levels = label), y = pct_low_ri)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Interval",
    y = paste0("% obs where RI ≤ ", ri_threshold),
    title = "Low RI by Calendar Interval"
  ) +
  theme_minimal(base_size = 14) +  
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20),         # major breaks every 20
    minor_breaks = seq(0, 100, 10)    # minor breaks every 10
  ) 




```


## heatmap

```{r}


# Step 1: Long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  ) %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(RI)) %>%
  mutate(low_ri = RI < ri_threshold)

# Step 2: Summarize by year × interval
heatmap_data <- dm_long %>%
  group_by(year, interval) %>%
  summarise(
    pct_low_ri = mean(low_ri) * 100,
    .groups = "drop"
  )

# Step 3: Add readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

heatmap_data <- heatmap_data %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))  # Preserve order

# Step 4: Plot heatmap
ggplot(heatmap_data, aes(x = label, y = factor(year), fill = pct_low_ri)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = paste0("% RI ≤ ", ri_threshold), option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of Low RI Frequency (RI ≤ ", ri_threshold, ")"),
    fill = "% Low RI"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

```

## payouts



## chatgpt, DM ≥3
```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)

# 1. Pivot DM_max columns to long

dm_long <- grids %>%
  st_drop_geometry() %>%
  select(GRIDCODE, starts_with("DM_max_")) %>%
  pivot_longer(
    cols = starts_with("DM_max_"),
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_max_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6)),
    row_id = row_number()  # optional: track row identity
  )

# 2. Pivot cpc_index columns to long
cpc_long <- grids %>% st_drop_geometry() %>% 
  select(GRIDCODE, starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = starts_with("cpc_index_"),
    names_to = "var",
    values_to = "cpc_value"
  ) %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  mutate(row_id = row_number())

# 3. Join DM_max and CPC on row + year + interval
joined <- left_join(dm_long, cpc_long, by = c("GRIDCODE", "row_id", "year", "interval"))

# 4. Filter where DM_max >= 3
filtered <- joined %>%
  filter(!is.na(dm_value), dm_value >= 3, !is.na(cpc_value))

# 5. Calculate percentage of cpc_value < ri_threshold among these filtered
summary_df <- filtered %>%
  group_by(year) %>%
  summarise(
    pct_cpc_below_0_7 = mean(cpc_value < ri_threshold, na.rm = TRUE) * 100,
    count = n()
  )

# 6. Plot
ggplot(summary_df, aes(x = year, y = pct_cpc_below_0_7)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(limits = c(0, 100), labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = paste0("% CPC Index < ", ri_threshold, " (where DM_max ≥ 3)"),
    title = "Frequency of Low CPC Index During High Drought Periods"
  ) +
  theme_minimal(base_size = 14)


# 4. Filter where DM_max >= 4
filtered <- joined %>%
  filter(!is.na(dm_value), dm_value >= 4, !is.na(cpc_value))

# 5. Calculate percentage of cpc_value < ri_threshold among these filtered
summary_df <- filtered %>%
  group_by(year) %>%
  summarise(
    pct_cpc_below_0_7 = mean(cpc_value < ri_threshold, na.rm = TRUE) * 100,
    count = n()
  )

# 6. Plot
ggplot(summary_df, aes(x = year, y = pct_cpc_below_0_7)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(limits = c(0, 100), labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = paste0("% CPC Index < ", ri_threshold, " (where DM_max ≥ 4)"),
    title = "Frequency of Low CPC Index During High Drought Periods"
  ) +
  theme_minimal(base_size = 14)



```

## missed payouts
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)


# 4. Filter to DM >= 3 observations
filtered <- joined %>% filter(!is.na(dm_value), dm_value >= 3)

# 5. Calculate counts by year
summary_counts <- filtered %>%
  group_by(year) %>%
  summarise(
    count_dm_ge_3 = n(),
    count_missed_payouts = sum(cpc_value >= ri_threshold, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(count_dm_ge_3, count_missed_payouts),
    names_to = "metric",
    values_to = "count"
  )

# 6. Plot two lines
ggplot(summary_counts, aes(x = year, y = count, color = metric)) +
  geom_line(size = 1.2) +
  geom_point() +
  scale_color_manual(
    values = c(count_dm_ge_3 = "steelblue", count_missed_payouts = "firebrick"),
    labels = c("DM ≥ 3 observations", paste0("Missed payouts (RI ≥ ", ri_threshold, " & DM ≥ 3)"))
  ) +
  labs(
    x = "Year",
    y = "Count of observations",
    color = "Metric",
    title = "Counts by Year of Drought Severity and Missed Payouts"
  ) +
  theme_minimal(base_size = 14)

# 4. Filter to DM >= 3 observations
filtered <- joined %>% filter(!is.na(dm_value), dm_value >= 4)

# 5. Calculate counts by year
summary_counts <- filtered %>%
  group_by(year) %>%
  summarise(
    count_dm_ge_4 = n(),
    count_missed_payouts = sum(cpc_value >= ri_threshold, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(count_dm_ge_4, count_missed_payouts),
    names_to = "metric",
    values_to = "count"
  )

# 6. Plot two lines
ggplot(summary_counts, aes(x = year, y = count, color = metric)) +
  geom_line(size = 1.2) +
  geom_point() +
  scale_color_manual(
    values = c(count_dm_ge_4 = "steelblue", count_missed_payouts = "firebrick"),
    labels = c("DM ≥ 4 observations", paste0("Missed payouts (RI ≥ ", ri_threshold, " & DM ≥ 4)") )
  ) +
  labs(
    x = "Year",
    y = "Count of observations",
    color = "Metric",
    title = "Counts by Year of Drought Severity and Missed Payouts"
  ) +
  theme_minimal(base_size = 14)

```

## heatmap, DM ≥3

```{r}

# Filter for DM >= 3 only
filtered <- joined %>% filter(!is.na(dm_value), dm_value >= 3)

# Summarize counts by year × interval
heatmap_data <- filtered %>%
  group_by(year, interval) %>%
  summarise(
    count_dm_ge_3 = n(),
    count_missed_payouts = sum(cpc_value >= ri_threshold, na.rm = TRUE),
    pct_missed_payouts = 100 * count_missed_payouts / count_dm_ge_3,
    .groups = "drop"
  )

interval_start_months <- 1:11
names(interval_start_months) <- 625:635
interval_end_months <- 2:12
names(interval_end_months) <- 625:635

interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]], 
    "-", 
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

heatmap_data <- heatmap_data %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))

ggplot(heatmap_data, aes(x = label, y = factor(year), fill = count_missed_payouts)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Count Missed Payouts", option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of Missed Payout Counts (DM ≥ 3 & RI ≥ ", ri_threshold, ")")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

ggplot(heatmap_data, aes(x = label, y = factor(year), fill = pct_missed_payouts)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "% Missed Payouts", option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of % Missed Payouts (DM ≥ 3 & RI ≥ ", ri_threshold, ")")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )



```
## export DM = 4 counts

```{r}

joined$missed_payout <- (joined$dm_value == 4 & joined$cpc_value >= ri_threshold)

grids_missed_payouts <- joined %>%
  group_by(GRIDCODE, year) %>%
  summarise(
    count_4 = sum(dm_value == 4, na.rm = TRUE),
    count_4_and_true = sum(dm_value == 4 & missed_payout == TRUE, na.rm = TRUE),
    .groups = "drop"
  )

grids_missed_payouts$missed_payout_count <- grids_missed_payouts$count_4_and_true/25

grids_missed_payouts <- grids_missed_payouts %>%
  mutate(
    missed_payout_pct = ifelse(count_4 == 0, 0, 100*(count_4_and_true/count_4))
  )

save(grids_missed_payouts, file = here("data/grids_missed_payouts_dm4.rda"))

```



## heatmap, DM ≥ 4

```{r}

# Filter for DM >= 4 only
filtered <- joined %>% filter(!is.na(dm_value), dm_value >= 4)

# Summarize counts by year × interval
heatmap_data <- filtered %>%
  group_by(year, interval) %>%
  summarise(
    count_dm_ge_4 = n(),
    count_missed_payouts = sum(cpc_value >= ri_threshold, na.rm = TRUE),
    pct_missed_payouts = 100 * count_missed_payouts / count_dm_ge_4,
    .groups = "drop"
  )

interval_start_months <- 1:11
names(interval_start_months) <- 625:635
interval_end_months <- 2:12
names(interval_end_months) <- 625:635

interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]], 
    "-", 
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

heatmap_data <- heatmap_data %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))

ggplot(heatmap_data, aes(x = label, y = factor(year), fill = count_missed_payouts)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Count Missed Payouts", option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of Missed Payout Counts (DM ≥ 4 & RI ≥ ", ri_threshold, ")")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

ggplot(heatmap_data, aes(x = label, y = factor(year), fill = pct_missed_payouts)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "% Missed Payouts", option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of % Missed Payouts (DM ≥ 4 & RI ≥ ", ri_threshold, ")")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )



```

## maps

```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(sf)
library(ggplot2)
library(viridis)

# Assume grids is an sf object with geometry and lots of DM_max_ & cpc_index_ columns

# Get row ID for spatial join (row number)
grids <- grids %>% mutate(row_id = row_number())

# DM_max long format
dm_long <- grids %>%
  st_drop_geometry() %>%
  select(row_id, starts_with("DM_max_")) %>%
  pivot_longer(
    cols = -row_id,
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_max_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )

# cpc_index long format
cpc_long <- grids %>%
  st_drop_geometry() %>%
  select(row_id, starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = -row_id,
    names_to = "var",
    values_to = "ri_value"
  ) %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )

# Join DM and RI by row_id, year, interval
joined <- left_join(dm_long, cpc_long, by = c("row_id", "year", "interval"))

# Filter where DM >= 3
filtered <- joined %>%
  filter(!is.na(dm_value), dm_value >= 3)

# Define missed payout = RI >= ri_threshold
filtered <- filtered %>%
  mutate(missed_payout = ri_value >= ri_threshold)

# Aggregate per spatial cell (row_id)
agg <- filtered %>%
  group_by(row_id) %>%
  summarise(
    missed_payout_count = sum(missed_payout, na.rm = TRUE),
    total_dm_ge_3 = n(),
    missed_payout_pct = 100 * missed_payout_count / total_dm_ge_3
  )

# Join back to grids sf object
grids_map <- grids %>%
  left_join(agg, by = "row_id")

# Map 1: Missed payout counts
ggplot(grids_map) +
  geom_sf(aes(fill = missed_payout_count), color = NA) +
  scale_fill_viridis_c(option = "C", na.value = "grey90", direction = -1) +
  labs(fill = "Missed Payout Count", title = "Count of Missed Payouts per Cell") +
#  zlim(0,100) +
  theme_minimal()

# Map 2: Missed payout percentage
ggplot(grids_map) +
  geom_sf(aes(fill = missed_payout_pct), color = NA) +
  scale_fill_viridis_c(option = "C", na.value = "grey90", direction = -1) +
  labs(fill = "% Missed Payouts", title = "Percent of Missed Payouts per Cell") +
  theme_minimal()



```


## comparison


```{r}

a <- st_read("/Users/mjcecil/Documents/GitHub/usdm-prf/data/USDM_GIS/2022_USDM_M/USDM_20220802.shp")
a <- st_make_valid(a)
tx_counties <- st_read(here("data/Tx_CntyBndry_Jurisdictional_TIGER/Tx_CntyBndry_Jurisdictional_TIGER.shp"))


# 2. Ensure both are in the same CRS
a <- st_transform(a, st_crs(tx_counties))

# 3. Clip drought data to TX counties
a_tx <- st_intersection(a, tx_counties)

# 4. Identify counties not intersecting any drought zone
# First, union drought geometries
a_union <- st_union(a)

# Find areas in TX counties not covered by drought
no_drought_geom <- st_difference(tx_counties, a_union)

# Add DM = -1 to those
no_drought_sf <- no_drought_geom %>%
  mutate(DM = -1)

# 5. Combine drought data with the -1 zones
a_tx_cleaned <- a_tx %>%
  select(DM) %>%
  mutate(DM = as.numeric(as.character(DM)))  # ensure numeric

combined <- bind_rows(a_tx_cleaned, no_drought_sf)

# 6. Plot with custom colors
ggplot(combined) +
  geom_sf(aes(fill = factor(DM)), color = NA) +
  scale_fill_manual(
    values = c(
      "-1" = "darkgreen",
      "0"  = "lightgreen",
      "1"  = "yellow",
      "2"  = "orange",
      "3"  = "red",
      "4"  = "darkred"
    ),
    name = "Drought Level",
    labels = c(
      "-1" = "No Classification",
      "0" = "D0",
      "1" = "D1",
      "2" = "D2",
      "3" = "D3",
      "4" = "D4"
    )
  ) +
  theme_minimal() +
  labs(title = "USDM Aug 2, 2022") + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank())



plot(a)

grids_pt <- grids %>% st_drop_geometry() %>% st_as_sf(., coords = c("lon", "lat"), crs = 4326)

grids_joined <- st_join(grids_pt, a, left = TRUE) %>%
  mutate(DM = coalesce(DM, -1))

grids_joined$DM[is.na(grids_joined$DM)] <- -1

plot(grids_joined['DM'])

rm(a)

```

```{r}

ggplot(grids) +
  geom_sf(aes(fill = factor(DM_max_2022_628)), color = NA) +
  scale_fill_manual(
    values = c(
      "-1" = "darkgreen",
      "0"  = "lightgreen",
      "1"  = "yellow",
      "2"  = "orange",
      "3"  = "red",
      "4"  = "darkred"
    ),
    name = NULL,  # Removes legend title
    labels = c(
      "-1" = "No Classification",
      "0" = "D0",
      "1" = "D1",
      "2" = "D2",
      "3" = "D3",
      "4" = "D4"
    )
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "Max USDM July-August 2022")

```

```{r}

# 1. Recode values
grids <- grids %>%
  mutate(DM_class = ifelse(DM_max_2022_631 == 4, "D4", "Other"))

# 2. Plot with two classes
ggplot(grids) +
  geom_sf(aes(fill = DM_class), color = NA) +
  scale_fill_manual(
    values = c(
      "D4" = "darkred",
      "Other" = "grey"
    ),
    name = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "Max USDM July–August 2022")

```

## table counts for DM prevalence

```{r}

# Define year and interval ranges
years <- 2007:2023
intervals <- 625:635

# Generate column names of interest
target_cols <- outer(years, intervals, function(y, i) paste0("DM_max_", y, "_", i)) |> as.vector()

# Filter columns that actually exist
target_cols <- target_cols[target_cols %in% names(grids_collapsed)]

# Extract and convert to matrix
dm_matrix <- grids_collapsed[, target_cols]
if ("sf" %in% class(grids_collapsed)) dm_matrix <- st_drop_geometry(dm_matrix)
dm_matrix <- as.matrix(dm_matrix)
dm_matrix <- apply(dm_matrix, 2, as.numeric)  # ensure numeric

# Calculate percent of all values >= each threshold
thresholds <- 0:4
percent_ge <- sapply(thresholds, function(t) mean(dm_matrix >= t, na.rm = TRUE) * 100)

# Create and print results table
result_df <- data.frame(
  Threshold = thresholds,
  Percent_GE = round(percent_ge, 1)
)

print(result_df)



```

## table counts for RI prevalence

```{r}

# Define year and interval ranges
years <- 2007:2023
intervals <- 625:635

# Generate column names of interest
target_cols <- outer(years, intervals, function(y, i) paste0("cpc_index_", y, "_", i)) |> as.vector()

# Filter columns that actually exist
target_cols <- target_cols[target_cols %in% names(grids_collapsed)]

# Extract and convert to matrix
dm_matrix <- grids_collapsed[, target_cols]
if ("sf" %in% class(grids_collapsed)) dm_matrix <- st_drop_geometry(dm_matrix)
dm_matrix <- as.matrix(dm_matrix)
dm_matrix <- apply(dm_matrix, 2, as.numeric)  # ensure numeric

# Calculate percent of all values >= each threshold
thresholds <- seq(0.70, 0.90, 0.05)
percent_le <- sapply(thresholds, function(t) mean(dm_matrix <= t, na.rm = TRUE) * 100)

# Create and print results table
result_df <- data.frame(
  Threshold = thresholds,
  Percent_GE = round(percent_le, 1)
)

print(result_df)



```