---
title: "Untitled"
output: html_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(stringr)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```



## initial plots

## summary stats
###mean RI by year

```{r}
load("data/grids_usdm_stats.rda")

ri_threshold <- 0.9
```

```{r}

load("/Users/mjcecil/Documents/GitHub/prf-ri/data/outputs/payouts/tpu_Grazing_2023_cpc_index.rda")

tpu$roi <- (tpu$Indemnity_cpc_index)/tpu$Premium_cpc_index

```

## enrollment by year (acres, liability)


## coverage level breakdown by year


## interval breakdown by year

## define when a county experiences a drought event (base on intersecting CPC grids, use max or median)

```{r}

## load counties shapefile
tx_counties <- st_read(here("data/Tx_CntyBndry_Jurisdictional_TIGER/Tx_CntyBndry_Jurisdictional_TIGER.shp"))

## for each county, get list of grid codes

tx_county_areas <- read.csv(here("data/tx_county-areas/cpc-county-areas.csv"))

## take median (or max) of grid code values





```

```{r}
# Set your working directory if needed
# setwd("path/to/your/files")

# List of years
years <- 2007:2023

# Initialize an empty list to store dataframes
tpu_list <- list()

# Loop through each year
for (yr in years) {
  # Construct file name
  file_name <- sprintf(here("data/payouts/tpu_Grazing_%d_cpc_index.rda"), yr)
  
  # Load file, `tpu` will be created in the environment
  load(file_name)
  
  # Add year column
  tpu$year <- yr
  
  # Append to list
  tpu_list[[as.character(yr)]] <- tpu
}

# Combine all into one data frame
tpu_all_years <- do.call(rbind, tpu_list)

# Check the result
head(tpu_all_years)

```


```{r}
## fix interval codes for early years

   ## in 2009 and before, there were only 6 intervals 221... 226, represeting Jan-Feb, Mar-Apr
interval_dict <- c("221" = 625, "222" = 627, "223" = 629, "224" = 631, "225" = 633, "226" = 635)
      
# Convert names of interval_dict to numeric for comparison
interval_keys <- as.numeric(names(interval_dict))

tpu_all_years <- tpu_all_years %>%
  mutate(
    Practice.Code = case_when(
      Practice.Code %in% interval_keys ~ interval_dict[as.character(Practice.Code)],
      TRUE ~ Practice.Code  # keep original if no match
    )
  )

```


```{r}

# Create empty vector to store results
DM_max_values <- numeric(nrow(tpu_all_years))

# Loop over each row of tpu_all_years
for (i in seq_len(nrow(tpu_all_years))) {
  # Extract identifiers
  county <- tpu_all_years$County.Code[i]
  year <- tpu_all_years$year[i]
  practice <- tpu_all_years$Practice.Code[i]
  
  # Step 2: Filter tx_county_areas
  area_rows <- tx_county_areas[
    tx_county_areas$County.Code == county, ]
  
  # Skip if no matching area rows
  if (nrow(area_rows) == 0) {
    DM_max_values[i] <- NA
    next
  }

  # Step 2b: Extract unique CPC grid codes
  grid_codes <- unique(area_rows$CPC.Grid.Code)

  # Step 3: Filter grids
  county_rows <- grids[grids$GRIDCODE %in% grid_codes, ]
  
  # Construct the column name
  col_name <- paste0("DM_max_", year, "_", practice)

  # Step 4: Check if column exists
  if (!col_name %in% colnames(county_rows)) {
    DM_max_values[i] <- NA
    next
  }

  # Step 4b: Get max value (excluding NAs)
  DM_max_values[i] <- max(county_rows[[col_name]], na.rm = TRUE)
}

# Add to tpu_all_years
tpu_all_years$DM_max_county <- DM_max_values

save(tpu_all_years, file = here('data/tpu_all_years.rda'))

```





## make a visualization of CPC grids, counties, and USDM classification 

```{r}

load(file = here('data/tpu_all_years.rda'))

```


```{r}
usdm <- st_read("/Users/mjcecil/Documents/GitHub/usdm-prf/data/USDM_GIS/2022_USDM_M/USDM_20220503.shp") %>% st_intersection(., tx_counties)

plot(usdm['DM'])

plot(grids["DM_max_2022_628"])

library(viridis)
# Extract the numeric range for coloring
zvals <- grids$cpc_index_2022_628

# Create a reversed color palette (adjust number of colors as needed)
col_pal <- rev(terrain.colors(20))  # or use viridis(20), etc.

# Plot with reversed colors
plot(grids["cpc_index_2022_628"])


ggplot(grids) +
  geom_sf(aes(fill = cpc_index_2022_628), color = "white", size = 0.1) +
  scale_fill_viridis_c(
    option = "plasma",  # or "plasma", "magma", "cividis", etc.
    direction = -1,      # <- this flips the scale
    na.value = "grey80",
    name = "CPC Index 2022"
  ) +
  theme_minimal()

grids_2 <- grids
grids_2$payout <- as.numeric(grids_2$cpc_index_2022_628 < 0.9)

ggplot(grids_2) +
  geom_sf(aes(fill = payout), color = "white", size = 0.1) +
  scale_fill_viridis_c(
    option = "plasma",  # or "plasma", "magma", "cividis", etc.
    direction = 1,      # <- this flips the scale
    na.value = "grey80",
    name = "Payout"
  ) +
  theme_minimal()


```

```{r}

# Perform left join, selecting only needed columns
tx_counties_joined <- tx_counties %>%
  left_join(
    tpu_all_years_filt %>%
      select(County.Code, DM_max_county, ROI, Net.Reporting.Level.Amount, Indemnity_cpc_index),
    by = "County.Code"
  )

plot(tx_counties_joined['Indemnity_cpc_index'])


## county acreage (0.90)
zvals <- tx_counties_joined$Net.Reporting.Level.Amount
brks <- pretty(zvals, n = 6)

plot(
  tx_counties_joined["Net.Reporting.Level.Amount"],
  breaks = brks,
  col = viridis::viridis(length(brks) - 1),
  main = "Net Reporting Level Amount (Acreage)"
)

## county ROI (0.90)



```



## for each year-interval combination, counties are divided into those that experienced severe drought and those that didn't

```{r}

tpu_all_years$ROI <- tpu_all_years$Indemnity_cpc_index/tpu_all_years$Premium_cpc_index

current_year <- 2022
next_year <- current_year + 1
current_interval <- 628
current_coverage_level <- 0.90

tpu_all_years_filt_all_coverage <- tpu_all_years %>% filter(year == current_year) %>% filter(Practice.Code == current_interval)

tpu_all_years_filt_next_year_all_coverage <- tpu_all_years %>% filter(year == next_year) %>% filter(Practice.Code == current_interval)

tpu_all_years_filt <- tpu_all_years %>% filter(year == current_year) %>% filter(Coverage.Level.Percent == current_coverage_level) %>% filter(Practice.Code == current_interval)



```

## example viz

```{r}

tx_counties$County.Code <- as.numeric(tx_counties$COUNTYID)

# Perform left join, selecting only needed columns
tx_counties_joined <- tx_counties %>%
  left_join(
    tpu_all_years_filt %>%
      select(County.Code, DM_max_county, ROI),
    by = "County.Code"
  )

ggplot(tx_counties_joined) +
  geom_sf(aes(fill = DM_max_county), color = "white") +
  scale_fill_gradientn(
    name = "DM_max_county",
    colours = c(
      "#08306B",  # Dark blue (-1)
      "#2171B5",  # Blue (0)
      "#66C2A4",  # Light green (1)
      "#FFFFB2",  # Yellow (2)
      "#FE9929",  # Orange (3)
      "#CC0000"   # Red (4)
    ),
    values = scales::rescale(c(-1, 0, 1, 2, 3, 4)),
    limits = c(-1, 4),
    na.value = "grey"
  ) +
  theme_minimal()



```


```{r}


# Define breaks and labels including >4
breaks <- c(0, 0.5, 0.8, 1, 1.5, 2, 4, Inf)
labels <- c("0–0.5", "0.5–0.8", "0.8–1", "1–1.5", "1.5–2", "2–4", ">4")

# Bin the ROI values
tx_counties_joined$ROI_binned <- cut(
  tx_counties_joined$ROI,
  breaks = breaks,
  labels = labels,
  include.lowest = TRUE,
  right = FALSE
)

# Reverse factor levels so legend order reverses
tx_counties_joined$ROI_binned <- factor(
  tx_counties_joined$ROI_binned,
  levels = rev(labels)  # Reverse the order here
)

# Get viridis colors for 7 bins
vir_colors <- viridis::viridis(length(labels))

# Since factor levels reversed, reverse colors too to keep mapping consistent
vir_colors_rev <- rev(vir_colors)

ggplot(tx_counties_joined) +
  geom_sf(aes(fill = ROI_binned), color = "white") +
  scale_fill_manual(
    name = "ROI range",
    values = vir_colors_rev,
    na.value = "grey",
    breaks = rev(labels),
    labels = rev(labels),
    drop = FALSE       # <--- Keep NA in the legend
  ) +
  theme_minimal()

```



## violin 

```{r}

# Exclude NA from DM_max_county and convert to factor
tx_counties_joined$DM_max_factor <- factor(
  tx_counties_joined$DM_max_county,
  levels = sort(na.omit(unique(tx_counties_joined$DM_max_county)))
)

ggplot(tx_counties_joined %>% filter(!is.na(ROI)), aes(x = DM_max_factor, y = ROI)) +
  geom_violin(fill = "skyblue", color = "black", alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.3, size = 0.8) +
  geom_point(aes(x = Inf, y = Inf, shape = "Median"), color = "red", fill = "red", size = 3) +
  stat_summary(
    fun = median, geom = "point", shape = 23, size = 3, fill = "red",
    aes(shape = "Median"), show.legend = TRUE
  ) +
  scale_shape_manual(
    name = "", 
    values = 23, 
    labels = "Median (red diamond)"
  ) +
  labs(
    x = "DM Max",
    y = "ROI",
    title = "Distribution of ROI by DM Max Groups"
  ) +
  theme_minimal()  +
  theme(legend.position = "right")

```

## producer decision points - enrollment (acres, liability), coverage level, interval change

```{r}


# Step 1: Create lookup from 2022
dm_lookup <- tpu_all_years_filt_all_coverage %>%
  select(County.Code, DM_max_county) %>%
  distinct()

# Step 2: Join 2022 DM category to 2023 data
tpu_all_years_filt_next_year_all_coverage$DM_max_county <- NULL
tpu_2023_with_dm <- tpu_all_years_filt_next_year_all_coverage %>%
  left_join(dm_lookup, by = "County.Code")

# Step 3: Aggregate acreage for both years using 2022 DM category
enrollment_2022 <- tpu_all_years_filt_all_coverage %>%
  group_by(DM_max_county) %>%
  summarise(acres_2022 = sum(Net.Reporting.Level.Amount, na.rm = TRUE))

enrollment_2023 <- tpu_2023_with_dm %>%
  group_by(DM_max_county) %>%
  summarise(acres_2023 = sum(Net.Reporting.Level.Amount, na.rm = TRUE))

# Step 4: Combine and calculate percent change
enrollment_compare <- full_join(enrollment_2022, enrollment_2023, by = "DM_max_county") %>%
  mutate(
    percent_change = 100 * (acres_2023 - acres_2022) / acres_2022,
    DM_max_county = factor(DM_max_county, levels = sort(unique(DM_max_county)))
  )


```

```{r}

library(tidyr)

# Pivot longer for plotting
enrollment_long <- enrollment_compare %>%
  pivot_longer(cols = c(acres_2022, acres_2023), names_to = "year", values_to = "acres")

ggplot(enrollment_long, aes(x = DM_max_county, y = acres, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "DM Max (2022)",
    y = "Acres Enrolled",
    fill = "Year",
    title = "Enrollment (Acres) by 2022 DM Category"
  ) +
  scale_fill_manual(values = c("acres_2022" = "skyblue", "acres_2023" = "orange"),
                    labels = c("2022", "2023")) +
  theme_minimal()

```

```{r}

ggplot(enrollment_compare, aes(x = DM_max_county, y = percent_change)) +
  geom_bar(stat = "identity", fill = "purple") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "DM Max (2022)",
    y = "% Change in Enrollment",
    title = "Percent Change in Enrollment from 2022 to 2023"
  ) +
  theme_minimal()


```


### what is the metric we most expect producers to respond and change behavior.




## among those that experienced drought, you will have different levels of payout. we want to compare these levels of ROI (i.e. (indemnity - premium)/indemnity) to enrollment in the following year

## similar as above but for counties that did not experience severe drought

## first comparison is between all drought vs all non-drought

## second comparison is based on level of payout for drought, level of payout for non-drought

## two years to look at to start 2011 to 2012 (severe drought, many payouts)

## 2018 to 2019 (more limited severe drought)





