---
title: "Untitled"
output: html_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(stringr)
library(fixest)
library(rlang)
library(exactextractr)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```



## create data frame
### for each year (e.g. 2023, need to do the following)
### 1. Take the tpu file, collapse on coverage level. get one row per county-year
### 2. calculate additional stats for each county-year: ROI.
### 3. Do the same for previous year. Join columns from previous year for # policies, acreage, liability, ROI
### 4. Also need stats on drought presence/absence from previous year, and missed payout %
### 5. Calculate year-over-year % growth for dependent variables: policies, acreage, liability $
### 6. rbind all years
### 7. perform some checks to make sure data frame is well formed (look at unique values of different columns...)
### filter to remove first year (or few years)
### 8. perform regressions 

```{r}

load("/Users/mjcecil/Documents/GitHub/usdm-prf/data/tpu_all_years.rda")
load("data/grids_usdm_stats.rda")

tx_counties <- st_read(here("data/Tx_CntyBndry_Jurisdictional_TIGER/Tx_CntyBndry_Jurisdictional_TIGER.shp"))
tx_counties <- st_set_crs(tx_counties, 4326)  # Example: EPSG 4326 (WGS84)
  
```

## collapse to unique county-year-intervals

```{r}



# Define column groups
group_cols <- c("Commodity.Year", "County.Code", "Practice.Code")
sum_cols   <- c("Net.Reporting.Level.Amount", "Liability.Amount", "Total.Premium.Amount", "Subsidy.Amount", "Indemnity.Amount", "Liability_cpc_index", "Premium_cpc_index", "Indemnity_cpc_index" )
drop_cols  <- c("Coverage.Level.Percent", "Practice.Name", "Loss.Ratio")  # <-- columns to drop

# Compute remaining columns to take first of
all_cols <- colnames(tpu_all_years)
first_cols <- setdiff(all_cols, c(group_cols, sum_cols, drop_cols))

# Collapse the dataframe


tpu_year_county_interval <- tpu_all_years %>%
  select(-all_of(drop_cols)) %>%
  group_by(across(all_of(group_cols))) %>%
  summarise(
    across(all_of(sum_cols), \(x) sum(x, na.rm = TRUE)),
    across(all_of(first_cols), ~ first(.)),
    .groups = "drop"
  )

```

## collapse to unique county-years, count % of time county is in DM ≥ 3 or DM = 4

```{r}


# Define grouping and summary column sets
group_cols <- c("Commodity.Year", "County.Code")
sum_cols   <- c("Net.Reporting.Level.Amount", "Liability.Amount", "Total.Premium.Amount", "Subsidy.Amount", "Indemnity.Amount", "Liability_cpc_index", "Premium_cpc_index", "Indemnity_cpc_index" )
max_cols   <- c("DM_max_county")

# Compute remaining columns to take first of
all_cols <- colnames(tpu_year_county_interval)
first_cols <- setdiff(all_cols, c(group_cols, sum_cols, max_cols))


tpu_year_county <- tpu_year_county_interval %>%
  group_by(across(all_of(group_cols))) %>%
  summarise(
    # Sum columns
    across(all_of(sum_cols), \(x) sum(x, na.rm = TRUE)),
    
    # Custom DM count logic
    DM_3_pct = mean(DM_max_county >= 3, na.rm = TRUE) * 100,
    DM_4_pct = mean(DM_max_county >= 4, na.rm = TRUE) * 100,
    
    # Max columns
    across(all_of(max_cols), \(x) max(x, na.rm = TRUE)),
    
    # first columns
    across(all_of(first_cols), ~ first(.)),

    .groups = "drop"
  )



```

## alternative calculation of DM percent

```{r}

tx_county_areas <- read.csv(here("data/tx_county-areas/cpc-county-areas.csv"))

tpu_year_county <- tpu_year_county %>%
  rowwise() %>%
  mutate(DM_4_mean_grids_intervals = {
    year <- Commodity.Year
    county <- County.Code

    # Step 1: Filter tx_county_areas for this county
    area_rows <- tx_county_areas %>%
      filter(County.Code == county)

    if (nrow(area_rows) == 0) return(NA_real_)

    # Step 2: Extract unique grid codes
    grid_codes <- unique(area_rows$CPC.Grid.Code)

    # Step 3: Filter grids to those grid codes
    county_rows <- grids %>%
      filter(GRIDCODE %in% grid_codes)

    # Step 4: Select relevant DM_max columns
    pattern <- paste0("^DM_max_", year, "_\\d{3}$")
    matched_cols <- county_rows %>%
      select(matches(pattern)) %>%
      st_drop_geometry()

    if (ncol(matched_cols) == 0) return(NA_real_)

    # Step 5: Proportion of D4 values
    mean(as.matrix(matched_cols) == 4, na.rm = TRUE)
  }) %>%
  ungroup()



```




## calculate additional columns

```{r}

tpu_year_county$ROI <- tpu_year_county$Indemnity_cpc_index/tpu_year_county$Premium_cpc_index

```

## join missed payout count and percent

```{r}

tx_county_areas <- read.csv(here("data/tx_county-areas/cpc-county-areas.csv"))

load("/Users/mjcecil/Documents/GitHub/usdm-prf/data/grids_missed_payouts_dm4.rda")

tpu_year_county$missed_payout_count <- sapply(1:nrow(tpu_year_county), function(x){
  current_year <- tpu_year_county$Commodity.Year[x]
  current_county_id <- tpu_year_county$County.Code[x]
  county_grid_codes <- tx_county_areas %>% filter(County.Code == current_county_id) %>% pull(CPC.Grid.Code) %>% unique()
  missed_payout_count <- grids_missed_payouts %>% filter(year == current_year) %>% filter(GRIDCODE %in% county_grid_codes) %>% pull(missed_payout_count) %>% sum()
})

tpu_year_county$missed_payout_count_per_grid <- sapply(1:nrow(tpu_year_county), function(x){
  current_year <- tpu_year_county$Commodity.Year[x]
  current_county_id <- tpu_year_county$County.Code[x]
  county_grid_codes <- tx_county_areas %>% filter(County.Code == current_county_id) %>% pull(CPC.Grid.Code) %>% unique()
  n_grids <- length(county_grid_codes)
  missed_payout_count <- grids_missed_payouts %>% filter(year == current_year) %>% filter(GRIDCODE %in% county_grid_codes) %>% pull(missed_payout_count) %>% sum()
  missed_payout_rate <- missed_payout_count/n_grids
})

tpu_year_county$missed_payout_pct <- sapply(1:nrow(tpu_year_county), function(x){
  current_year <- tpu_year_county$Commodity.Year[x]
  current_county_id <- tpu_year_county$County.Code[x]
  county_grid_codes <- tx_county_areas %>% filter(County.Code == current_county_id) %>% pull(CPC.Grid.Code) %>% unique()
  missed_payout_pct <- grids_missed_payouts %>% filter(year == current_year) %>% filter(GRIDCODE %in% county_grid_codes) %>% pull(missed_payout_pct) %>% mean()
})

```


## calculate and join previous year statistics

```{r}


# Define your target columns
prev_year_cols <- c("Net.Reporting.Level.Amount", "Liability_cpc_index", "Premium_cpc_index", "Indemnity_cpc_index", "ROI", "DM_3_pct", "DM_4_pct", "DM_max_county", "missed_payout_count", "missed_payout_count_per_grid", "missed_payout_pct"  )

# Step 1: Create a lagged version of the data
tpu_year_county_prev <- tpu_year_county %>%
  select(Commodity.Year, County.Code, all_of(prev_year_cols)) %>%
  mutate(Commodity.Year = Commodity.Year + 1)  # shift forward to match with current year

# Rename lagged columns
tpu_year_county_prev <- tpu_year_county_prev %>%
  rename_with(~ paste0(., "_prev_year"), all_of(prev_year_cols))

# Step 2: Join original df with lagged version
tpu_year_county_joined <- tpu_year_county %>%
  left_join(tpu_year_county_prev, by = c("Commodity.Year", "County.Code"))

# Step 3: Compute YoY columns
for (col in prev_year_cols) {
  current_col <- sym(col)
  prev_col <- sym(paste0(col, "_prev_year"))
  yoy_col <- paste0(col, "_yoy")
  
  tpu_year_county_joined <- tpu_year_county_joined %>%
    mutate(!!yoy_col := {
        ratio <- !!current_col / !!prev_col
        ifelse(is.infinite(ratio), NA, ratio)
      })
}

```








## run model, drought index
```{r}
# Run without fixed effects
feols(Net.Reporting.Level.Amount_yoy ~ DM_4_pct_prev_year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount_yoy ~ DM_4_pct_prev_year | County.Code, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount_yoy ~ DM_4_pct_prev_year | Commodity.Year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount_yoy ~ DM_4_pct_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)

```

## run model, ROI
```{r}

# tpu_year_county_joined$Net.Reporting.Level.Amount_yoy <- 100 * tpu_year_county_joined$Net.Reporting.Level.Amount_yoy

# Run without fixed effects
# feols(Net.Reporting.Level.Amount_yoy ~ ROI_prev_year, data = tpu_year_county_joined)
# 
# feols(Net.Reporting.Level.Amount_yoy ~ ROI_prev_year | County.Code, data = tpu_year_county_joined)
# 
# feols(Net.Reporting.Level.Amount_yoy ~ ROI_prev_year | Commodity.Year, data = tpu_year_county_joined)
# 
# feols(Net.Reporting.Level.Amount_yoy ~ ROI_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)
# 
# feols(Net.Reporting.Level.Amount_yoy ~ missed_payout_pct_prev_year, data = tpu_year_county_joined)
# 
# feols(Net.Reporting.Level.Amount_yoy ~ missed_payout_pct_prev_year | County.Code, data = tpu_year_county_joined)
# 
# feols(Net.Reporting.Level.Amount_yoy ~ missed_payout_pct_prev_year | Commodity.Year, data = tpu_year_county_joined)
# 
# feols(Net.Reporting.Level.Amount_yoy ~ missed_payout_pct_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount_yoy ~ DM_4_pct_prev_year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount_yoy ~ DM_4_pct_prev_year | County.Code, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount_yoy ~ DM_4_pct_prev_year | Commodity.Year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount_yoy ~ DM_4_pct_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)


# feols(log(Net.Reporting.Level.Amount_yoy) ~ DM_4_pct_prev_year, data = tpu_year_county_joined)
# 
# feols(log(Net.Reporting.Level.Amount_yoy) ~ DM_4_pct_prev_year | County.Code, data = tpu_year_county_joined)
# 
# feols(log(Net.Reporting.Level.Amount_yoy) ~ DM_4_pct_prev_year | Commodity.Year, data = tpu_year_county_joined)
# 
# feols(log(Net.Reporting.Level.Amount_yoy) ~ DM_4_pct_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)

```

```{r}
# Run without fixed effects
feols(Net.Reporting.Level.Amount ~ ROI_prev_year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ ROI_prev_year | County.Code, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ ROI_prev_year | Commodity.Year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ ROI_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ missed_payout_pct_prev_year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ missed_payout_pct_prev_year | County.Code, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ missed_payout_pct_prev_year | Commodity.Year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ missed_payout_pct_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ DM_4_pct_prev_year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ DM_4_pct_prev_year | County.Code, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ DM_4_pct_prev_year | Commodity.Year, data = tpu_year_county_joined)

feols(Net.Reporting.Level.Amount ~ DM_4_pct_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)

```

```{r}
# Run without fixed effects
feols(log(Net.Reporting.Level.Amount) ~ ROI_prev_year, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ ROI_prev_year | County.Code, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ ROI_prev_year | Commodity.Year, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ ROI_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ missed_payout_pct_prev_year, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ missed_payout_pct_prev_year | County.Code, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ missed_payout_pct_prev_year | Commodity.Year, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ missed_payout_pct_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ DM_4_pct_prev_year, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ DM_4_pct_prev_year | County.Code, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ DM_4_pct_prev_year | Commodity.Year, data = tpu_year_county_joined)

feols(log(Net.Reporting.Level.Amount) ~ DM_4_pct_prev_year | County.Code + Commodity.Year, data = tpu_year_county_joined)
```

## non-regression comparisons
## what is typical year over year growth in acreage?
## split to with a drought vs without (DM3, DM4), what is typical growth
## split to E vs W Texas, with and without drought
## split by interval



```{r}

mean_yoy <- mean(tpu_year_county_joined$Net.Reporting.Level.Amount_yoy, na.rm = T)
median_yoy <- median(tpu_year_county_joined$Net.Reporting.Level.Amount_yoy, na.rm = T)

mean_drought_4_yoy <- tpu_year_county_joined %>%
  filter(DM_max_county_prev_year == 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(., na.rm = T)

median_drought_4_yoy <- tpu_year_county_joined %>%
  filter(DM_max_county_prev_year == 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(., na.rm = T)

mean_non_drought_4_yoy <- tpu_year_county_joined %>%
  filter(DM_max_county_prev_year < 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(., na.rm = T)

median_non_drought_4_yoy <- tpu_year_county_joined %>%
  filter(DM_max_county_prev_year < 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(., na.rm = T)

mean_drought_3_yoy <- tpu_year_county_joined %>%
  filter(DM_max_county_prev_year >= 3) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(., na.rm = T)

median_drought_3_yoy <- tpu_year_county_joined %>%
  filter(DM_max_county_prev_year >= 3) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(., na.rm = T)

mean_non_drought_3_yoy <- tpu_year_county_joined %>%
  filter(DM_max_county_prev_year < 3) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(., na.rm = T)

median_non_drought_3_yoy <- tpu_year_county_joined %>%
  filter(DM_max_county_prev_year < 3) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(., na.rm = T)

median_cols <- c("Net.Reporting.Level.Amount_yoy")

a <- tpu_year_county_joined %>% 
  select(County.Code, Net.Reporting.Level.Amount_yoy) %>%
  group_by(County.Code) %>% 
  summarise(across(all_of(median_cols), \(x) median(x, na.rm = TRUE)))

b <- tpu_year_county_joined %>% 
  select(Commodity.Year, Net.Reporting.Level.Amount_yoy) %>%
  group_by(Commodity.Year) %>% 
  summarise(across(all_of(median_cols), \(x) median(x, na.rm = TRUE)))

tx_counties$COUNTYID <- as.numeric(tx_counties$COUNTYID)

tx_counties <- tx_counties %>%
  left_join(
    a,
    by = c( "COUNTYID" = "County.Code")
  )

# Define bin breaks and labels
breaks <- c(-Inf, seq(0.7, 1.3, by = 0.1), Inf)
labels <- c("<0.7", "0.7–0.8", "0.8–0.9", "0.9–1.0", "1.0–1.1", "1.1–1.2", "1.2–1.3", ">1.3")

# Create a binned factor column
tx_counties <- tx_counties %>%
  mutate(Net_Reporting_Bin = cut(Net.Reporting.Level.Amount_yoy, breaks = breaks, labels = labels, include.lowest = TRUE))

# Plot the map
ggplot(tx_counties) +
  geom_sf(aes(fill = Net_Reporting_Bin), color = "white", size = 0.1) +
  scale_fill_viridis_d(option = "plasma", na.value = "grey90") +
  labs(
    title = "Net Reporting Level (Median YoY)",
    fill = "YoY Bin"
  ) +
  theme_minimal()

# plot time-series
ggplot(b) + 
  geom_point(aes(x = Commodity.Year, y = Net.Reporting.Level.Amount_yoy)) +
  geom_line(aes(x = Commodity.Year, y = Net.Reporting.Level.Amount_yoy)) +
  theme_bw() +
  ylab("Acreage YoY") +
  xlab("year") +
  ggtitle("Year over Year acreage growth")


```

## East vs West

### East vs West Texas classification (counties)

```{r}

tx_counties <- st_read(here("data/Tx_CntyBndry_Jurisdictional_TIGER/Tx_CntyBndry_Jurisdictional_TIGER.shp"))
tx_counties <- st_set_crs(tx_counties, 4326)  # Example: EPSG 4326 (WGS84)
  
koppen_grid <- raster(here("data/koppen-climate-classification/koppen_ics/w001001.adf"))

tx_counties_rpj <- st_transform(tx_counties, crs = crs(koppen_grid))

## get unique classifications in TX
summary_stats <- exact_extract(koppen_grid, tx_counties_rpj %>% st_union(), function(values, coverage_fraction) {
    table(values)
}, progress = TRUE)

## 1 value is for "B" classifications (arid) 
## 2 value is for "C" classification (temperate)
reclass_df <- data.frame(c(7, 8, 15, 18, 22, 23),
                         c(1, 1, 2, 2, 2, 2))
koppen_grid_rcl <- reclassify(koppen_grid, rcl = as.matrix(reclass_df))

## extract values for TX counties
summary_stats_rcl <- exact_extract(
    koppen_grid_rcl,
    tx_counties_rpj,
    function(values, coverage_fraction) {
        tbl <- table(values)
        as.numeric(names(tbl)[which.max(tbl)])
    },
    progress = TRUE
)

tx_counties_rpj$region <- summary_stats_rcl
tx_counties$region <- summary_stats_rcl

east_tx_county_codes <- tx_counties %>% filter(region == 2) %>% pull(COUNTYID) %>% as.numeric()
west_tx_county_codes <- tx_counties %>% filter(region == 1) %>% pull(COUNTYID) %>% as.numeric()

plot(tx_counties_rpj['region'])

```

## join region data

```{r}

tx_counties_rpj$COUNTYID <- as.numeric(tx_counties_rpj$COUNTYID)

tpu_year_county_joined <- tpu_year_county_joined %>%
  left_join(
    tx_counties_rpj %>% 
      select(COUNTYID, region) %>% 
      st_drop_geometry(),
    by = c("County.Code" = "COUNTYID")
  )


```



```{r}

## e v w

mean_yoy_e <- tpu_year_county_joined %>% filter(region == 2) %>% 
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(na.rm = T)
mean_yoy_w <- tpu_year_county_joined %>% filter(region == 1) %>% 
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(na.rm = T)
median_yoy_e <- tpu_year_county_joined %>% filter(region == 2) %>% 
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(na.rm = T)
median_yoy_w <- tpu_year_county_joined %>% filter(region == 1) %>% 
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(na.rm = T)

mean_drought_4_yoy_e <- tpu_year_county_joined %>%
  filter(region == 2) %>% 
  filter(DM_max_county_prev_year == 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(., na.rm = T)

median_drought_4_yoy_e <- tpu_year_county_joined %>%
  filter(region == 2) %>% 
  filter(DM_max_county_prev_year == 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(., na.rm = T)

mean_non_drought_4_yoy_e <- tpu_year_county_joined %>%
  filter(region == 2) %>% 
  filter(DM_max_county_prev_year < 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(., na.rm = T)

median_non_drought_4_yoy_e <- tpu_year_county_joined %>%
  filter(region == 2) %>% 
  filter(DM_max_county_prev_year < 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(., na.rm = T)

mean_drought_4_yoy_w <- tpu_year_county_joined %>%
  filter(region == 1) %>% 
  filter(DM_max_county_prev_year == 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(., na.rm = T)

median_drought_4_yoy_w <- tpu_year_county_joined %>%
  filter(region == 1) %>% 
  filter(DM_max_county_prev_year == 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(., na.rm = T)

mean_non_drought_4_yoy_w <- tpu_year_county_joined %>%
  filter(region == 1) %>% 
  filter(DM_max_county_prev_year < 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  mean(., na.rm = T)

median_non_drought_4_yoy_w <- tpu_year_county_joined %>%
  filter(region == 1) %>% 
  filter(DM_max_county_prev_year < 4) %>%
  pull(Net.Reporting.Level.Amount_yoy) %>% 
  median(., na.rm = T)



```


## look at with, without drought
```{r}

## with drought
median_cols <- c("Net.Reporting.Level.Amount_yoy")

a <- tpu_year_county_joined %>% 
  filter(DM_max_county_prev_year == 4) %>% 
  select(County.Code, Net.Reporting.Level.Amount_yoy) %>%
  group_by(County.Code) %>% 
  summarise(across(all_of(median_cols), \(x) median(x, na.rm = TRUE)))

tx_counties$COUNTYID <- as.numeric(tx_counties$COUNTYID)

tx_counties <- tx_counties %>%
  left_join(
    a,
    by = c( "COUNTYID" = "County.Code")
  )

# Define bin breaks and labels
breaks <- c(-Inf, seq(0.7, 1.3, by = 0.1), Inf)
labels <- c("<0.7", "0.7–0.8", "0.8–0.9", "0.9–1.0", "1.0–1.1", "1.1–1.2", "1.2–1.3", ">1.3")

# Create a binned factor column
tx_counties <- tx_counties %>%
  mutate(Net_Reporting_Bin = cut(Net.Reporting.Level.Amount_yoy, breaks = breaks, labels = labels, include.lowest = TRUE))

# Plot the map
ggplot(tx_counties) +
  geom_sf(aes(fill = Net_Reporting_Bin), color = "white", size = 0.1) +
  scale_fill_viridis_d(option = "plasma", na.value = "grey90") +
  labs(
    title = "Net Reporting Level (YoY) under Drought Conditions",
    fill = "YoY Bin"
  ) +
  theme_minimal()


```

## without drought

```{r}


## with drought
median_cols <- c("Net.Reporting.Level.Amount_yoy")

a <- tpu_year_county_joined %>% 
  filter(DM_max_county_prev_year < 4) %>% 
  select(County.Code, Net.Reporting.Level.Amount_yoy) %>%
  group_by(County.Code) %>% 
  summarise(across(all_of(median_cols), \(x) median(x, na.rm = TRUE)))

tx_counties$COUNTYID <- as.numeric(tx_counties$COUNTYID)

tx_counties <- tx_counties %>%
  left_join(
    a,
    by = c( "COUNTYID" = "County.Code")
  )

# Define bin breaks and labels
breaks <- c(-Inf, seq(0.7, 1.3, by = 0.1), Inf)
labels <- c("<0.7", "0.7–0.8", "0.8–0.9", "0.9–1.0", "1.0–1.1", "1.1–1.2", "1.2–1.3", ">1.3")

# Create a binned factor column
tx_counties <- tx_counties %>%
  mutate(Net_Reporting_Bin = cut(Net.Reporting.Level.Amount_yoy, breaks = breaks, labels = labels, include.lowest = TRUE))

# Plot the map
ggplot(tx_counties) +
  geom_sf(aes(fill = Net_Reporting_Bin), color = "white", size = 0.1) +
  scale_fill_viridis_d(option = "plasma", na.value = "grey90") +
  labs(
    title = "Net Reporting Level (YoY) under Drought Conditions",
    fill = "YoY Bin"
  ) +
  theme_minimal()



```

## bar chart for Yoy growth

```{r}

# Create a named vector
combo_totals <- c(
  "All county-years" = median_yoy,
  "No exceptional drought" = median_non_drought_4_yoy,
  "Exceptional Drought" = median_drought_4_yoy
)


library(ggplot2)

# Create a data frame, preserving order
combo_df <- data.frame(
  group = factor(names(combo_totals), levels = names(combo_totals)),
  value = as.numeric(combo_totals)
)

# Create a "height above 1" column
combo_df$bar_height <- combo_df$value - 1

# Plot with manual baseline at y = 1 using geom_rect
ggplot(combo_df, aes(x = group)) +
  geom_rect(aes(
    xmin = as.numeric(group) - 0.4,
    xmax = as.numeric(group) + 0.4,
    ymin = 1,
    ymax = value
  ), fill = "steelblue") +
  geom_text(aes(x = group, y = value + 0.005, label = paste0(round(value * 100), "%")),
            size = 4.5) +
  scale_y_continuous(limits = c(1, max(combo_df$value) + 0.05), expand = c(0, 0)) +
  labs(
    title = "Median Year-over-Year Acreage Change",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(size = 16, face = "bold")
  )


```

## plot of dm yearly county pct

```{r}

tx_counties <- st_read(here("data/Tx_CntyBndry_Jurisdictional_TIGER/Tx_CntyBndry_Jurisdictional_TIGER.shp"))
tx_counties <- st_set_crs(tx_counties, 4326)  # Example: EPSG 4326 (WGS84)
  

a <- tpu_year_county_joined %>% 
  filter(year == 2022) %>% 
  select(County.Code, DM_4_mean_grids_intervals)

tx_counties$COUNTYID <- as.numeric(tx_counties$COUNTYID)

tx_counties <- tx_counties %>%
  left_join(
    a,
    by = c( "COUNTYID" = "County.Code")
  )




# Step 1: Create bins by 10%, with last bin >50%
tx_counties <- tx_counties %>%
  mutate(D4_bin = cut(
    DM_4_mean_grids_intervals,
    breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, Inf),
    labels = c("0–10%", "10–20%", "20–30%", "30–40%", "40–50%", ">50%"),
    include.lowest = TRUE,
    right = FALSE
  ))

# Step 2: Plot with reversed "plasma" palette to make purple the highest
ggplot(tx_counties) +
  geom_sf(aes(fill = D4_bin), color = "white", size = 0.1) +
  scale_fill_viridis_d(
    option = "plasma",
    direction = -1,   # Reverse palette to make purple highest
    na.value = "grey90"
  ) +
  labs(
    title = "2022 Exceptional Drought %",
    fill = "D4 %"
  ) +
  theme_minimal()

```



