---
title: "ARER workflow"
author: Michael Cecil
output: html_notebook
---

This notebook performs the rainfall index and payout calculations for much of the manuscript.

## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(tigris)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```

## Inputs

```{r}

start_year_payouts <- 2010
end_year_payouts <- 2025

```


## Payouts

First determine payouts for given inputs (year, grid, interval, coverage level, productivity factor, grazing/haying, )

Assumptions made:
- constant productivity factor of 1
- rates are constant across years
- rates are consistent despite changing temporal baseline or spatial resolution
- acreage proportionally split among cells in county
- liability, premia, payouts calculated based on acreage for each interval

Equations used:
Dollar Amount of Insurance Per Acre = County-Base Value * Productivity Factor (also = Liability per Acre?)
Liability = Dollar Amount * # Acres
Premium per acre = Liability per acre * rate 
Payment factor = (trigger_threshold - RI)/trigger_threshold
Indemnity = Liability * payment factor 

Columns to add
CHIRPS data set
- rates, 
- payment factor for each interval x year


## load rates data, join to grids_intersecting

Note - the rates files have a funny format. They are actually downloading two pieces of information. The Premium.Rate is based on the coverage level (70%, 75% ... 90% etc), while the County.Base.Value is based on the use type (Grazing, Haying non irrigated etc).

We will just be focusing on Grazing for now as that is the largest use case. 

```{r}

load(file = paste0(here("data/outputs/grids_intersecting.rda")))

## get state names in usable vector
states <- unique(grids_intersecting$STATE)
full_names <- state.name[match(states, state.abb)]
lower_hyphen_names <- gsub(" ", "-", tolower(full_names)) # Convert to lowercase + hyphenated

## find rates files

files <- list.files(
  path = here("data/rates"),
  recursive = TRUE,
  full.names = TRUE
)

rates_files <- files[grepl(
  paste(full_names, collapse = "|"),
  basename(files),
  ignore.case = TRUE
)]





## create single df with all rates information
rates_df <- lapply(rates_files, function(x){
  print(x)
  file_path_length <- length(strsplit(x, "/")[[1]])
  file_name_length <- length(strsplit(strsplit(x, "/")[[1]][file_path_length], "-")[[1]])
  coverage_level <- strsplit(strsplit(x, "/")[[1]][file_path_length], "-")[[1]][file_name_length - 1] %>% as.numeric()
  # print(strsplit(strsplit(x, "/")[[1]][file_path_length], "-")[[1]])
  # print(file_name_length)
  use_level <- strsplit(strsplit(x, "/")[[1]][file_path_length], "-")[[1]][3:(file_name_length - 2)] 
  use_level <- paste(use_level, collapse = "-")
  a <- read.csv(x)
  a$coverage <- coverage_level
  a$use <- use_level
  return(a)
}) %>% do.call('rbind', .)


## add rates for CPC grids and intervals 
## includes a check to confirm that if a grid is listed multiple times 
## (with the same coverage level), then it must have the same premium rate


for (interval in 625:635){
  for(coverage_level in seq(70, 90, 5)){

    rates_df_filt <- rates_df %>% 
      filter(Interval.Code == interval,
             coverage == coverage_level)

    # Validate duplicates
    dup_check <- rates_df_filt %>%
      group_by(CPC.Grid.Code) %>%
      summarise(n_rates = n_distinct(Premium.Rate), .groups = "drop") %>%
      filter(n_rates > 1)

    if (nrow(dup_check) > 0) {
      stop(paste(
        "Conflicting Premium.Rate values found for interval", interval,
        "coverage", coverage_level,
        "CPC grids:",
        paste(dup_check$CPC.Grid.Code, collapse = ", ")
      ))
    }

    # Deduplicate safely
    rates_df_filt <- rates_df_filt %>%
      distinct(CPC.Grid.Code, .keep_all = TRUE)

    rate_lookup_table <- setNames(
      rates_df_filt$Premium.Rate,
      rates_df_filt$CPC.Grid.Code
    )

    grids_intersecting[[paste0(
      "rate_interval_", interval, "_coverage_", coverage_level
    )]] <- rate_lookup_table[as.character(grids_intersecting$GRIDCODE)]
  }
}


```

## save grids_intersecting to file with rates included 
Also remove NA values, for grid cells that were not included in rates files

```{r}

grids_intersecting_rates <- grids_intersecting %>% filter(!is.na(rate_interval_625_coverage_70))

save(grids_intersecting_rates, 
     file = here("data/grids_intersecting_rates.rda"))

```

## create CPC area proportion files
output table has columns "State.Code"     "County.Code"    "CPC.Grid.Code"  "Area"           "Proportion"    

```{r}

## download counties from tigris
state_fips <- fips_codes %>%   # built into tigris
  filter(state %in% states) %>%
  distinct(state, state_code)

counties_sf <- counties(
  state = states,
  cb = TRUE,      # cartographic boundary (simpler, faster)
  year = 2022
)

aea_crs <- 5070  # US Albers Equal Area

counties_sf <- st_transform(counties_sf, aea_crs)
grids_sf    <- st_transform(grids_intersecting_rates, aea_crs)

county_grid_intersections <- st_intersection(
  counties_sf,
  grids_sf
)

county_grid_intersections <- county_grid_intersections %>%
  mutate(
    Area = as.numeric(st_area(geometry))
  )

area_table <- county_grid_intersections %>%
  group_by(STATEFP, COUNTYFP) %>%
  mutate(
    Proportion = Area / sum(Area)
  ) %>%
  ungroup()

cpc_county_areas <- area_table %>%
  mutate(
    # ---- Your requested schema ----
    State.Code    = as.numeric(STATEFP),  # numeric state FIPS
    State         = STUSPS,               # TX, ND, etc.
    County.Code   = COUNTYFP,             # 3-digit county FIPS
    CPC.Grid.Code = GRIDCODE
  ) %>%
  st_drop_geometry() %>%
  select(
    State.Code,
    State,
    County.Code,
    CPC.Grid.Code,
    Area,
    Proportion
  )

cpc_county_areas$County.Code <- as.numeric(cpc_county_areas$County.Code)


save(cpc_county_areas, file = here("data/outputs/cpc_county_areas.rda"))

```

## validation check

```{r}
sums_by_county <- cpc_county_areas %>%
  group_by(State.Code, County.Code) %>%
  summarise(sum_prop = sum(Proportion))

```








## calculate payouts
## we iterate through TPU file, and calculate payouts for different rainfall indices
## below chunk contains payment_calcs function used to calculate payouts

```{r}

load_rda_to_variable <- function(file_path, variable_name) {
  # Load the .rda file
  loaded_objects <- load(file_path)
  
  # Assign the loaded object to the specified variable name
  if (length(loaded_objects) != 1) {
    stop("The .rda file must contain exactly one object.")
  }
  
  assign(variable_name, get(loaded_objects), envir = .GlobalEnv)
}

payment_calcs <- function(year, 
                          use_type = "Grazing",
                          index_prefix = "cpc_index", ## index to calculate indemnities for
                          resolution = "CPC",
                          county_areas_file_path = "data/outputs/cpc_county_areas.rda",
                          proportion_column = "Proportion",  
                          input_grid_data = grids_intersecting_rates,
                          CDL = F){

  year_2 <- substr(year, 3, 4)
  tpu_names <- read.csv(here("data/PRF_tpu/tpu_names.csv"))
  tpu <- read.csv(here(paste0("data/PRF_tpu/SOBSCCTPU", year_2, ".TXT")), sep = "|", header = F)
  names(tpu) <- names(tpu_names)
  
  tpu$Type.Name <- trimws(tpu$Type.Name) # remove white space

  if(year <= 2012){ ## 2012 had different type names
    tpu[tpu$Type.Name == "Grazingland", "Type.Name"] <- "Grazing"
    tpu[tpu$Type.Name == "Hayland", "Type.Name"] <- "Haying"
    tpu[tpu$Type.Name == "GRAZINGLAND", "Type.Name"] <- "Grazing"
    tpu[tpu$Type.Name == "HAYLAND", "Type.Name"] <- "Haying"
  }
  
  print(nrow(tpu))
  print(tpu[1,])

  tpu$Commodity.Code <- as.numeric(tpu$Commodity.Code)

  
  tpu <- tpu %>% 
    filter(State.Abbreviation %in% states) %>% 
 #   filter(Insurance.Plan.Abbreviation == "RI") %>% 
    filter(Liability.Amount > 0) %>% 
    filter(Commodity.Code == 88) %>% ## for PRF
    filter(Type.Name == use_type)# %>% 
 #   filter(Reporting.Level.Type == "Acres") 
  
  print(tpu[1,])
  print(nrow(tpu))

  ## add columns to store outputs
  ri_premium_column <- paste0("Premium_", index_prefix)
  ri_indemnity_column <- paste0("Indemnity_", index_prefix)
  tpu[[ri_premium_column]] <- NA
  tpu[[ri_indemnity_column]] <- NA
  
  ## filter to only non-organic non-irrigated haying
  if(use_type == "Haying" & year >= 2011){
    tpu <- tpu %>% filter(Practice.Code %in% (c(525:535, 625:635))) 
  }
  
  cbv_use = c("Grazing" = "grazing", "Haying" = "hayingnonirrigatednonorganic") ## dictionary to use with rates files
  
  tpu$CountyBaseValue <- mapply(
    function(state_code, county_code){
    
      rates_df_filtered <- rates_df %>%
        dplyr::filter(
          State.Code  == state_code,
          County.Code == county_code,
          use == cbv_use[use_type]
        )
    
      # ---- Case 1: No values ----
      if (nrow(rates_df_filtered) == 0) {
    
        message(
          paste0(
            "No County.Base.Value for State.Code=",
            state_code,
            ", County.Code=",
            county_code
          )
        )
    
        return(NA_real_)
      }
    
      # ---- Case 2: Multiple values (keep your assertion) ----
      assertthat::assert_that(
        length(unique(rates_df_filtered$County.Base.Value)) == 1
      )
    
      # ---- Case 3: Exactly one value ----
      rates_df_filtered$County.Base.Value[1]
    
    },
    tpu$State.Code,
    tpu$County.Code
  )

  ri_liability_column <- paste0("Liability_", index_prefix)
  tpu[[ri_liability_column]] <- tpu$CountyBaseValue * tpu$Net.Reporting.Level.Amount
  
  ## iterate through rows of tpu data one at a time, calculating alternate payouts
  ##  for each tpu row, create mini df of overlapping CPC/CHIRPS grid cells with proportions
  for(k in 1:nrow(tpu)){
    current_county_code <- tpu$County.Code[k]
    current_state_code <- tpu$State.Code[k]
    total_liability <- tpu[[ri_liability_column]][k]
    current_interval <- tpu$Practice.Code[k]
    
    if(use_type == "Haying" && year >=2016){## before 2016, the haying intervals are 625:635. Starting 2016 they are 525:535
      current_interval <- current_interval + 100
    }
    
   ## in certain years there were intervals 221... 226, represeting Jan-Feb, Mar-Apr
   ## also in years 2010... there were intervals 648 representing Apr-June ... 652 representing Aug-Oct. we've assigned these to the last two months
    interval_dict <- c("221" = 625, "222" = 627, "223" = 629, "224" = 631, "225" = 633, "226" = 635, "648" = 629, "649" = 630, "650" = 631, "651" = 632, "652" = 633)
      
     
    if (as.character(current_interval) %in% names(interval_dict)){
      current_interval <- interval_dict[as.character(current_interval)]
      tpu$Practice.Code[k] <- current_interval
    }
      ## update practice code
    
    

    
    
    
    current_coverage <- (tpu$Coverage.Level.Percent[k]) * 100
    trigger_grid_index <- tpu$Coverage.Level.Percent[k]
    
    ## get RI column
    current_interval_ri_column <- paste0(index_prefix, "_", year, "_", current_interval)

    if(!(current_interval_ri_column %in% names(input_grid_data))){
      print(current_interval_ri_column)
      print(current_interval)
      print(tpu$Practice.Code[k])
      print(k)
      # print(names(input_grid_data))
      assert_that(current_interval_ri_column %in% names(input_grid_data))
    }
    
    ## get rate column
    current_interval_rate_column <- paste0("rate_interval_", current_interval, "_coverage_", current_coverage)
    assert_that(current_interval_rate_column %in% names(input_grid_data))

    
    
    ## calculate area-weighted (by grid) premium and payout
    load_rda_to_variable(here(county_areas_file_path), "cpc_county_areas")
    cpc_county_areas$Proportion <- cpc_county_areas[[proportion_column]]

    ## create mini_df of grid cells within county for area weighting
    mini_df <- cpc_county_areas %>% filter(County.Code == current_county_code) %>% 
      filter(State.Code == current_state_code)
    mini_df$total_liability <- total_liability
    total_area <- sum(mini_df$Area)
    mini_df$weighted_liability <- mini_df$total_liability * mini_df$Proportion
    
    input_grid_data_dedupe <- input_grid_data[!duplicated(input_grid_data$GRIDCODE), ]
    ## merge w input_grid_data, get columns for ri, rating
    mini_df <- merge(mini_df, input_grid_data_dedupe[, c("GRIDCODE", 
                                              current_interval_ri_column, 
                                              current_interval_rate_column)],
                     by.x = "CPC.Grid.Code",
                     by.y = "GRIDCODE",
                     all.x = T)
  #  print(mini_df)
    mini_df$weighted_premium <- mini_df$weighted_liability * mini_df[[current_interval_rate_column]]
    
    mini_df$payment_factor <- pmax(0, 
                                  (trigger_grid_index - mini_df[[current_interval_ri_column]])/trigger_grid_index)
    mini_df$weighted_indemnity <- mini_df$weighted_liability * mini_df$payment_factor
   
    weighted_premium <- sum(mini_df$weighted_premium, na.rm = T)
    tpu[[ri_premium_column]][k] <- weighted_premium
    tpu[[ri_indemnity_column]][k] <- sum(mini_df$weighted_indemnity, na.rm = T) 
    
    
  }

  
  


  save(tpu, file = here(paste0("data/outputs/payouts/tpu_", use_type, "_", year, "_", index_prefix, ".rda")))
  print(paste0("processing ", year, " ", index_prefix, " completed"))
  return(tpu)
  

}

```




## below chunk loops through each year
## it calculates payouts for the following options
### Grazing - CPC index 


```{r}

payouts_outputs_folder <- here("data/outputs/payouts")

if (!dir.exists(payouts_outputs_folder)) {
  dir.create(payouts_outputs_folder)
}

for(year in start_year_payouts:end_year_payouts){
  print(year)

  a4 <- payment_calcs(year,
              use_type = "Grazing",
              index_prefix = "cpc_index",
              resolution = "CPC",
              county_areas_file_path = "data/outputs/cpc_county_areas.rda",
              input_grid_data = grids_intersecting_rates,
              proportion_column = "Proportion")

}

```


## count and remove NA values (due to missing county base values)

```{r}

for(year in start_year_payouts:end_year_payouts){
  print(year)
    load_rda_to_variable(here(paste0("data/outputs/payouts/tpu_Grazing_", year, "_cpc_index.rda")), 
                       "tpu_with_na")
  print(paste0("NA count ", sum(is.na(tpu_with_na$CountyBaseValue)) ))
  save(tpu_with_na, file = here(paste0("data/outputs/payouts/tpu_Grazing_", year, "_cpc_index_W_NA.rda")))
  
  tpu <- tpu_with_na %>% filter(!is.na(CountyBaseValue))
  print(nrow(tpu_with_na))
  print(nrow(tpu))
  
  save(tpu, file = here(paste0("data/outputs/payouts/tpu_Grazing_", year, "_cpc_index.rda")))

}



```

## consolidate statistics for each year
## create year_df, a data frame listing liabilities/premia/indemnities
## from the statement of business files (raw)
## and comparing payouts between cpc_index estimated and "raw" payouts from SOB files
```{r}

year_df <- data.frame("year" = start_year_payouts:end_year_payouts)
year_df$grazing_raw_liability <- NA
year_df$grazing_raw_premium <- NA
year_df$grazing_estimated_liability <- NA
year_df$grazing_cpc_premium <- NA
year_df$grazing_cpc_indemnity <- NA


for(k in 1:nrow(year_df)){
  year <- year_df$year[k]
  
  load_rda_to_variable(here(paste0("data/outputs/payouts/tpu_Grazing_", year, "_cpc_index.rda")), 
                       "df")
  

  year_df$grazing_cpc_indemnity[k] <- sum(df$Indemnity_cpc_index)
  year_df$grazing_raw_indemnity[k] <- sum(df$Indemnity.Amount)
  year_df$grazing_cpc_premium[k] <- sum(df$Premium_cpc_index)
  year_df$grazing_raw_premium[k] <- sum(df$Total.Premium.Amount)
  year_df$grazing_estimated_liability[k] <- sum(df$Liability_cpc_index)
  year_df$grazing_raw_liability[k] <- sum(df$Liability.Amount)

}


# ## combined grazing + haying indemnities
# year_df$combined_raw_indemnity <- year_df$grazing_raw_indemnity + year_df$haying_raw_indemnity
# year_df$combined_cpc_indemnity <- year_df$grazing_cpc_indemnity + year_df$haying_cpc_indemnity
# 
# 
# ## combined ratios
# 
# year_df$combined_ratio_cpc_raw <- round(year_df$combined_cpc_indemnity/year_df$combined_raw_indemnity, 2)



## changes in indemnities, premia
# grazing indemnity
year_df$grazing_ratio_indemnity_grazing_cpc <- round((year_df$grazing_cpc_indemnity)/year_df$grazing_raw_indemnity, 3)

# grazing premia
year_df$grazing_ratio_premium_grazing_cpc <- round((year_df$grazing_cpc_premium)/year_df$grazing_raw_premium, 3)

## cost ratios
year_df$grazing_cr_raw <- round(year_df$grazing_raw_indemnity/ year_df$grazing_raw_premium, 3)
year_df$grazing_cr_cpc <- round(year_df$grazing_cpc_indemnity/ year_df$grazing_cpc_premium, 3)

## cost ratio comparisons
year_df$grazing_cr_comparison_cpc <- round(year_df$grazing_cr_cpc/ year_df$grazing_cr_raw, 3)


year_df$grazing_liability_ratio <- round(year_df$grazing_estimated_liability/year_df$grazing_raw_liability, 3)

write.csv(year_df, file = here("data/outputs/payouts/payouts_by_year.csv"))

```
