---
title: "Untitled"
output: html_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(stringr)
library(tidyr)
library(gt)
library(viridis)
library(scales)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```



## initial plots

## summary stats
###mean RI by year

```{r}
load(here("data/outputs/grids_usdm_stats.rda"))

grids <- grids %>% filter(!is.na(cpc_index_2010_625)) %>% filter(!(is.na(DM_median_2010_625)))

ri_threshold <- 0.9
high_drought_threshold <- 4
high_drought_label <-"D4"

start_year <- 2010
end_year <- 2025

aoi_states <- c("CA", "NV", "AZ", "NM", "CO", "UT", "TX", "OK", "KS")

interval_start_months <- 1:11
names(interval_start_months) <- 625:635

interval_end_months <- 2:12
names(interval_end_months) <- 625:635

```

## function for exporting tables

```{r}

export_df_to_latex <- function(df, file_path, caption = NULL) {
  
  #-----------------------------
  # Helper: escape LaTeX specials
  #-----------------------------
  escape_latex <- function(x) {
    x <- gsub("\\\\", "\\\\textbackslash{}", x)
    x <- gsub("([%&_#\\$])", "\\\\\\1", x)
    x <- gsub("_", "\\\\_", x)
    return(x)
  }
  
  # Convert to character
  df[] <- lapply(df, as.character)
  
  # Escape LaTeX characters
  df[] <- lapply(df, escape_latex)
  colnames(df) <- escape_latex(colnames(df))
  
  # Alignment: first left, rest right
  n_cols <- ncol(df)
  align_string <- paste0("l", paste(rep("r", n_cols - 1), collapse = ""))
  
  # Build body rows
  body_rows <- apply(df, 1, function(row) {
    paste(row, collapse = " & ")
  })
  
  body_rows <- paste0(body_rows, " \\\\")
  body_text <- paste(body_rows, collapse = "\n")
  
  # Header row
  header <- paste(colnames(df), collapse = " & ")
  header <- paste0(header, " \\\\")
  
  # Caption + label handling
  file_name <- tools::file_path_sans_ext(basename(file_path))
  
  if (is.null(caption)) {
    caption_text <- gsub("_", " ", file_name)
  } else {
    caption_text <- caption
  }
  
  label <- paste0("tab:", file_name)
  
  # Build LaTeX table
  latex_table <- paste0(
"\\begin{table}[t]
\\centering
\\fontsize{12pt}{14pt}\\selectfont
\\begin{tabular}{@{\\extracolsep{\\fill}}", align_string, "}
\\toprule
", header, "
\\midrule\\addlinespace[2.5pt]
", body_text, "
\\bottomrule
\\end{tabular}
\\caption{", caption_text, "}
\\label{", label, "}
\\end{table}"
  )
  
  # Write file
  writeLines(latex_table, file_path)
  
  message("LaTeX table exported to: ", file_path)
}



```



## Summary stats for RI (across all grid-intervals)

```{r}
years <- start_year:end_year
intervals <- 625:635

RI_cols <- expand.grid(years, intervals) |>
  apply(1, \(x) paste0("cpc_index_", x[1], "_", x[2]))

DM_cols <- expand.grid(years, intervals) |>
  apply(1, \(x) paste0("DM_median_", x[1], "_", x[2]))




#-----------------------------
# 1. Long format
#-----------------------------
df_long <- grids %>%
  select(STATE, any_of(RI_cols)) %>%
  st_drop_geometry() %>% 
  pivot_longer(
    cols = -STATE,
    names_to = "grid_interval",
    values_to = "RI"
  )

#-----------------------------
# 2. Classify payout status
#-----------------------------
df_long <- df_long %>%
  mutate(
    payout_class = case_when(
      RI <= 0.90 ~ "Payout",
      RI >  0.90 ~ "No Payout",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(payout_class))

#-----------------------------
# 3. State-level percentages
#-----------------------------
state_table <- df_long %>%
  group_by(STATE, payout_class) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(STATE) %>%
  mutate(percent = n / sum(n)) %>%
  ungroup() %>%
  select(-n) %>%
  pivot_wider(
    names_from = payout_class,
    values_from = percent
  )

#-----------------------------
# 4. Total row
#-----------------------------
total_row <- df_long %>%
  group_by(payout_class) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(percent = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(
    names_from = payout_class,
    values_from = percent
  ) %>%
  mutate(STATE = "TOTAL") %>%
  select(STATE, everything())

#-----------------------------
# 5. Combine + round + reorder
#-----------------------------
final_table <- bind_rows(state_table, total_row) %>%
  mutate(
    Payout     = paste0(round(Payout * 100), "%"),
    `No Payout` = paste0(round(`No Payout` * 100), "%")
  ) %>%
  select(STATE, Payout, `No Payout`) %>%
  arrange(STATE == "TOTAL", STATE)   # puts TOTAL last

final_table

export_df_to_latex(
  df = final_table,
  file_path = here("data/outputs/tables/state_RI_payout_table.tex"),
  caption = "RI payout frequencies by state, 2010-2025"
)



```






## DM table

```{r}

#-----------------------------
# 0. Build DM column list
#-----------------------------
years <- start_year:end_year
intervals <- 625:635

DM_cols <- expand.grid(years, intervals) |>
  apply(1, \(x) paste0("DM_median_", x[1], "_", x[2]))

#-----------------------------
# 1. Long format
#-----------------------------
df_long <- grids %>%
  select(STATE, any_of(DM_cols)) %>%
  st_drop_geometry() %>%
  pivot_longer(
    cols = -STATE,
    names_to = "grid_interval",
    values_to = "DM"
  )

#-----------------------------
# 2. Classify D4 status
#-----------------------------
df_long <- df_long %>%
  mutate(
    dm_class = case_when(
      DM >= 4 ~ "D4",
      DM <  4 ~ "Non-D4",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(dm_class))

#-----------------------------
# 3. State-level percentages
#-----------------------------
state_table <- df_long %>%
  group_by(STATE, dm_class) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(STATE) %>%
  mutate(percent = n / sum(n)) %>%
  ungroup() %>%
  select(-n) %>%
  pivot_wider(
    names_from = dm_class,
    values_from = percent
  )

#-----------------------------
# 4. Total row
#-----------------------------
total_row <- df_long %>%
  group_by(dm_class) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(percent = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(
    names_from = dm_class,
    values_from = percent
  ) %>%
  mutate(STATE = "TOTAL") %>%
  select(STATE, everything())

#-----------------------------
# 5. Combine + round + reorder
#-----------------------------
final_table <- bind_rows(state_table, total_row) %>%
  mutate(
    D4      = paste0(round(D4 * 100), "%"),
    `Non-D4` = paste0(round(`Non-D4` * 100), "%")
  ) %>%
  select(STATE, D4, `Non-D4`) %>%
  arrange(STATE == "TOTAL", STATE)   # TOTAL last

final_table


export_df_to_latex(
  df = final_table,
  file_path = here("data/outputs/tables/state_DM_payout_table.tex"),
  caption = "Exceptional drought (D4) frequencies by state, 2010-2025"
)




```





## RI-DM Combos table

```{r}

library(dplyr)
library(tidyr)

# Thresholds
high_drought_threshold <- 4   # DM threshold for D4
ri_threshold <- 0.90          # RI threshold for payout

# Years & intervals
years <- 2010:2025
intervals <- 625:635

# Generate column names
DM_cols <- expand.grid(years, intervals) |>
  apply(1, \(x) paste0("DM_median_", x[1], "_", x[2]))
RI_cols <- expand.grid(years, intervals) |>
  apply(1, \(x) paste0("cpc_index_", x[1], "_", x[2]))

#-----------------------------
# 1. Pivot grids long for DM & RI
#-----------------------------
grids_long <- grids %>%
  st_drop_geometry() %>%
  select(STATE, all_of(DM_cols), all_of(RI_cols)) %>%
  pivot_longer(
    cols = -STATE,
    names_to = "colname",
    values_to = "value"
  ) %>%
  # Identify DM vs RI
  mutate(var_type = case_when(
    grepl("^DM_median_", colname)  ~ "DM",
    grepl("^cpc_index_", colname) ~ "RI"
  )) %>%
  select(-colname)

#-----------------------------
# 2. Pivot wider to get DM & RI per observation
#-----------------------------
grids_wide <- grids_long %>%
  group_by(STATE, var_type) %>%
  mutate(obs = row_number()) %>%
  ungroup() %>%
  pivot_wider(
    id_cols = c(STATE, obs),
    names_from = var_type,
    values_from = value
  )

#-----------------------------
# 3. Create classes & combo factor
#-----------------------------
grids_wide <- grids_wide %>%
  mutate(
    DM_class = ifelse(DM >= high_drought_threshold, "D4", "Other"),
    RI_class = ifelse(RI <= ri_threshold, "Payout", "No Payout"),
    case_high_drought = DM >= high_drought_threshold,
    case_ri_threshold = RI <= ri_threshold,
    combo = factor(
      paste(case_high_drought, case_ri_threshold, sep = "_"),
      levels = c("FALSE_FALSE", "FALSE_TRUE", "TRUE_FALSE", "TRUE_TRUE"),
      labels = c("No drought without payout", 
                 "No drought with payout",
                 "Drought without payout", 
                 "Drought with payout")
    )
  )

#-----------------------------
# 4. Aggregate by state
#-----------------------------
combo_summary <- grids_wide %>%
  group_by(STATE, combo) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(STATE) %>%
  mutate(percent = n / sum(n)) %>%
  ungroup() %>%
  select(-n) %>%
  pivot_wider(
    names_from = combo,
    values_from = percent
  )

#-----------------------------
# 5. Add total row
#-----------------------------
total_row <- grids_wide %>%
  group_by(combo) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(percent = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(
    names_from = combo,
    values_from = percent
  ) %>%
  mutate(STATE = "TOTAL") %>%
  select(STATE, everything())

final_combo_table <- bind_rows(combo_summary, total_row) 

final_combo_table <- final_combo_table %>%
  mutate(
    across(2:3, 
           ~ paste0(format(round(as.numeric(sub("%","",.x)), 3)*100, nsmall=1), "%")),
    # Keep STATE as is
    # Format last two columns with 1 decimal
    across((ncol(final_combo_table)-1):ncol(final_combo_table), 
           ~ paste0(format(round(as.numeric(sub("%","",.x)), 3)*100, nsmall=1), "%"))
  )

final_combo_table

export_df_to_latex(
  df = final_combo_table,
  file_path = here("data/outputs/tables/RI_DM_combo.tex"),
  caption = "Combinations of RI Payouts and DM Exceptional drought classifications, 2010-2025"
)





```


## RI line plot by year

```{r}


library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(sf)

#-----------------------------
# 1. Settings
#-----------------------------
years <- 2010:2025
intervals <- 625:635

thresholds <- c(0.70, 0.80, 0.90)

RI_cols <- expand.grid(years, intervals) |>
  apply(1, \(x) paste0("cpc_index_", x[1], "_", x[2]))

#-----------------------------
# 2. Pivot RI long
#-----------------------------
ri_long <- grids %>%
  st_drop_geometry() %>%
  select(all_of(RI_cols)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "colname",
    values_to = "RI"
  ) %>%
  # Extract year + interval
  mutate(
    year = as.integer(str_extract(colname, "(?<=cpc_index_)\\d{4}")),
    interval = as.integer(str_extract(colname, "(?<=_)\\d{3}$"))
  ) %>%
  filter(!is.na(RI))

#-----------------------------
# 3. Expand for all thresholds
#-----------------------------
ri_thresh_long <- ri_long %>%
  crossing(threshold = thresholds) %>%
  mutate(
    payout = RI <= threshold
  )

#-----------------------------
# 4. Aggregate % payout by year
#-----------------------------
ts_data <- ri_thresh_long %>%
  group_by(year, threshold) %>%
  summarise(
    pct_payout = mean(payout) * 100,
    .groups = "drop"
  )

#-----------------------------
# 5. Plot
#-----------------------------
# 1. Load required libraries

# 2. Example plot object (replace df + vars with yours)
library(dplyr)
library(ggplot2)

# 1. Data for line labels (last year only, selected thresholds)
line_labels <- ts_data %>%
  filter(threshold %in% c(0.9, 0.8, 0.7)) %>%
  group_by(threshold) %>%
  filter(year == max(year)) %>%
  ungroup()

line_labels$label = paste0("RI = ", str_pad(line_labels$threshold, side = "right", pad = 0, width = 4))

p_ri_year <- ggplot(ts_data, aes(x = year, y = pct_payout, group = threshold)) +
  
  # Grey thin lines (all except 0.90)
  geom_line(
    data = ts_data %>% filter(threshold != 0.90),
    color = "grey60",
    linewidth = 0.6
  ) +
  
  # Bold black 0.90 line
  geom_line(
    data = ts_data %>% filter(threshold == 0.90),
    color = "black",
    linewidth = 1.4
  ) +
  
  # 2. Line end labels
  geom_text(
    data = line_labels,
    aes(label = label),
    hjust = 0.5,          # push text slightly right
    vjust = 0.4,
    size = 3.8
  ) +
  
  scale_x_continuous(breaks = seq(2010, 2025, 2)) +
  
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10),
    expand = c(0, 0)
  ) +
  
  labs(
    title = "Percent of Grid-Intervals Triggering PRF Payouts by Year",
    subtitle = "RI Threshold Sensitivity",
    x = "Year",
    y = "Percent of Grid-Intervals with Payout"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )


png(here("data/outputs/figures/p_ri_year.png"),
    width = 600,
    height = 400)
plot(p_ri_year)
dev.off()






```


## DM by year plot

```{r}


library(dplyr)
library(ggplot2)
library(here)

#-----------------------------
# 1. Settings
#-----------------------------
years <- 2010:2025
intervals <- 625:635
dm_thresholds <- 0:4  # thresholds for DM

#-----------------------------
# 2. Compute % of grid-intervals per year and threshold
#-----------------------------
ts_dm_data <- expand.grid(year = years, threshold = dm_thresholds) %>%
  rowwise() %>%
  mutate(
    # Store columns as a list
    DM_median_cols = list(paste0("DM_median_", year, "_", intervals)),
    
    # Compute #cells >= threshold
    n_hit = sum(as.matrix(grids[, DM_median_cols]) >= threshold, na.rm = TRUE),
    
    # Total valid cells
    n_total = sum(!is.na(as.matrix(grids[, DM_median_cols]))),
    
    pct_grids = n_hit / n_total * 100
  ) %>%
  ungroup() %>%
  select(year, threshold, pct_grids)

#-----------------------------
# 3. Line-end labels (last year)
#-----------------------------
line_labels <- ts_dm_data %>%
  filter(year == max(year)) %>%
  mutate(label = paste0("DM ≥ ", threshold))

#-----------------------------
# 4. Plot
#-----------------------------
p_dm_year <- ggplot(ts_dm_data, aes(x = year, y = pct_grids, group = threshold)) +
  
  # Thin grey lines for DM 0–3
  geom_line(data = ts_dm_data %>% filter(threshold != 4),
            color = "grey60", linewidth = 0.6) +
  
  # Thick black line for DM 4
  geom_line(data = ts_dm_data %>% filter(threshold == 4),
            color = "black", linewidth = 1.4) +
  
  # Line-end labels
  geom_text(data = line_labels,
            aes(label = label),
            hjust = 0.4,
            vjust = 0.5,
            size = 3.8) +
  
  scale_x_continuous(breaks = seq(2010, 2025, 2)) +
  
  scale_y_continuous(limits = c(0, 100),
                     breaks = seq(0, 100, 10),
                     expand = c(0, 0)) +
  
  labs(
    title = "Percent of Grid-Intervals by Drought Severity",
    subtitle = "DM Threshold Sensitivity",
    x = "Year",
    y = "Percent of Grid-Intervals Meeting Threshold"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

p_dm_year
#-----------------------------
# 5. Export PNG
#-----------------------------
png(here("data/outputs/figures/p_dm_year.png"),
    width = 600, height = 400)

plot(p_dm_year)

dev.off()

```

## missed payout by year

```{r}

library(dplyr)
library(ggplot2)
library(here)

#-----------------------------
# 1. Settings
#-----------------------------
years <- 2010:2025
intervals <- 625:635
ri_threshold <- 0.9
dm_threshold <- 4

#-----------------------------
# 2. Compute % for DM=4 total and split by RI
#-----------------------------
ts_dm_ri <- lapply(years, function(yr){
  
  DM_cols <- paste0("DM_median_", yr, "_", intervals)
  RI_cols <- paste0("cpc_index_", yr, "_", intervals)
  
  dm_vals <- as.matrix(grids[, DM_cols])
  ri_vals <- as.matrix(grids[, RI_cols])
  
  # Total DM >= 4
  total_hit <- sum(dm_vals >= dm_threshold, na.rm = TRUE)
  total_cells <- sum(!is.na(dm_vals))
  pct_dm4 <- total_hit / total_cells * 100
  
  # DM >=4 AND RI <= 0.9
  payout_hit <- sum(dm_vals >= dm_threshold & ri_vals <= ri_threshold, na.rm = TRUE)
  pct_payout <- payout_hit / total_cells * 100
  
  # DM >=4 AND RI > 0.9
  no_payout_hit <- sum(dm_vals >= dm_threshold & ri_vals > ri_threshold, na.rm = TRUE)
  pct_no_payout <- no_payout_hit / total_cells * 100
  
  data.frame(
    year = yr,
    pct_dm4 = pct_dm4,
    pct_payout = pct_payout,
    pct_no_payout = pct_no_payout
  )
  
}) %>% bind_rows()

#-----------------------------
# 3. Plot
#-----------------------------
p_dm4_zoom <- ggplot(ts_dm_ri, aes(x = year)) +
  
  # DM >=4 total (optional, black line)
  geom_line(aes(y = pct_dm4), color = "black", linewidth = 1.2) +
  
  # DM=4 & RI <=0.9
  geom_line(aes(y = pct_payout), color = "blue", linewidth = 1.2) +
  
  # DM=4 & RI >0.9
  geom_line(aes(y = pct_no_payout), color = "red", linewidth = 1.2) +
  
  scale_x_continuous(breaks = seq(2010, 2025, 2)) +
  scale_y_continuous(limits = c(0, 25), breaks = seq(0, 25, 5), expand = c(0, 0)) +
  
  labs(
    title = "DM = 4 Grid-Intervals Split by RI",
    subtitle = "Blue = payout (RI ≤ 0.9), Red = no payout (RI > 0.9)",
    x = "Year",
    y = "Percent of Grid-Intervals"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

p_dm4_zoom

#-----------------------------
# 4. Export PNG
#-----------------------------
png(here("data/outputs/figures/p_dm4_zoom.png"),
    width = 600, height = 400)

plot(p_dm4_zoom)

dev.off()











```

## heat maps





## lag analysis (during drought by month, what is the % likelihood of a payout)
## also, perhaps, categorize droughts by length for a grid cell, are short/long droughts more likely to receive a payout? 






## sample grid maps
```{r}

# 1. Recode values
grids <- grids %>%
  mutate(DM_class = ifelse(DM_median_2022_631 >= high_drought_threshold, "D4", "Other")) %>% 
  mutate(RI_class = ifelse(cpc_index_2022_631 <= ri_threshold, "Payout", "No Payout")) %>% 
  mutate(case_high_drought = ifelse(DM_median_2022_631 >= high_drought_threshold, TRUE, FALSE)) %>% 
  mutate(case_ri_threshold = ifelse(cpc_index_2022_631 <= ri_threshold, TRUE, FALSE))



# 2. Plot with two classes
ggplot(grids) +
  geom_sf(aes(fill = DM_class), color = NA) +
  scale_fill_manual(
    values = c(
      "D4" = "darkred",
      "Other" = "grey"
    ),
    name = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "Max USDM July–August 2022")

ggplot(grids) +
  geom_sf(aes(fill = RI_class), color = NA) +
  scale_fill_manual(
    values = c(
      "Payout" = "darkred",
      "No Payout" = "grey"
    ),
    name = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "PRF Payouts July–August 2022")


```

## crosstab map

```{r}

library(sf)
library(dplyr)
library(ggplot2)
library(tigris)

options(tigris_use_cache = TRUE)

grids_collapsed <- grids

grids_collapsed$combo <- with(grids_collapsed,
  factor(
    paste(case_high_drought, case_ri_threshold, sep = "_"),
    levels = c("FALSE_FALSE", "FALSE_TRUE", "TRUE_FALSE", "TRUE_TRUE"),
    labels = c("No drought without payout", 
               "No drought with payout",
               "Drought without payout", "Drought with payout")
  )
)
#-----------------------------
# 2. Get AOI State polygons
#-----------------------------
states_aoi <- states(cb = TRUE, resolution = "20m") %>%
  filter(STUSPS %in% aoi_states) %>%
  st_transform(st_crs(grids_collapsed))  # match CRS of grids_collapsed

#-----------------------------
# 3. Clip grids_collapsed to AOI
#-----------------------------
grids_aoi <- st_intersection(grids_collapsed, st_union(states_aoi)) %>%
  filter(!is.na(combo))  # remove NA values

#-----------------------------
# 4. Recreate combo factor (if needed)
#-----------------------------
grids_aoi$combo <- factor(
  grids_aoi$combo,
  levels = c("No drought without payout", 
             "No drought with payout",
             "Drought without payout", 
             "Drought with payout")
)

combo_colors <- c(
  "No drought without payout" = "lightgrey",
  "No drought with payout" = "skyblue",
  "Drought without payout" = "red",
  "Drought with payout" = "orange"
)

#-----------------------------
# 5. Plot
#-----------------------------
p_crosstab <- ggplot() +
  geom_sf(data = grids_aoi, aes(fill = combo), color = NA) +
  geom_sf(data = states_aoi, fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_manual(
    values = combo_colors,
    name = "Drought / Payout Classification"
  ) +
  labs(title = "Crosstab Map : DM = 4 vs RI ≤ 0.9 July–August 2022") +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )


png(here("data/outputs/figures/p_crosstab.png"),
    width = 800,
    height = 600)
plot(p_crosstab)
dev.off()


```




## crosstab table - all years 

```{r}

years <- start_year:end_year
intervals <- 625:635

for (year in years) {
  for (interval in intervals) {
    # Construct column names dynamically
    dm_col <- paste0("DM_median_", year, "_", interval)
    ri_col <- paste0("cpc_index_", year, "_", interval)
    
    case_dm_col <- paste0("case_high_drought_", year, "_", interval)
    case_ri_col <- paste0("case_ri_threshold_", year, "_", interval)
    combo_col   <- paste0("combo_", year, "_", interval)
    
    # Only create if both source columns exist
    if (all(c(dm_col, ri_col) %in% names(grids_collapsed))) {
      
      # Boolean columns
      grids_collapsed[[case_dm_col]] <- grids_collapsed[[dm_col]] == 4
      grids_collapsed[[case_ri_col]] <- grids_collapsed[[ri_col]] <= 0.9
      
      # Combo factor column
      grids_collapsed[[combo_col]] <- with(grids_collapsed,
        factor(
          paste(grids_collapsed[[case_dm_col]], grids_collapsed[[case_ri_col]], sep = "_"),
          levels = c("FALSE_FALSE", "FALSE_TRUE", "TRUE_FALSE", "TRUE_TRUE"),
          labels = c("Neither", "RI only", "DM only", "Both")
        )
      )
    }
  }
}

# Initialize count vectors
dm_only     <- integer(nrow(grids_collapsed))
ri_only     <- integer(nrow(grids_collapsed))
both        <- integer(nrow(grids_collapsed))
neither     <- integer(nrow(grids_collapsed))
total_valid <- integer(nrow(grids_collapsed))  # count of total comparisons per row

# Loop through each year and interval
for (year in years) {
  for (interval in intervals) {
    dm_col <- paste0("case_high_drought_", year, "_", interval)
    ri_col <- paste0("case_ri_threshold_", year, "_", interval)
    
    # Only proceed if both columns exist
    if (all(c(dm_col, ri_col) %in% names(grids_collapsed))) {
      dm_vals <- grids_collapsed[[dm_col]]
      ri_vals <- grids_collapsed[[ri_col]]
      
      valid <- !is.na(dm_vals) & !is.na(ri_vals)  # handle missing data safely
      total_valid <- total_valid + valid
      
      dm_only     <- dm_only     + (dm_vals & !ri_vals)
      ri_only     <- ri_only     + (!dm_vals & ri_vals)
      both        <- both        + (dm_vals & ri_vals)
      neither     <- neither     + (!dm_vals & !ri_vals)
    }
  }
}

# Add raw count columns
grids_collapsed$dm_only_count  <- dm_only
grids_collapsed$ri_only_count  <- ri_only
grids_collapsed$both_count     <- both
grids_collapsed$neither_count  <- neither
grids_collapsed$total_valid    <- total_valid

grids_collapsed$dm_true_count <- grids_collapsed$dm_only_count + grids_collapsed$both_count

grids_collapsed$missed_payout_pct <- 100*(grids_collapsed$dm_only_count/grids_collapsed$dm_true_count)

# Calculate percentages for each class
grids_collapsed$dm_only_pct   <- 100 * grids_collapsed$dm_only_count / grids_collapsed$total_valid
grids_collapsed$ri_only_pct   <- 100 * grids_collapsed$ri_only_count / grids_collapsed$total_valid
grids_collapsed$both_pct      <- 100 * grids_collapsed$both_count    / grids_collapsed$total_valid
grids_collapsed$neither_pct   <- 100 * grids_collapsed$neither_count / grids_collapsed$total_valid

summary_table <- grids_collapsed %>%
  st_drop_geometry() %>%  # remove spatial data for printing
  select(dm_only_pct, ri_only_pct, both_pct, neither_pct) %>%
  summarise(across(everything(), ~ round(mean(.x, na.rm = TRUE), 1)))

print(summary_table)

```


## crosstab table
```{r}

# Sum total counts across all rows
dm_only_total   <- sum(grids_collapsed$dm_only_count, na.rm = TRUE)
ri_only_total   <- sum(grids_collapsed$ri_only_count, na.rm = TRUE)
both_total      <- sum(grids_collapsed$both_count, na.rm = TRUE)
neither_total   <- sum(grids_collapsed$neither_count, na.rm = TRUE)

# Total number of valid combinations evaluated
total_combos <- dm_only_total + ri_only_total + both_total + neither_total

# Percentages
dm_only_pct   <- 100 * dm_only_total / total_combos
ri_only_pct   <- 100 * ri_only_total / total_combos
both_pct      <- 100 * both_total / total_combos
neither_pct   <- 100 * neither_total / total_combos

# Print summary
summary_table <- data.frame(
  category = c("DM only", "RI only", "Both", "Neither"),
  count    = c(dm_only_total, ri_only_total, both_total, neither_total),
  percent  = round(c(dm_only_pct, ri_only_pct, both_pct, neither_pct), 1)
)

print(summary_table)

```



## absent payout %
```{r}

library(RColorBrewer)


# Plot: Continuous scale with viridis "C" reversed
ggplot(data = grids_collapsed) +
  geom_sf(aes(fill = missed_payout_pct), color = NA) +
  scale_fill_viridis_c(
    name = "% Absent Payout",
    option = "C",
    direction = -1,
    limits = c(0, 100),
    na.value = "grey90"  # NA values in light grey
  ) +
  labs(title = "Absent Payout % by Grid Cell") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

```

## absent payout count

```{r}

grids_collapsed$dm_only_count[is.na(grids_collapsed$missed_payout_pct)] <- NA

# Plot: Continuous scale with viridis "C" reversed
ggplot(data = grids_collapsed) +
  geom_sf(aes(fill = dm_only_count), color = NA) +
  scale_fill_viridis_c(
    name = "# Absent Payouts",
    option = "C",
    direction = -1,
    na.value = "grey90"  # NA values in light grey
  ) +
  labs(title = paste0("Absent Payout Count ", start_year, "-", end_year)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )


```


```{r}

save(grids_collapsed, file = here("data/grids_collapsed.rda"))

```





```{r}


# 1. Gather DM_median columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )


# 4. Summarize: for each year, calculate % of intervals with DM ≥ 3
dm_summary <- dm_long %>%
  group_by(year) %>%
  summarise(
    mean_ri = mean(RI, na.rm = TRUE)
  )

# 5. Plot
ggplot(dm_summary, aes(x = year, y = mean_ri)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = "Mean RI",
    title = "Mean RI by Year"
  ) +
  ylim(0,2) +
  theme_minimal(base_size = 14)






```



### % obs for different RI thresholds

```{r}

# 1. Gather DM_median columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )


# 4. Summarize: for each year, calculate % of intervals with RI < thresholds
for(threshold in seq(0.7, 0.9, 0.05)){
  print(threshold)
  pct <- 100*(sum(dm_long$RI < threshold)/nrow(dm_long))
  print(paste0(round(pct, 1), "%"))
}



```

### summary stats, DM category
```{r}

# 1. Gather DM_median columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_median_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "DM_value"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "DM_median_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )


# 4. Summarize: for each year, calculate % of intervals with RI < thresholds
for(threshold in seq(0, 4, 1)){
  print(threshold)
  pct <- 100*(sum(dm_long$DM_value >= threshold)/nrow(dm_long))
  print(paste0(round(pct, 1), "%"))
}


```


## by year, >= 3, 4

```{r}



# 1. Gather DM_median columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_median_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "dm_value"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "DM_median_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )

# 3. For each row (site × interval), define if DM ≥ 3
dm_long <- dm_long %>%
  mutate(high_dm = dm_value >= 3)

# 4. Summarize: for each year, calculate % of intervals with DM ≥ 3
dm_summary <- dm_long %>%
  group_by(year) %>%
  summarise(
    pct_high_dm = mean(high_dm, na.rm = TRUE) * 100
  )

# 5. Plot
ggplot(dm_summary, aes(x = year, y = pct_high_dm)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = "% intervals where max DM ≥ 3",
    title = "High Drought Frequency by Year"
  ) +
  ylim(0,100) +
  theme_minimal(base_size = 14)


# 3. For each row (site × interval), define if DM ≥ 4
dm_long <- dm_long %>%
  mutate(high_dm = dm_value >= 4)

# 4. Summarize: for each year, calculate % of intervals with DM ≥ 4
dm_summary <- dm_long %>%
  group_by(year) %>%
  summarise(
    pct_high_dm = mean(high_dm, na.rm = TRUE) * 100
  )

# 5. Plot
ggplot(dm_summary, aes(x = year, y = pct_high_dm)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = "% intervals where max DM ≥ 4",
    title = "High Drought Frequency by Year"
  ) +
  ylim(0,100) +
  theme_minimal(base_size = 14)



```


## intervals >=3, >= 4

```{r}


# 1. Make sure you have the correct long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_median_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_median_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(dm_value)) %>%
  mutate(high_dm = dm_value >= 3)

# 2. Summarize: % of rows × years where DM_median ≥ 3 by interval
interval_summary <- dm_long %>%
  group_by(interval) %>%
  summarise(
    pct_high_dm = mean(high_dm) * 100,
    .groups = "drop"
  )

# 3. Create readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

# 4. Join labels
interval_summary <- interval_summary %>%
  left_join(interval_labels, by = "interval")

# 5. Plot
ggplot(interval_summary, aes(x = factor(label, levels = label), y = pct_high_dm)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Interval",
    y = "% obs with DM ≥ 3",
    title = "High Drought Frequency by Calendar Interval"
  ) +
  theme_minimal(base_size = 14) +  
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20),         # major breaks every 20
    minor_breaks = seq(0, 100, 10)    # minor breaks every 10
  ) 



# 1. Make sure you have the correct long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_median_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_median_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(dm_value)) %>%
  mutate(high_dm = dm_value >= 4)

# 2. Summarize: % of rows × years where DM_median ≥ 4 by interval
interval_summary <- dm_long %>%
  group_by(interval) %>%
  summarise(
    pct_high_dm = mean(high_dm) * 100,
    .groups = "drop"
  )

# 3. Create readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

# 4. Join labels
interval_summary <- interval_summary %>%
  left_join(interval_labels, by = "interval")

# 5. Plot
ggplot(interval_summary, aes(x = factor(label, levels = label), y = pct_high_dm)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Interval",
    y = "% obs with DM ≥ 4",
    title = "High Drought Frequency by Calendar Interval"
  ) +
  theme_minimal(base_size = 14) +  
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20),         # major breaks every 20
    minor_breaks = seq(0, 100, 10)    # minor breaks every 10
  ) 

```


## heatmap DM = 4

```{r}



# Step 1: Long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("DM_median_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_median_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(dm_value)) %>%
  mutate(high_dm = dm_value >= 4)

# Step 2: Summarize by year × interval
heatmap_data <- dm_long %>%
  group_by(year, interval) %>%
  summarise(
    pct_high_dm = mean(high_dm) * 100,
    .groups = "drop"
  )

# Step 3: Add readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

heatmap_data <- heatmap_data %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))  # Preserve order


#-----------------------------
# Step 4: Plot heatmap with 0% in white
#-----------------------------
#-----------------------------
# Determine max value for color scaling
#-----------------------------
max_val <- max(heatmap_data$pct_high_dm, na.rm = TRUE)

# Small epsilon above 0 to allow white for 0
eps <- 0.0001

#-----------------------------
# Plot heatmap
#-----------------------------
p_heatmap_dm <- ggplot(heatmap_data, aes(x = label, y = factor(year), fill = pct_high_dm)) +
  geom_tile(color = "lightgrey", linewidth = 0.3) +  # light grey lines
  
  # Color ramp: white for 0, viridis from just above 0 to max_val
  scale_fill_gradientn(
    colors = c("white", viridis(100, option = "C", direction = -1)),
    values = rescale(c(0, eps, seq(eps, max_val, length.out = 100))),
    limits = c(0, max_val),
    name = "% DM = 4",
    na.value = "grey90"
  ) +
  
  labs(
    x = "Interval",
    y = "Year",
    title = "Heatmap of High Drought Frequency (DM Max = 4)",
    fill = "% High DM"
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

png(here("data/outputs/figures/heatmap_DM.png"),
    width = 800,
    height = 600)
plot(p_heatmap_dm)
dev.off()

```

## heatmap RI
```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)
library(here)

#-----------------------------
# 1. Long format RI
#-----------------------------
ri_long <- grids %>%
  st_drop_geometry() %>%
  select(starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  ) %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(RI)) %>%
  mutate(payout = RI <= 0.90)  # threshold for payout

#-----------------------------
# 2. Summarize by year × interval
#-----------------------------
heatmap_ri <- ri_long %>%
  group_by(year, interval) %>%
  summarise(
    pct_payout = mean(payout) * 100,
    .groups = "drop"
  )

#-----------------------------
# 3. Add readable interval labels
#-----------------------------
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

heatmap_ri <- heatmap_ri %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))  # preserve order

#-----------------------------
# 4. Determine min/max for color scaling
#-----------------------------
min_val <- min(heatmap_ri$pct_payout, na.rm = TRUE)
max_val <- max(heatmap_ri$pct_payout, na.rm = TRUE)

#-----------------------------
# 5. Plot RI heatmap
#-----------------------------
p_heatmap_ri <- ggplot(heatmap_ri, aes(x = label, y = factor(year), fill = pct_payout)) +
  geom_tile(color = "lightgrey", linewidth = 0.3) +
  
  scale_fill_viridis_c(
    name = "% Payout (RI ≤ 0.90)",
    option = "C",
    direction = -1,
    limits = c(min_val, max_val)
  ) +
  
  labs(
    x = "Interval",
    y = "Year",
    title = "Heatmap of Payout Frequency (RI ≤ 0.90)",
    fill = "% Payout"
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

p_heatmap_ri
#-----------------------------
# 6. Export PNG
#-----------------------------
png(here("data/outputs/figures/heatmap_RI.png"),
    width = 800,
    height = 600)

plot(p_heatmap_ri)
dev.off()



```

## drought length effects

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

#-----------------------------
# 1. Identify DM=4 per grid × interval
#-----------------------------
# Get only DM_median columns
dm_cols <- grep("^DM_median_", names(grids), value = TRUE)

dm_long <- grids %>%
  st_drop_geometry() %>%
  select(GRIDCODE, all_of(dm_cols)) %>%  # assume grids has a GRIDCODE column
  pivot_longer(
    cols = -GRIDCODE,
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    dm_flag = as.integer(dm_value >= 4),
    year = as.integer(substr(var, 8, 11)),
    interval = as.integer(substr(var, 13, 15))
  ) %>%
  arrange(GRIDCODE, year, interval)

#-----------------------------
# 2. Compute consecutive runs (RLE)
#-----------------------------
# For each grid, compute runs of DM=4
drought_runs <- dm_long %>%
  group_by(GRIDCODE) %>%
  summarise(
    runs = {
      rle_vals <- rle(dm_flag)
      # Only keep runs where dm_flag == 1
      rle_lengths <- rle_vals$lengths[rle_vals$values == 1]
      list(rle_lengths)
    },
    .groups = "drop"
  ) %>%
  unnest(cols = c(runs))  # one row per drought event

#-----------------------------
# 3. Histogram of drought lengths
#-----------------------------
hist_data <- drought_runs %>%
  count(runs) %>%
  mutate(pct = n / sum(n) * 100)

ggplot(hist_data, aes(x = runs, y = pct)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  scale_x_continuous(breaks = seq(1, max(hist_data$runs), 1)) +
  labs(
    x = "Length of Consecutive DM = 4 Intervals",
    y = "Percent of Drought Events",
    title = "Distribution of Exceptional Drought Lengths (DM = 4)"
  ) +
  theme_minimal(base_size = 13)





```

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

#-----------------------------
# 1. Compute cumulative distribution
#-----------------------------
cum_dist <- drought_runs %>%
  count(runs) %>%
  arrange(runs) %>%
  mutate(
    pct = n / sum(n) * 100,
    cum_pct = cumsum(pct)
  )

#-----------------------------
# 2. Print cumulative distribution
#-----------------------------
print(cum_dist)

# Example output columns:
# runs     n     pct    cum_pct

#-----------------------------
# 3. Plot cumulative distribution
#-----------------------------
ggplot(cum_dist, aes(x = runs, y = cum_pct)) +
  geom_step(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_continuous(breaks = seq(1, max(cum_dist$runs), 1)) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = "Length of Consecutive DM = 4 Intervals",
    y = "Cumulative Percent of Drought Events",
    title = "Cumulative Distribution of Exceptional Drought Lengths (DM = 4)"
  ) +
  theme_minimal(base_size = 13)

```

## drought length by state

```{r}

library(dplyr)
library(tidyr)

#-----------------------------
# 1. Add STATE information to drought_runs
#-----------------------------
drought_runs_state <- drought_runs %>%
  left_join(
    grids %>% st_drop_geometry() %>% select(GRIDCODE, STATE),
    by = "GRIDCODE"
  )

#-----------------------------
# 2. Categorize drought lengths
#-----------------------------
drought_runs_state <- drought_runs_state %>%
  mutate(
    length_category = case_when(
      runs >= 1 & runs <= 5 ~ "Short (1-5)",
      runs >= 6 & runs <= 10 ~ "Medium (6-10)",
      runs >= 11 ~ "Long (11+)"
    )
  )

#-----------------------------
# 3a. Count table per state
#-----------------------------
count_table <- drought_runs_state %>%
  group_by(STATE, length_category) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = length_category,
    values_from = count,
    values_fill = 0
  ) %>%
  # Ensure column order
  select(STATE, `Short (1-5)`, `Medium (6-10)`, `Long (11+)`)

# Add TOTAL row
total_counts <- count_table %>%
  summarise(across(-STATE, sum)) %>%
  mutate(STATE = "TOTAL") %>%
  select(STATE, everything())

count_table <- bind_rows(count_table, total_counts)

#-----------------------------
# 3b. Percent table per state
#-----------------------------
percent_table <- count_table %>%
  mutate(
    total = `Short (1-5)` + `Medium (6-10)` + `Long (11+)`,
    across(c(`Short (1-5)`, `Medium (6-10)`, `Long (11+)`),
           ~ paste0(round(.x / total * 100), "%"))
  ) %>%
  select(-total)

#-----------------------------
# 4. View tables
#-----------------------------
count_table
percent_table


```

```{r}

library(dplyr)
library(tidyr)

#-----------------------------
# 1. Add STATE info
#-----------------------------
drought_runs_state <- drought_runs %>%
  left_join(
    grids %>% st_drop_geometry() %>% select(GRIDCODE, STATE),
    by = "GRIDCODE"
  )

#-----------------------------
# 2. Categorize drought lengths
#-----------------------------
drought_runs_state <- drought_runs_state %>%
  mutate(
    length_category = case_when(
      runs >= 1 & runs <= 5 ~ "Short (1-5)",
      runs >= 6 & runs <= 10 ~ "Medium (6-10)",
      runs >= 11 ~ "Long (11+)"
    )
  )

#-----------------------------
# 3a. Count table weighted by grid-intervals
#-----------------------------
count_table_intervals <- drought_runs_state %>%
  group_by(STATE, length_category) %>%
  summarise(grid_intervals = sum(runs), .groups = "drop") %>%   # sum of intervals per category
  pivot_wider(
    names_from = length_category,
    values_from = grid_intervals,
    values_fill = 0
  ) %>%
  select(STATE, `Short (1-5)`, `Medium (6-10)`, `Long (11+)`)  # ensure order

# Add TOTAL row
total_counts <- count_table_intervals %>%
  summarise(across(-STATE, sum)) %>%
  mutate(STATE = "TOTAL") %>%
  select(STATE, everything())

count_table_intervals <- bind_rows(count_table_intervals, total_counts)

#-----------------------------
# 3b. Percent table weighted by grid-intervals
#-----------------------------
percent_table_intervals <- count_table_intervals %>%
  mutate(
    total = `Short (1-5)` + `Medium (6-10)` + `Long (11+)`,
    across(c(`Short (1-5)`, `Medium (6-10)`, `Long (11+)`),
           ~ paste0(round(.x / total * 100), "%"))
  ) %>%
  select(-total)

#-----------------------------
# 4. View tables
#-----------------------------
count_table_intervals
percent_table_intervals






```

## payouts by drought length

```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(sf)

conflicts_prefer(dplyr::lag)

#====================================================
# 1. Build long dataset (DM + RI aligned)
#====================================================
long_data <- grids %>%
  st_drop_geometry() %>%
  select(
    GRIDCODE, STATE,
    starts_with("DM_median_"),
    starts_with("cpc_index_")
  ) %>%
  pivot_longer(
    cols = -c(GRIDCODE, STATE),
    names_to = "var",
    values_to = "value"
  ) %>%
  mutate(
    type = case_when(
      str_detect(var, "^DM_median_") ~ "DM",
      str_detect(var, "^cpc_index_") ~ "RI"
    ),
    var = str_remove(var, "^DM_median_|^cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  select(-var) %>%
  pivot_wider(names_from = type, values_from = value) %>%
  arrange(GRIDCODE, year, interval)

#====================================================
# 2. Identify drought runs (DM = 4)
#====================================================
long_data <- long_data %>%
  mutate(
    drought = DM == 4,
    payout  = RI <= 0.90
  ) %>%
  group_by(GRIDCODE) %>%
  mutate(
    drought_group = cumsum(drought != lag(drought, default = FALSE))
  ) %>%
  ungroup()

#====================================================
# 3. Compute drought run length
#====================================================
drought_lengths <- long_data %>%
  filter(drought) %>%
  group_by(GRIDCODE, drought_group) %>%
  summarise(
    run_length = n(),
    STATE = first(STATE),
    .groups = "drop"
  )

long_data <- long_data %>%
  left_join(
    drought_lengths,
    by = c("GRIDCODE", "drought_group", "STATE")
  )

#====================================================
# 4. Categorize drought length
#====================================================
long_data <- long_data %>%
  mutate(
    length_category = case_when(
      run_length >= 1 & run_length <= 5  ~ "Short (1-5)",
      run_length >= 6 & run_length <= 10 ~ "Medium (6-10)",
      run_length >= 11                   ~ "Long (11+)"
    )
  )

#====================================================
# 5. Calculate payout % (interval-weighted)
#====================================================
payout_summary <- long_data %>%
  filter(drought) %>%
  group_by(STATE, length_category) %>%
  summarise(
    total_intervals  = n(),
    payout_intervals = sum(payout, na.rm = TRUE),
    payout_pct       = payout_intervals / total_intervals * 100,
    .groups = "drop"
  )

#====================================================
# 6. Format final table
#====================================================
payout_table <- payout_summary %>%
  select(STATE, length_category, payout_pct) %>%
  pivot_wider(
    names_from = length_category,
    values_from = payout_pct
  ) %>%
  select(STATE, `Short (1-5)`, `Medium (6-10)`, `Long (11+)`)

# TOTAL row (interval-weighted across all states)
total_row <- long_data %>%
  filter(drought) %>%
  group_by(length_category) %>%
  summarise(
    payout_pct = sum(payout, na.rm = TRUE) / n() * 100,
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = length_category,
    values_from = payout_pct
  ) %>%
  mutate(STATE = "TOTAL") %>%
  select(STATE, `Short (1-5)`, `Medium (6-10)`, `Long (11+)`)

payout_table <- bind_rows(payout_table, total_row)

# Round and add %
payout_table <- payout_table %>%
  mutate(across(-STATE, ~ paste0(round(.x), "%")))

payout_table


```



## lag effects 

```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(sf)

#====================================================
# 1. Build long dataset (DM + RI aligned)
#====================================================
long_data <- grids %>%
  st_drop_geometry() %>%
  select(
    GRIDCODE, STATE,
    starts_with("DM_median_"),
    starts_with("cpc_index_")
  ) %>%
  pivot_longer(
    cols = -c(GRIDCODE, STATE),
    names_to = "var",
    values_to = "value"
  ) %>%
  mutate(
    type = case_when(
      str_detect(var, "^DM_median_") ~ "DM",
      str_detect(var, "^cpc_index_") ~ "RI"
    ),
    var = str_remove(var, "^DM_median_|^cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  select(-var) %>%
  pivot_wider(names_from = type, values_from = value) %>%
  arrange(GRIDCODE, year, interval)

#====================================================
# 2. Identify drought runs
#====================================================
long_data <- long_data %>%
  mutate(
    drought = DM == 4,
    payout  = RI <= 0.90
  ) %>%
  group_by(GRIDCODE) %>%
  mutate(
    drought_group = cumsum(drought != dplyr::lag(drought, default = FALSE))
  ) %>%
  ungroup()

#====================================================
# 3. Compute position within each drought
#====================================================
drought_positions <- long_data %>%
  filter(drought) %>%
  group_by(GRIDCODE, drought_group) %>%
  mutate(
    position_in_drought = row_number()
  ) %>%
  ungroup()

#====================================================
# 4. Calculate payout likelihood by position
#====================================================
position_summary <- drought_positions %>%
  group_by(position_in_drought) %>%
  summarise(
    total_intervals  = n(),
    payout_intervals = sum(payout, na.rm = TRUE),
    payout_prob      = payout_intervals / total_intervals,
    .groups = "drop"
  )

#====================================================
# 5. Plot
#====================================================
ggplot(position_summary,
       aes(x = position_in_drought,
           y = payout_prob)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Position Within Consecutive Drought",
    y = "Likelihood of Payout",
    title = "Payout Probability by Drought Position",
    subtitle = "Conditional on DM = 4"
  ) +
  theme_minimal(base_size = 13) 


ggplot(position_summary,
       aes(x = position_in_drought,
           y = total_intervals)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    x = "Position Within Consecutive Drought",
    y = "Number of Drought Grid-Intervals",
    title = "Number of Observations by Drought Position",
    subtitle = "Counts of DM = 4 intervals at each position"
  ) +
  theme_minimal(base_size = 13)




```


```{r}

library(dplyr)
library(ggplot2)

#====================================================
# 1. Compute reverse position within each drought
#====================================================
reverse_positions <- long_data %>%
  filter(DM == 4) %>%
  group_by(GRIDCODE, drought_group) %>%
  mutate(
    run_length = n(),
    position_from_end = run_length - row_number() + 1
  ) %>%
  ungroup()

#====================================================
# 2. Compute payout probability by reverse position
#====================================================
reverse_summary <- reverse_positions %>%
  group_by(position_from_end) %>%
  summarise(
    total_intervals  = n(),
    payout_intervals = sum(RI <= 0.90, na.rm = TRUE),
    payout_prob      = payout_intervals / total_intervals,
    .groups = "drop"
  )

#====================================================
# 3. Plot payout probability (reverse axis)
#====================================================
ggplot(reverse_summary,
       aes(x = position_from_end,
           y = payout_prob)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_x_reverse() +   # <-- makes drought end on the right
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Position from End of Drought",
    y = "Likelihood of Payout",
    title = "Payout Probability by Position from Drought End",
    subtitle = "Conditional on DM = 4"
  ) +
  theme_minimal(base_size = 13)

#====================================================
# 4. Plot observation counts (reverse axis)
#====================================================
ggplot(reverse_summary,
       aes(x = position_from_end,
           y = total_intervals)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_x_reverse() +
  labs(
    x = "Position from End of Drought",
    y = "Number of Drought Grid-Intervals",
    title = "Number of Observations by Position from Drought End"
  ) +
  theme_minimal(base_size = 13)




```








## similar plots for RI
### year

```{r}
 
# 1. Gather DM_median columns into long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  )

# 2. Extract year and interval from the column names
dm_long <- dm_long %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )

# 3. For each row (site × interval), define if DM ≥ 3
dm_long <- dm_long %>%
  mutate(low_ri = RI < ri_threshold)

# 4. Summarize: for each year, calculate % of intervals with DM ≥ 3
dm_summary <- dm_long %>%
  group_by(year) %>%
  summarise(
    pct_low_ri = mean(low_ri, na.rm = TRUE) * 100
  )

# 5. Plot
ggplot(dm_summary, aes(x = year, y = pct_low_ri)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = paste0("% obs where RI ≤ ", ri_threshold), 
    title = "Low RI Frequency by Year"
  ) +
  ylim(0,100) +
  theme_minimal(base_size = 14)

# # 3. For each row (site × interval), define if DM ≥ 3
# dm_long <- dm_long %>%
#   mutate(low_ri = RI < 0.9)
# 
# # 4. Summarize: for each year, calculate % of intervals with DM ≥ 3
# dm_summary <- dm_long %>%
#   group_by(year) %>%
#   summarise(
#     pct_low_ri = mean(low_ri, na.rm = TRUE) * 100
#   )
# 
# # 5. Plot
# ggplot(dm_summary, aes(x = year, y = pct_low_ri)) +
#   geom_line(color = "firebrick", size = 1.2) +
#   geom_point(color = "black") +
#   scale_y_continuous(labels = scales::percent_format(scale = 1)) +
#   labs(
#     x = "Year",
#     y = "% obs where RI ≤ 0.9", 
#     title = "High Drought Frequency by Year"
#   ) +
#   ylim(0,100) +
#   theme_minimal(base_size = 14)
# 


```



## interval

```{r}


# 1. Make sure you have the correct long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  ) %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(RI)) %>%
  mutate(low_ri = RI < ri_threshold)

# 2. Summarize: % of rows × years where DM_median ≥ 3 by interval
interval_summary <- dm_long %>%
  group_by(interval) %>%
  summarise(
    pct_low_ri = mean(low_ri) * 100,
    .groups = "drop"
  )

# 3. Create readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

# 4. Join labels
interval_summary <- interval_summary %>%
  left_join(interval_labels, by = "interval")

# 5. Plot
ggplot(interval_summary, aes(x = factor(label, levels = label), y = pct_low_ri)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Interval",
    y = paste0("% obs where RI ≤ ", ri_threshold),
    title = "Low RI by Calendar Interval"
  ) +
  theme_minimal(base_size = 14) +  
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20),         # major breaks every 20
    minor_breaks = seq(0, 100, 10)    # minor breaks every 10
  ) 




```


## heatmap

```{r}


# Step 1: Long format
dm_long <- grids %>% st_drop_geometry() %>% 
  select(starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "var",
    values_to = "RI"
  ) %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  filter(!is.na(RI)) %>%
  mutate(low_ri = RI < ri_threshold)

# Step 2: Summarize by year × interval
heatmap_data <- dm_long %>%
  group_by(year, interval) %>%
  summarise(
    pct_low_ri = mean(low_ri) * 100,
    .groups = "drop"
  )

# Step 3: Add readable interval labels
interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]],
    "-",
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

heatmap_data <- heatmap_data %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))  # Preserve order

# Step 4: Plot heatmap
ggplot(heatmap_data, aes(x = label, y = factor(year), fill = pct_low_ri)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = paste0("% RI ≤ ", ri_threshold), option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of Low RI Frequency (RI ≤ ", ri_threshold, ")"),
    fill = "% Low RI"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

```

## payouts



## chatgpt, DM ≥3
```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)

# 1. Pivot DM_median columns to long

dm_long <- grids %>%
  st_drop_geometry() %>%
  select(GRIDCODE, starts_with("DM_median_")) %>%
  pivot_longer(
    cols = starts_with("DM_median_"),
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_median_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6)),
    row_id = row_number()  # optional: track row identity
  )

# 2. Pivot cpc_index columns to long
cpc_long <- grids %>% st_drop_geometry() %>% 
  select(GRIDCODE, starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = starts_with("cpc_index_"),
    names_to = "var",
    values_to = "cpc_value"
  ) %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  ) %>%
  mutate(row_id = row_number())

# 3. Join DM_median and CPC on row + year + interval
joined <- left_join(dm_long, cpc_long, by = c("GRIDCODE", "row_id", "year", "interval"))

# 4. Filter where DM_median >= 3
filtered <- joined %>%
  filter(!is.na(dm_value), dm_value >= 3, !is.na(cpc_value))

# 5. Calculate percentage of cpc_value < ri_threshold among these filtered
summary_df <- filtered %>%
  group_by(year) %>%
  summarise(
    pct_cpc_below_0_7 = mean(cpc_value < ri_threshold, na.rm = TRUE) * 100,
    count = n()
  )

# 6. Plot
ggplot(summary_df, aes(x = year, y = pct_cpc_below_0_7)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(limits = c(0, 100), labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = paste0("% CPC Index < ", ri_threshold, " (where DM_median ≥ 3)"),
    title = "Frequency of Low CPC Index During High Drought Periods"
  ) +
  theme_minimal(base_size = 14)


# 4. Filter where DM_median >= 4
filtered <- joined %>%
  filter(!is.na(dm_value), dm_value >= 4, !is.na(cpc_value))

# 5. Calculate percentage of cpc_value < ri_threshold among these filtered
summary_df <- filtered %>%
  group_by(year) %>%
  summarise(
    pct_cpc_below_0_7 = mean(cpc_value < ri_threshold, na.rm = TRUE) * 100,
    count = n()
  )

# 6. Plot
ggplot(summary_df, aes(x = year, y = pct_cpc_below_0_7)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "black") +
  scale_y_continuous(limits = c(0, 100), labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Year",
    y = paste0("% CPC Index < ", ri_threshold, " (where DM_median ≥ 4)"),
    title = "Frequency of Low CPC Index During High Drought Periods"
  ) +
  theme_minimal(base_size = 14)



```

## missed payouts
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)


# 4. Filter to DM >= 3 observations
filtered <- joined %>% filter(!is.na(dm_value), dm_value >= 3)

# 5. Calculate counts by year
summary_counts <- filtered %>%
  group_by(year) %>%
  summarise(
    count_dm_ge_3 = n(),
    count_missed_payouts = sum(cpc_value >= ri_threshold, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(count_dm_ge_3, count_missed_payouts),
    names_to = "metric",
    values_to = "count"
  )

# 6. Plot two lines
ggplot(summary_counts, aes(x = year, y = count, color = metric)) +
  geom_line(size = 1.2) +
  geom_point() +
  scale_color_manual(
    values = c(count_dm_ge_3 = "steelblue", count_missed_payouts = "firebrick"),
    labels = c("DM ≥ 3 observations", paste0("Missed payouts (RI ≥ ", ri_threshold, " & DM ≥ 3)"))
  ) +
  labs(
    x = "Year",
    y = "Count of observations",
    color = "Metric",
    title = "Counts by Year of Drought Severity and Missed Payouts"
  ) +
  theme_minimal(base_size = 14)

# 4. Filter to DM >= 3 observations
filtered <- joined %>% filter(!is.na(dm_value), dm_value >= 4)

# 5. Calculate counts by year
summary_counts <- filtered %>%
  group_by(year) %>%
  summarise(
    count_dm_ge_4 = n(),
    count_missed_payouts = sum(cpc_value >= ri_threshold, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # Add missing years 2010–2025
  complete(
    year = 2010:2025,
    fill = list(
      count_dm_ge_4 = 0,
      count_missed_payouts = 0
    )
  ) %>%
  
  pivot_longer(
    cols = c(count_dm_ge_4, count_missed_payouts),
    names_to = "metric",
    values_to = "count"
  )

# 6. Plot two lines
ggplot(summary_counts, aes(x = year, y = count, color = metric)) +
  geom_line(size = 1.2) +
  geom_point() +
  scale_color_manual(
    values = c(count_dm_ge_4 = "steelblue", count_missed_payouts = "firebrick"),
    labels = c("DM ≥ 4 observations", paste0("Missed payouts (RI ≥ ", ri_threshold, " & DM ≥ 4)") )
  ) +
  labs(
    x = "Year",
    y = "Count of observations",
    color = "Metric",
    title = "Counts by Year of Drought Severity and Missed Payouts"
  ) +
  theme_minimal(base_size = 14)

```

## heatmap, DM ≥3

```{r}

# Filter for DM >= 3 only
filtered <- joined %>% filter(!is.na(dm_value), dm_value >= 3)

# Summarize counts by year × interval
heatmap_data <- filtered %>%
  group_by(year, interval) %>%
  summarise(
    count_dm_ge_3 = n(),
    count_missed_payouts = sum(cpc_value >= ri_threshold, na.rm = TRUE),
    pct_missed_payouts = 100 * count_missed_payouts / count_dm_ge_3,
    .groups = "drop"
  )

interval_start_months <- 1:11
names(interval_start_months) <- 625:635
interval_end_months <- 2:12
names(interval_end_months) <- 625:635

interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]], 
    "-", 
    month.abb[interval_end_months[as.character(625:635)]]
  )
)

heatmap_data <- heatmap_data %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))

ggplot(heatmap_data, aes(x = label, y = factor(year), fill = count_missed_payouts)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Count Missed Payouts", option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of Missed Payout Counts (DM ≥ 3 & RI ≥ ", ri_threshold, ")")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

ggplot(heatmap_data, aes(x = label, y = factor(year), fill = pct_missed_payouts)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "% Missed Payouts", option = "C", direction = -1) +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of % Missed Payouts (DM ≥ 3 & RI ≥ ", ri_threshold, ")")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )



```
## export DM = 4 counts

```{r}

joined$missed_payout <- (joined$dm_value == 4 & joined$cpc_value >= ri_threshold)

grids_missed_payouts <- joined %>%
  group_by(GRIDCODE, year) %>%
  summarise(
    count_4 = sum(dm_value == 4, na.rm = TRUE),
    count_4_and_true = sum(dm_value == 4 & missed_payout == TRUE, na.rm = TRUE),
    .groups = "drop"
  )

grids_missed_payouts$missed_payout_count <- grids_missed_payouts$count_4_and_true/25

grids_missed_payouts <- grids_missed_payouts %>%
  mutate(
    missed_payout_pct = ifelse(count_4 == 0, 0, 100*(count_4_and_true/count_4))
  )

save(grids_missed_payouts, file = here("data/grids_missed_payouts_dm4.rda"))

```



## heatmap, DM ≥ 4

```{r}

# Filter for DM >= 4 only
filtered <- joined %>% filter(!is.na(dm_value), dm_value >= 4)

# Summarize counts by year × interval
heatmap_data <- filtered %>%
  group_by(year, interval) %>%
  summarise(
    count_dm_ge_4 = n(),
    count_missed_payouts = sum(cpc_value >= ri_threshold, na.rm = TRUE),
    pct_missed_payouts = 100 * count_missed_payouts / count_dm_ge_4,
    .groups = "drop"
  )

interval_start_months <- 1:11
names(interval_start_months) <- 625:635
interval_end_months <- 2:12
names(interval_end_months) <- 625:635

interval_labels <- tibble(
  interval = 625:635,
  label = paste0(
    month.abb[interval_start_months[as.character(625:635)]], 
    "-", 
    month.abb[interval_end_months[as.character(625:635)]]
  )
)


## complete missing values
heatmap_data <- heatmap_data %>% 
  # Add missing years 2010–2025
  complete(
    year = 2010:2025,
    interval = 625:635,
    fill = list(
      count_missed_payouts = NA,
      pct_missed_payouts = NA
    ))

## add labels
heatmap_data <- heatmap_data %>%
  left_join(interval_labels, by = "interval") %>%
  mutate(label = factor(label, levels = interval_labels$label))



ggplot(heatmap_data, aes(x = label, y = factor(year), fill = count_missed_payouts)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Count Missed Payouts", option = "C", direction = -1, na.value = "white") +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of Missed Payout Counts (DM ≥ 4 & RI ≥ ", ri_threshold, ")")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

ggplot(heatmap_data, aes(x = label, y = factor(year), fill = pct_missed_payouts)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "% Missed Payouts", option = "C", direction = -1, na.value = "white") +
  labs(
    x = "Interval",
    y = "Year",
    title = paste0("Heatmap of % Missed Payouts (DM ≥ 4 & RI ≥ ", ri_threshold, ")")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )



```

## maps

```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(sf)
library(ggplot2)
library(viridis)

# Assume grids is an sf object with geometry and lots of DM_median_ & cpc_index_ columns

# Get row ID for spatial join (row number)
grids <- grids %>% mutate(row_id = row_number())

# DM_median long format
dm_long <- grids %>%
  st_drop_geometry() %>%
  select(row_id, starts_with("DM_median_")) %>%
  pivot_longer(
    cols = -row_id,
    names_to = "var",
    values_to = "dm_value"
  ) %>%
  mutate(
    var = str_remove(var, "DM_median_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )

# cpc_index long format
cpc_long <- grids %>%
  st_drop_geometry() %>%
  select(row_id, starts_with("cpc_index_")) %>%
  pivot_longer(
    cols = -row_id,
    names_to = "var",
    values_to = "ri_value"
  ) %>%
  mutate(
    var = str_remove(var, "cpc_index_"),
    year = as.integer(str_sub(var, 1, 4)),
    interval = as.integer(str_sub(var, 6))
  )

# Join DM and RI by row_id, year, interval
joined <- left_join(dm_long, cpc_long, by = c("row_id", "year", "interval"))

# Filter where DM >= 3
filtered <- joined %>%
  filter(!is.na(dm_value), dm_value >= 3)

# Define missed payout = RI >= ri_threshold
filtered <- filtered %>%
  mutate(missed_payout = ri_value >= ri_threshold)

# Aggregate per spatial cell (row_id)
agg <- filtered %>%
  group_by(row_id) %>%
  summarise(
    missed_payout_count = sum(missed_payout, na.rm = TRUE),
    total_dm_ge_3 = n(),
    missed_payout_pct = 100 * missed_payout_count / total_dm_ge_3
  )

# Join back to grids sf object
grids_map <- grids %>%
  left_join(agg, by = "row_id")

# Map 1: Missed payout counts
ggplot(grids_map) +
  geom_sf(aes(fill = missed_payout_count), color = NA) +
  scale_fill_viridis_c(option = "C", na.value = "grey90", direction = -1) +
  labs(fill = "Missed Payout Count", title = "Count of Missed Payouts per Cell") +
#  zlim(0,100) +
  theme_minimal()

# Map 2: Missed payout percentage
ggplot(grids_map) +
  geom_sf(aes(fill = missed_payout_pct), color = NA) +
  scale_fill_viridis_c(option = "C", na.value = "grey90", direction = -1) +
  labs(fill = "% Missed Payouts", title = "Percent of Missed Payouts per Cell") +
  theme_minimal()



```


## comparison


```{r}

a <- st_read("/Users/mjcecil/Documents/GitHub/usdm-prf/data/USDM_GIS/2022_USDM_M/USDM_20220802.shp")
a <- st_make_valid(a)
tx_counties <- st_read(here("data/Tx_CntyBndry_Jurisdictional_TIGER/Tx_CntyBndry_Jurisdictional_TIGER.shp"))


# 2. Ensure both are in the same CRS
a <- st_transform(a, st_crs(tx_counties))

# 3. Clip drought data to TX counties
a_tx <- st_intersection(a, tx_counties)

# 4. Identify counties not intersecting any drought zone
# First, union drought geometries
a_union <- st_union(a)

# Find areas in TX counties not covered by drought
no_drought_geom <- st_difference(tx_counties, a_union)

# Add DM = -1 to those
no_drought_sf <- no_drought_geom %>%
  mutate(DM = -1)

# 5. Combine drought data with the -1 zones
a_tx_cleaned <- a_tx %>%
  select(DM) %>%
  mutate(DM = as.numeric(as.character(DM)))  # ensure numeric

combined <- bind_rows(a_tx_cleaned, no_drought_sf)

# 6. Plot with custom colors
ggplot(combined) +
  geom_sf(aes(fill = factor(DM)), color = NA) +
  scale_fill_manual(
    values = c(
      "-1" = "darkgreen",
      "0"  = "lightgreen",
      "1"  = "yellow",
      "2"  = "orange",
      "3"  = "red",
      "4"  = "darkred"
    ),
    name = "Drought Level",
    labels = c(
      "-1" = "No Classification",
      "0" = "D0",
      "1" = "D1",
      "2" = "D2",
      "3" = "D3",
      "4" = "D4"
    )
  ) +
  theme_minimal() +
  labs(title = "USDM Aug 2, 2022") + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank())



plot(a)

grids_pt <- grids %>% st_drop_geometry() %>% st_as_sf(., coords = c("lon", "lat"), crs = 4326)

grids_joined <- st_join(grids_pt, a, left = TRUE) %>%
  mutate(DM = coalesce(DM, -1))

grids_joined$DM[is.na(grids_joined$DM)] <- -1

plot(grids_joined['DM'])

rm(a)

```

```{r}

ggplot(grids) +
  geom_sf(aes(fill = factor(DM_median_2022_631)), color = NA) +
  scale_fill_manual(
    values = c(
      "-1" = "darkgreen",
      "0"  = "lightgreen",
      "1"  = "yellow",
      "2"  = "orange",
      "3"  = "red",
      "4"  = "darkred"
    ),
    name = NULL,  # Removes legend title
    labels = c(
      "-1" = "No Classification",
      "0" = "D0",
      "1" = "D1",
      "2" = "D2",
      "3" = "D3",
      "4" = "D4"
    )
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "Max USDM July-August 2022")

```

```{r}

# 1. Recode values
grids <- grids %>%
  mutate(DM_class = ifelse(DM_median_2022_631 == 4, "D4", "Other"))

# 2. Plot with two classes
ggplot(grids) +
  geom_sf(aes(fill = DM_class), color = NA) +
  scale_fill_manual(
    values = c(
      "D4" = "darkred",
      "Other" = "grey"
    ),
    name = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "Max USDM July–August 2022")

```

## table counts for DM prevalence

```{r}

# Define year and interval ranges
years <- start_year:end_year
intervals <- 625:635

# Generate column names of interest
target_cols <- outer(years, intervals, function(y, i) paste0("DM_median_", y, "_", i)) |> as.vector()

# Filter columns that actually exist
target_cols <- target_cols[target_cols %in% names(grids_collapsed)]

# Extract and convert to matrix
dm_matrix <- grids_collapsed[, target_cols]
if ("sf" %in% class(grids_collapsed)) dm_matrix <- st_drop_geometry(dm_matrix)
dm_matrix <- as.matrix(dm_matrix)
dm_matrix <- apply(dm_matrix, 2, as.numeric)  # ensure numeric

# Calculate percent of all values >= each threshold
thresholds <- 0:4
percent_ge <- sapply(thresholds, function(t) mean(dm_matrix >= t, na.rm = TRUE) * 100)

# Create and print results table
result_df <- data.frame(
  Threshold = thresholds,
  Percent_GE = round(percent_ge, 1)
)

print(result_df)



```

## table counts for RI prevalence

```{r}

# Define year and interval ranges
years <- start_year:end_year
intervals <- 625:635

# Generate column names of interest
target_cols <- outer(years, intervals, function(y, i) paste0("cpc_index_", y, "_", i)) |> as.vector()

# Filter columns that actually exist
target_cols <- target_cols[target_cols %in% names(grids_collapsed)]

# Extract and convert to matrix
dm_matrix <- grids_collapsed[, target_cols]
if ("sf" %in% class(grids_collapsed)) dm_matrix <- st_drop_geometry(dm_matrix)
dm_matrix <- as.matrix(dm_matrix)
dm_matrix <- apply(dm_matrix, 2, as.numeric)  # ensure numeric

# Calculate percent of all values >= each threshold
thresholds <- seq(0.70, 0.90, 0.05)
percent_le <- sapply(thresholds, function(t) mean(dm_matrix <= t, na.rm = TRUE) * 100)

# Create and print results table
result_df <- data.frame(
  Threshold = thresholds,
  Percent_GE = round(percent_le, 1)
)

print(result_df)



```