---
title: "ARER workflow - index calculation"
author: Michael Cecil
output: html_notebook
---

This notebook groups raw CPC data into months and intervals, and also calculates the rainfall index using the full baseline from 1948 to year (t-2)


## load required libraries

```{r}

library(tidyverse)
library(lubridate)
library(stringr)
library(here)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```

## inputs

Folder and year inputs 

```{r}

# ---- Configuration: raw CPC data location ----------------------------------
# 1_1 is the ONLY script that reads the raw CPC daily precipitation files
# (~30 GB), which are stored OUTSIDE this repo due to size. Downstream scripts
# do NOT need these files -- they read the derived monthly_averages_na_rm.csv
# produced below. Point this at your local CPC download directory, either by
# setting the CPC_DOWNLOAD_DIR environment variable or editing the default here.
cpc_download_folder <- Sys.getenv(
  "CPC_DOWNLOAD_DIR",
  unset = "~/Documents/GitHub/prf-ri/data/cpc_downloads"   # <-- edit for your machine
)
cpc_download_folder <- path.expand(cpc_download_folder)

stopifnot(
  "CPC download folder not found -- set CPC_DOWNLOAD_DIR or edit the default in 1_1" =
    dir.exists(cpc_download_folder)
)

outputs_folder <- here("data/outputs")

if (!dir.exists(outputs_folder)) {
  dir.create(outputs_folder)
}

start_year <- 1948
end_year <- 2025

start_year_ri <- 2007
end_year_ri <- 2025

```

## functions

```{r}


# Function to process a month's worth of CPC files
process_monthly_files <- function(files) {
  file_1 <- read.csv(files[1])
  monthly_data <- file_1[c('lon', 'lat')]
  for(file in files){
    print(file)
    file_date <- sub("^.*cpc_(\\d{4}-\\d{2}-\\d{2})\\.csv$", "\\1", file)
    file_data <- read.csv(file)
  #  print(file_data)
    monthly_data[[paste0("p_", file_date)]] <- file_data$precip
  }
  monthly_data$lon <- monthly_data$lat <- NULL
#  print(monthly_data)
  monthly_data[monthly_data == -99.9] <- NA
  na_count <- rowSums(is.na(monthly_data))

  return(list(rowSums(monthly_data[,], na.rm = T),
              na_count))
}

# Function to extract month from filename
extract_month <- function(filename) {
  str_extract(basename(filename), "\\d{4}-\\d{2}")
}

# Function to update column names
replace_dots_in_colnames <- function(df) {
  colnames(df) <- gsub("\\.", "-", colnames(df))
  return(df)
}

```

## Calculate monthly averages
Convert CPC data from daily to monthly, store in monthly_averages df

```{r}


# List files matching the pattern (adjust the pattern as needed)
files <- list.files(path = cpc_download_folder,
                    pattern = "cpc_\\d{4}-\\d{2}-\\d{2}.csv",
                    full.names = T)

# Create a data frame with file paths and extracted months
file_df <- data.frame(file_path = files,
                      month = extract_month(files))


cpc_data <- read.csv(files[1])
monthly_averages <- cpc_data[c('lon', 'lat')]

for (current_month in unique(file_df$month)){
  month_files <- file_df %>% filter(month == current_month)
  month_precip <- process_monthly_files(month_files$file_path)
  monthly_averages[[paste0('precip_', current_month)]] <- month_precip[[1]]
  monthly_averages[[paste0('na_count_', current_month)]] <- month_precip[[2]]
}


## save output with monthly precipitation sums
write.csv(monthly_averages, here("data/outputs/monthly_averages_na_rm.csv"))

```

## calculate two-month interval values

```{r}

# calc index values
monthly_averages <- read.csv(here("data/outputs/monthly_averages_na_rm.csv"))
monthly_averages <- replace_dots_in_colnames(monthly_averages)

## calculate interval precipitation
for(year in start_year:end_year){
  for(month in 1:11){
    current_interval <- month + 624
    monthly_averages[[paste0("precip_",
                             year,
                             "-",
                             current_interval )]] <-
      monthly_averages[[paste0('precip_', year, "-", str_pad(month,
                                                             2,
                                                             pad = "0"))]] +
      monthly_averages[[paste0('precip_', year, "-", str_pad(month + 1,
                                                             2,
                                                             pad = "0"))]]
  }
}

```

## Calculate rainfall index (RI) values

```{r}


# calculate RI value, columns start with "cpc_index
for(year in start_year_ri:end_year_ri){
  print(year)
  for(interval in 625:635){
    current_year_precip <- monthly_averages[[paste0("precip_",
                                                    year,
                                                    "-",
                                                    interval) ]]
    baseline_precip_columns <- paste0("precip_",
                                      start_year:(year - 2),
                                      "-",
                                      interval)
    baseline_avg <- rowMeans(monthly_averages[, baseline_precip_columns ])

    monthly_averages[[paste0("cpc_index_", year, "_", interval)]] <- round(current_year_precip/baseline_avg,
                                                                           5)

  }
}

write.csv(monthly_averages, here("data/outputs/monthly_averages_na_rm.csv"))

```

## Get NA count by interval
```{r}
## get NA count by interval
monthly_averages <- read.csv(here("data/outputs/monthly_averages_na_rm.csv"))
monthly_averages <- replace_dots_in_colnames(monthly_averages)

monthly_averages$total_interval_na_count <- rowSums(is.na(monthly_averages %>% select(contains("cpc_index_2"))))

print(table(monthly_averages$total_interval_na_count))

```


```{r}
## QC check (exploratory) -- inspect valid grids; not saved, not used downstream
valid_grids <- monthly_averages %>% filter(total_interval_na_count == 0)

cpc_index_columns <- names(valid_grids)[grepl("cpc_index_2", names(valid_grids))]

valid_grids_cpc_index <- valid_grids[,cpc_index_columns]

```


