---
title: "PRF-USDM - grid intersection and RI assignment"
author: Michael Cecil
output: html_notebook
---

This notebook intersects the CPC rainfall-index grids with the RMA PRF grid
definitions for the nine-state study region, attaching the RI (cpc_index)
columns to each grid. Output: grids_intersecting.rda.


## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(tigris)   # or use rnaturalearth
options(tigris_use_cache = TRUE)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```

## inputs

```{r}

aoi_states <- c("CA", "NV", "AZ", "NM", "CO", "UT", "TX", "OK", "KS")


```


## Load pre-calculated monthly averages for CPC data, merge with RMA grid shapefile

```{r}

monthly_averages <- read.csv(here("data/outputs/monthly_averages_na_rm.csv"))

## adjust longitude in above to be negative in western hemisphere
monthly_averages$lon <- monthly_averages$lon - 360

## Load CPC grids as spatial object, merge monthly average columns to spatial object
grids <- st_read(here("data/rainfall_index_grids/official_RMA_RI_grid.shp"))
grids$lon <- rowMeans(cbind(grids$X_MIN, grids$X_MAX))
grids$lat <- rowMeans(cbind(grids$Y_MIN, grids$Y_MAX))

monthly_averages_merge <- merge(grids, monthly_averages,
                          by = c("lon", "lat"))



```

## Filter to CPC grids in states of interest

```{r}

# --- Pinned state boundaries (reproducibility) --------------------------------
# The grid set depends on the state boundary used below. tigris::states() pulls
# from the Census over the web, and the returned (cb=TRUE) geometry can change
# between downloads / cache states -- which silently shifts which CPC grids are
# kept. To make the grid set reproducible, we download the boundary ONCE and
# save it in the repo; every subsequent run loads that pinned copy.
#
# To refresh the boundary deliberately, delete the .rda below and re-run.
us_states_path <- here("data/boundaries/us_states_cb_2022.rda")

if (file.exists(us_states_path)) {
  load(us_states_path)                       # loads `us_states`
} else {
  us_states <- states(cb = TRUE, year = 2022)  # generalized boundaries (web download)
  dir.create(dirname(us_states_path), recursive = TRUE, showWarnings = FALSE)
  save(us_states, file = us_states_path)
  message("Downloaded and pinned state boundaries to: ", us_states_path)
}

sel_states <- us_states %>%
  filter(STUSPS %in% aoi_states)

sel_states <- st_transform(sel_states, st_crs(monthly_averages_merge))

idx <- st_intersects(monthly_averages_merge, sel_states)

grids_intersecting <- st_join(
  monthly_averages_merge,
  sel_states,
  join = st_intersects,
  left = FALSE   # keeps only intersecting grids
) %>% rename(STATE = STUSPS)



```


## save grids_intersecting
```{r}
save(grids_intersecting, 
     file = here("data/outputs/grids_intersecting.rda"))

```



## plot

```{r}
plot(st_geometry(monthly_averages_merge), border = "grey")
plot(st_geometry(sel_states), add = TRUE, col = NA, border = "red", lwd = 2)
plot(st_geometry(grids_intersecting), add = TRUE, col = "blue", alpha = 0.5)

```


