---
title: "PRF-USDM - county-grid area weights and raw TPU (no simulation)"
author: Michael Cecil
output: html_notebook
---

This notebook computes county-to-grid area proportions (cpc_county_areas.rda)
and saves the raw PRF-Grazing TPU records per year (tpu_Grazing_<year>_no_sim.rda)
using the county SOB numbers. It does NOT run the per-grid payout simulation
(that approach was superseded; see the archived 1_3_payout_calculations.Rmd).

## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(tigris)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```

## Inputs

```{r}

start_year_payouts <- 2010
end_year_payouts <- 2025

```


## Payouts

First determine payouts for given inputs (year, grid, interval, coverage level, productivity factor, grazing/haying, )

Assumptions made:
- constant productivity factor of 1
- rates are constant across years
- rates are consistent despite changing temporal baseline or spatial resolution
- acreage proportionally split among cells in county
- liability, premia, payouts calculated based on acreage for each interval

Equations used:
Dollar Amount of Insurance Per Acre = County-Base Value * Productivity Factor (also = Liability per Acre?)
Liability = Dollar Amount * # Acres
Premium per acre = Liability per acre * rate 
Payment factor = (trigger_threshold - RI)/trigger_threshold
Indemnity = Liability * payment factor 

Columns to add
CHIRPS data set
- rates, 
- payment factor for each interval x year


## load rates data, join to grids_intersecting

Note - the rates files have a funny format. They are actually downloading two pieces of information. The Premium.Rate is based on the coverage level (70%, 75% ... 90% etc), while the County.Base.Value is based on the use type (Grazing, Haying non irrigated etc).

We will just be focusing on Grazing for now as that is the largest use case. 

```{r}

load(file = paste0(here("data/outputs/grids_intersecting.rda")))

## get state names in usable vector
states <- unique(grids_intersecting$STATE)
full_names <- state.name[match(states, state.abb)]
lower_hyphen_names <- gsub(" ", "-", tolower(full_names)) # Convert to lowercase + hyphenated




```

## create CPC area proportion files
output table has columns "State.Code"     "County.Code"    "CPC.Grid.Code"  "Area"           "Proportion"    

```{r}

## download counties from tigris
state_fips <- fips_codes %>%   # built into tigris
  filter(state %in% states) %>%
  distinct(state, state_code)

# --- Pinned county boundaries (reproducibility) -------------------------------
# These counties feed the county-grid area weights (cpc_county_areas), which
# flow into the enrollment panel. tigris::counties() is a web download whose
# geometry can vary between runs, so we pin it: download once, save in the repo,
# load the pinned copy thereafter. Delete the .rda to refresh deliberately.
counties_path <- here("data/boundaries/counties_cb_2022.rda")

if (file.exists(counties_path)) {
  load(counties_path)                        # loads `counties_sf`
} else {
  counties_sf <- counties(
    state = states,
    cb = TRUE,      # cartographic boundary (simpler, faster)
    year = 2022
  )
  dir.create(dirname(counties_path), recursive = TRUE, showWarnings = FALSE)
  save(counties_sf, file = counties_path)
  message("Downloaded and pinned county boundaries to: ", counties_path)
}

aea_crs <- 5070  # US Albers Equal Area

counties_sf <- st_transform(counties_sf, aea_crs)
grids_sf    <- st_transform(grids_intersecting, aea_crs)

county_grid_intersections <- st_intersection(
  counties_sf,
  grids_sf
)

county_grid_intersections <- county_grid_intersections %>%
  mutate(
    Area = as.numeric(st_area(geometry))
  )

area_table <- county_grid_intersections %>%
  group_by(STATEFP, COUNTYFP) %>%
  mutate(
    Proportion = Area / sum(Area)
  ) %>%
  ungroup()

cpc_county_areas <- area_table %>%
  mutate(
    # ---- Your requested schema ----
    State.Code    = as.numeric(STATEFP),  # numeric state FIPS
    State         = STUSPS,               # TX, ND, etc.
    County.Code   = COUNTYFP,             # 3-digit county FIPS
    CPC.Grid.Code = GRIDCODE
  ) %>%
  st_drop_geometry() %>%
  select(
    State.Code,
    State,
    County.Code,
    CPC.Grid.Code,
    Area,
    Proportion
  )

cpc_county_areas$County.Code <- as.numeric(cpc_county_areas$County.Code)


save(cpc_county_areas, file = here("data/outputs/cpc_county_areas.rda"))

```

## validation check

```{r}
sums_by_county <- cpc_county_areas %>%
  group_by(State.Code, County.Code) %>%
  summarise(sum_prop = sum(Proportion))

```


## save raw (no-sim) TPU files, one per year
## These are the filtered PRF-Grazing TPU records with raw SOB columns only
## (no simulated payouts). Downstream, 2_3 stacks these into tpu_all_years and
## collapses to a county-year panel using the raw SOB fields.
## Filters mirror payment_calcs() in the full 1_3 script, minus the simulation.

```{r}

payouts_outputs_folder <- here("data/outputs/payouts")

if (!dir.exists(payouts_outputs_folder)) {
  dir.create(payouts_outputs_folder, recursive = TRUE)
}

for (year in start_year_payouts:end_year_payouts) {
  print(year)

  year_2 <- substr(year, 3, 4)
  tpu_names <- read.csv(here("data/PRF_tpu/tpu_names.csv"))
  tpu <- read.csv(here(paste0("data/PRF_tpu/SOBSCCTPU", year_2, ".TXT")),
                  sep = "|", header = F)
  names(tpu) <- names(tpu_names)

  tpu$Type.Name <- trimws(tpu$Type.Name) # remove white space

  if (year <= 2012) { ## 2012 had different type names
    tpu[tpu$Type.Name == "Grazingland", "Type.Name"] <- "Grazing"
    tpu[tpu$Type.Name == "Hayland",     "Type.Name"] <- "Haying"
    tpu[tpu$Type.Name == "GRAZINGLAND", "Type.Name"] <- "Grazing"
    tpu[tpu$Type.Name == "HAYLAND",     "Type.Name"] <- "Haying"
  }

  tpu$Commodity.Code <- as.numeric(tpu$Commodity.Code)

  tpu <- tpu %>%
    filter(State.Abbreviation %in% states) %>%
    filter(Liability.Amount > 0) %>%
    filter(Commodity.Code == 88) %>%   ## for PRF
    filter(Type.Name == "Grazing")

  print(nrow(tpu))

  save(tpu, file = here(paste0("data/outputs/payouts/tpu_Grazing_", year, "_no_sim.rda")))
  print(paste0("saved no-sim TPU for ", year))
}

```







