---
title: "Untitled"
output: html_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(stringr)
library(tidyr)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```




## summary stats
### mean RI by year

```{r}
load(here("data/outputs/grids_usdm_stats.rda"))

high_drought_threshold <- 4

ri_threshold <- 0.9

start_year <- 2010
end_year <- 2025

```

## load and merge all tpu files

```{r}

# List of years
years <- start_year:end_year

# Initialize an empty list to store dataframes
tpu_list <- list()

# Loop through each year
for (yr in years) {
  # Construct file name
  file_name <- sprintf(here("data/outputs/payouts/tpu_Grazing_%d_no_sim.rda"), yr)
  
  # Load file, `tpu` will be created in the environment
  load(file_name)
  
  # Add year column
  tpu$year <- yr

  # Append to list
  tpu_list[[as.character(yr)]] <- tpu
}

# Combine all into one data frame
tpu_all_years <- do.call(rbind, tpu_list)

# Check the result
head(tpu_all_years)

save(tpu_all_years, file =  here("data/outputs/tpu_all_years.rda"))

```


## collapse tpu_all_years to county-year (combining different intervals, coverage levels etc)

```{r}


sum_cols <- c(
  "Net.Reporting.Level.Amount",
  "Liability.Amount",
  "Total.Premium.Amount",
  "Subsidy.Amount",
  "Indemnity.Amount",
  "Endorsed.Commodity.Reporting.Level.Amount"
  # "Liability_cpc_index",
  # "Premium_cpc_index",
  # "Indemnity_cpc_index"
)

tpu_collapsed <- tpu_all_years %>%
  
  # Drop unwanted columns
  select(-Loss.Ratio) %>%
  
  # Group by the collapse keys
  group_by(year, State.Code, County.Code, State.Abbreviation, State.Name) %>%
  
  # Summarise
  summarise(
    
    # Sum the specified columns
    across(all_of(sum_cols), ~ sum(.x, na.rm = TRUE)),
    
    # Take first value of all remaining columns
    across(
      -all_of(sum_cols),
      ~ first(.x)
    ),
    
    .groups = "drop"
  )





```



## add columns for loss ratio and ROI (based on raw or estimated data)

```{r}

tpu_collapsed$loss_ratio_raw <- tpu_collapsed$Indemnity.Amount/tpu_collapsed$Total.Premium.Amount ## ROI


```


## add associated variables to tpu_all
### drought exposure 

### % grid-intervals with DM_max >= high drought
### mean USDM drought across all grid_intervals
### median USDM drought across all grid_intervals
### max USDM drought across all grid_intervals


```{r}

## load county areas file to identify grids in each county
load(file =  here("data/outputs/cpc_county_areas.rda"))
cpc_county_areas$County.Code <- as.numeric(cpc_county_areas$County.Code)


# Create empty vector to store results
DM_max_values    <- rep(NA_real_, nrow(tpu_collapsed))
DM_mean_values   <- rep(NA_real_, nrow(tpu_collapsed))
DM_median_values <- rep(NA_real_, nrow(tpu_collapsed))
DM_high_pct_values <- rep(NA_real_, nrow(tpu_collapsed))

# Loop over each row of tpu_collapsed
for (i in seq_len(nrow(tpu_collapsed))) {
  if(i %% 1000 == 0){
      print(i)
  }
  # Extract identifiers
  county <- tpu_collapsed$County.Code[i]
  state <- tpu_collapsed$State.Code[i]
  year <- tpu_collapsed$year[i]

  # Step 2: Filter cpc_county_areas 
  area_rows <- cpc_county_areas %>% 
    filter(State.Code == state) %>% 
    filter(County.Code == county)
  
  # Skip if no matching area rows
  if (nrow(area_rows) == 0) {
    print("no rows available")
    print(state)
    print(paste0("county, ", county))
    DM_max_values[i] <- NA
    DM_mean_values[i] <- NA
    DM_median_values[i] <- NA
    DM_high_pct_values[i] <- NA
    next
  }

  # Step 2b: Extract unique CPC grid codes
  grid_codes <- unique(area_rows$CPC.Grid.Code)

  # Step 3: Filter grids
  county_rows <- grids[grids$GRIDCODE %in% grid_codes, ]
  
  pattern_max <- paste0("^DM_max_", year, "_6(2[5-9]|3[0-5])$")
  pattern_median <- paste0("^DM_median_", year, "_6(2[5-9]|3[0-5])$")
  pattern_mean <- paste0("^DM_mean_", year, "_6(2[5-9]|3[0-5])$")

  # Step 4: Get values for DM indices
  DM_max_values[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_max)) %>%
    summarise(across(everything(), ~ max(.x, na.rm = TRUE))) %>%
    unlist() %>%
    max(na.rm = TRUE)
  
  DM_mean_values[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_mean)) %>%
    summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
    unlist() %>%
    mean(na.rm = TRUE)
  
  DM_median_values[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_median)) %>%
    summarise(across(everything(), ~ median(.x, na.rm = TRUE))) %>%
    unlist() %>%
    median(na.rm = TRUE)
  
  DM_high_pct_values[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_max)) %>%
    summarise(across(everything(), ~ mean(.x >= high_drought_threshold , na.rm = TRUE))) %>%
    unlist() %>%
    mean(na.rm = TRUE)
}

# Add columns to tpu_collapsed
tpu_collapsed$DM_max <- DM_max_values
tpu_collapsed$DM_mean <- DM_mean_values
tpu_collapsed$DM_median <- DM_median_values
tpu_collapsed$DM_high_pct <- DM_high_pct_values


save(tpu_collapsed, file = here("data/outputs/tpu_collapsed.rda"))




```


### payout variables
#### cost ratio variables are already calculated
#### need to calculate % payout frequency

```{r}


# Create empty vector to store results
RI_payout_pct <- numeric(nrow(tpu_collapsed))

# Loop over each row of tpu_collapsed
for (i in seq_len(nrow(tpu_collapsed))) {
  if(i %% 1000 == 0){
      print(i)
  }
  # Extract identifiers
  county <- tpu_collapsed$County.Code[i]
  state <- tpu_collapsed$State.Code[i]
  year <- tpu_collapsed$year[i]

  # Step 2: Filter cpc_county_areas 
  area_rows <- cpc_county_areas %>% 
    filter(State.Code == state) %>% 
    filter(County.Code == county)
  
  # Skip if no matching area rows
  if (nrow(area_rows) == 0) {
    print("no rows available")
    print(state)
    print(paste0("county, ", county))
    RI_payout_pct[i] <- NA
    next
  }

  # Step 2b: Extract unique CPC grid codes
  grid_codes <- unique(area_rows$CPC.Grid.Code)

  # Step 3: Filter grids
  county_rows <- grids[grids$GRIDCODE %in% grid_codes, ]
  
  pattern_RI <- paste0("^cpc_index_", year, "_6(2[5-9]|3[0-5])$")

  
  # Step 4: Get values for DM indices
  RI_payout_pct[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_RI)) %>%
    summarise(across(everything(), ~ mean(.x <= ri_threshold, na.rm = TRUE))) %>%
    unlist() %>%
    mean(na.rm = TRUE)
  
}

# Add columns to tpu_collapsed
tpu_collapsed$RI_payout_pct <- RI_payout_pct

save(tpu_collapsed, file = here("data/outputs/tpu_collapsed.rda"))




```


### missed payout (basis risk) variables

```{r}

# Create empty vectors
missed_payout_count_all  <- numeric(nrow(tpu_collapsed))  # count of D4 & no payout intervals
missed_payout_rate_all   <- numeric(nrow(tpu_collapsed))  # % of ALL intervals with D4 & no payout
dm_true_count  <- numeric(nrow(tpu_collapsed))  # count of D4 intervals
missed_payout_rate_d4    <- numeric(nrow(tpu_collapsed))  # % of D4 intervals with no payout

for (i in seq_len(nrow(tpu_collapsed))) {
  if(i %% 1000 == 0) print(i)

  county <- tpu_collapsed$County.Code[i]
  state  <- tpu_collapsed$State.Code[i]
  year   <- tpu_collapsed$year[i]

  # Filter county areas
  area_rows <- cpc_county_areas %>%
    filter(State.Code == state) %>%
    filter(County.Code == county)

  if (nrow(area_rows) == 0) {
    missed_payout_count_all[i] <- NA
    missed_payout_rate_all[i]  <- NA
    dm_true_count[i]    <- NA
    missed_payout_rate_d4[i]   <- NA
    next
  }

  grid_codes  <- unique(area_rows$CPC.Grid.Code)
  county_rows <- grids[grids$GRIDCODE %in% grid_codes, ] %>% st_drop_geometry()

  # Column patterns for this year
  pattern_DM <- paste0("^DM_median_", year, "_6(2[5-9]|3[0-5])$")
  pattern_RI <- paste0("^cpc_index_", year, "_6(2[5-9]|3[0-5])$")

  dm_cols <- grep(pattern_DM, names(county_rows), value = TRUE)
  ri_cols <- grep(pattern_RI, names(county_rows), value = TRUE)

  # Align columns (same intervals)
  # Extract interval suffix and match
  dm_intervals <- gsub(".*_(6[0-9]{2})$", "\\1", dm_cols)
  ri_intervals <- gsub(".*_(6[0-9]{2})$", "\\1", ri_cols)
  common_intervals <- base::intersect(dm_intervals, ri_intervals)

  dm_cols_use <- paste0("DM_median_", year, "_", common_intervals)
  ri_cols_use <- paste0("cpc_index_", year, "_", common_intervals)

  dm_mat <- as.matrix(county_rows[, dm_cols_use, drop = FALSE])
  ri_mat <- as.matrix(county_rows[, ri_cols_use, drop = FALSE])

  # Total valid intervals
  valid_mat <- !is.na(dm_mat) & !is.na(ri_mat)
  total_valid <- sum(valid_mat)

  if (total_valid == 0) {
    missed_payout_count_all[i] <- NA
    missed_payout_rate_all[i]  <- NA
    dm_true_count[i]    <- NA
    missed_payout_rate_d4[i]   <- NA
    next
  }

  # D4 intervals
  is_d4      <- dm_mat >= high_drought_threshold & valid_mat
  # Missed payout: D4 but RI > threshold (no payout)
  is_missed  <- is_d4 & ri_mat > ri_threshold

  n_d4       <- sum(is_d4,     na.rm = TRUE)
  n_missed   <- sum(is_missed, na.rm = TRUE)

  missed_payout_count_all[i] <- n_missed
  missed_payout_rate_all[i]  <- n_missed / total_valid          # % of ALL intervals
  dm_true_count[i]    <- n_d4
  missed_payout_rate_d4[i]   <- ifelse(n_d4 > 0,
                                        n_missed / n_d4,        # % of D4 intervals
                                        NA)
}

tpu_collapsed$missed_payout_count_all <- missed_payout_count_all
tpu_collapsed$missed_payout_rate_all  <- missed_payout_rate_all
tpu_collapsed$dm_true_count    <- dm_true_count
tpu_collapsed$missed_payout_rate_d4   <- missed_payout_rate_d4

save(tpu_collapsed, file = here("data/outputs/tpu_collapsed.rda"))

```

## spot-check

```{r}
# # ── Spot check: 2016, State 6, County 29 ─────────────────────────────────────
# check_year   <- 2016
# check_state  <- 6
# check_county <- 29
# 
# # Step 1: Get grid codes for this county
# area_rows_check <- cpc_county_areas %>%
#   filter(State.Code == check_state) %>%
#   filter(County.Code == check_county)
# 
# cat("Number of area rows:", nrow(area_rows_check), "\n")
# 
# grid_codes_check <- unique(area_rows_check$CPC.Grid.Code)
# cat("Number of unique grid codes:", length(grid_codes_check), "\n")
# 
# # Step 2: Get grid rows
# county_rows_check <- grids[grids$GRIDCODE %in% grid_codes_check, ] %>%
#   st_drop_geometry()
# 
# cat("Number of grid rows:", nrow(county_rows_check), "\n")
# 
# # Step 3: Get DM and RI columns for this year
# dm_cols_check <- grep(paste0("^DM_median_", check_year, "_6(2[5-9]|3[0-5])$"),
#                       names(county_rows_check), value = TRUE)
# ri_cols_check <- grep(paste0("^cpc_index_", check_year, "_6(2[5-9]|3[0-5])$"),
#                       names(county_rows_check), value = TRUE)
# 
# cat("\nDM columns found:", dm_cols_check, "\n")
# cat("RI columns found:", ri_cols_check, "\n")
# 
# # Step 4: Build matrices
# dm_mat_check <- as.matrix(county_rows_check[, dm_cols_check, drop = FALSE])
# ri_mat_check <- as.matrix(county_rows_check[, ri_cols_check, drop = FALSE])
# 
# cat("\nDimensions of DM matrix:", dim(dm_mat_check), "\n")
# cat("Dimensions of RI matrix:", dim(ri_mat_check), "\n")
# 
# # Step 5: Compute statistics manually
# valid_check   <- !is.na(dm_mat_check) & !is.na(ri_mat_check)
# total_valid   <- sum(valid_check)
# is_d4_check   <- dm_mat_check >= high_drought_threshold & valid_check
# is_missed_check <- is_d4_check & ri_mat_check > ri_threshold
# 
# n_d4_check     <- sum(is_d4_check,     na.rm = TRUE)
# n_missed_check <- sum(is_missed_check, na.rm = TRUE)
# 
# cat("\n=== Spot Check Results: Year", check_year,
#     "| State", check_state,
#     "| County", check_county, "===\n")
# cat("Total valid intervals:          ", total_valid, "\n")
# cat("D4 intervals:                   ", n_d4_check, "\n")
# cat("Missed payout intervals:        ", n_missed_check, "\n")
# cat("missed_payout_rate_all:         ", round(n_missed_check / total_valid, 4), "\n")
# cat("missed_payout_rate_d4:          ",
#     ifelse(n_d4_check > 0, round(n_missed_check / n_d4_check, 4), NA), "\n")
# 
# # Step 6: Compare to stored values
# stored <- tpu_collapsed %>%
#   filter(year == check_year,
#          State.Code == check_state,
#          County.Code == check_county)
# 
# cat("\n=== Stored values in tpu_collapsed ===\n")
# cat("missed_payout_count_all:  ", stored$missed_payout_count_all, "\n")
# cat("missed_payout_rate_all:   ", stored$missed_payout_rate_all,  "\n")
# cat("dm_true_count:     ", stored$dm_true_count,    "\n")
# cat("missed_payout_rate_d4:    ", stored$missed_payout_rate_d4,   "\n")
# 
# # Step 7: Show raw values for inspection
# cat("\n=== Raw DM values by interval ===\n")
# dm_summary <- data.frame(
#   interval   = dm_cols_check,
#   mean_DM    = colMeans(dm_mat_check, na.rm = TRUE),
#   pct_D4     = colMeans(dm_mat_check >= high_drought_threshold, na.rm = TRUE)
# )
# print(dm_summary)
# 
# cat("\n=== Raw RI values by interval ===\n")
# ri_summary <- data.frame(
#   interval      = ri_cols_check,
#   mean_RI       = colMeans(ri_mat_check, na.rm = TRUE),
#   pct_payout    = colMeans(ri_mat_check <= ri_threshold, na.rm = TRUE)
# )
# print(ri_summary)
```


### join relevant SOB columns (Policies Earning Premium Count, Units Earning Premium Count)
### very slow

```{r}
# 
# load(file = here("data/outputs/tpu_collapsed.rda"))
# 
# total_policies_sold_vct <- rep(NA_real_, nrow(tpu_collapsed))
# total_policies_earning_vct <- rep(NA_real_, nrow(tpu_collapsed))
# total_units_earning_vct <- rep(NA_real_, nrow(tpu_collapsed))
# 
# 
# 
# # Loop over each row of tpu_collapsed
# for (i in seq_len(nrow(tpu_collapsed))) {
#   if(i %% 100 == 0){
#       print(i)
#   }
#   # Extract identifiers
#   county <- tpu_collapsed$County.Code[i]
#   state <- tpu_collapsed$State.Code[i]
#   year <- tpu_collapsed$year[i]
#   
#   # Get column names
#   data_w_names <- read.csv(here("data/PRF_sob/sobcov_2024_names.csv"))
#   column_names <- colnames(data_w_names)
#   
#   # Read the data file for this year
#   year_2 <- substr(as.character(year), 3, 4)
#   sob_data <- read.csv(here(paste0("data/PRF_sob/raw_files/sobcov", year_2, ".txt")),
#                    sep = '|')
#   names(sob_data) <- column_names
#   
#     # Filter for the specific commodity and insurance plan
#   data_prf <- sob_data %>%
#     filter(Commodity.Code == 88) %>% 
#     filter(Insurance.Plan.Code == 13) %>% 
#     filter(Location.State.Code == state) %>% 
#     filter(Location.County.Code == county) %>% 
#     select(Policies.Sold.Count, Policies.Earning.Premium.Count, Units.Earning.Premium.Count) 
#   
#   n_obs <- nrow(data_prf)
#   if(n_obs == 0){
#      total_policies_sold_vct[i] <- 0
#      total_policies_earning_vct[i] <- 0
#      total_units_earning_vct[i] <- 0
#      next
#   }
# 
#   total_policies_sold_vct[i] <- sum(data_prf$Policies.Sold.Count)
#   total_policies_earning_vct[i] <- sum(data_prf$Policies.Earning.Premium.Count)
#   total_units_earning_vct[i] <- sum(data_prf$Units.Earning.Premium.Count)
# 
# }
# 
# tpu_collapsed$total_policies_sold <- total_policies_sold_vct
# tpu_collapsed$total_policies_earning <- total_policies_earning_vct
# tpu_collapsed$total_units_earning <- total_units_earning_vct
# 
# save(tpu_collapsed, file = here("data/outputs/tpu_collapsed.rda"))

```








