---
title: "Untitled"
output: html_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load required libraries

```{r}

library(dplyr)
library(here)
library(sf)
library(raster)
library(conflicted)
library(rlang)
library(scico)
library(ggpubr)
library(assertthat)
library(ggplot2)
library(stringr)
library(tidyr)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
options(max.print = 10000)
```




## summary stats
### mean RI by year

```{r}
load(here("data/outputs/grids_usdm_stats.rda"))

high_drought_threshold <- 4

ri_threshold <- 0.9

start_year <- 2010
end_year <- 2025

```

## load and merge all tpu files

```{r}

# List of years
years <- start_year:end_year

# Initialize an empty list to store dataframes
tpu_list <- list()

# Loop through each year
for (yr in years) {
  # Construct file name
  file_name <- sprintf(here("data/outputs/payouts/tpu_Grazing_%d_no_sim.rda"), yr)
  
  # Load file, `tpu` will be created in the environment
  load(file_name)
  
  # Add year column
  tpu$year <- yr

  # Append to list
  tpu_list[[as.character(yr)]] <- tpu
}

# Combine all into one data frame
tpu_all_years <- do.call(rbind, tpu_list)

# Check the result
head(tpu_all_years)

save(tpu_all_years, file =  here("data/outputs/tpu_all_years.rda"))

```


## collapse tpu_all_years to county-year (combining different intervals, coverage levels etc)

```{r}


sum_cols <- c(
  "Net.Reporting.Level.Amount",
  "Liability.Amount",
  "Total.Premium.Amount",
  "Subsidy.Amount",
  "Indemnity.Amount",
  "Endorsed.Commodity.Reporting.Level.Amount"
  # "Liability_cpc_index",
  # "Premium_cpc_index",
  # "Indemnity_cpc_index"
)

tpu_collapsed <- tpu_all_years %>%
  
  # Drop unwanted columns
  select(-Loss.Ratio) %>%
  
  # Group by the collapse keys
  group_by(year, State.Code, County.Code, State.Abbreviation, State.Name) %>%
  
  # Summarise
  summarise(
    
    # Sum the specified columns
    across(all_of(sum_cols), ~ sum(.x, na.rm = TRUE)),
    
    # Take first value of all remaining columns
    across(
      -all_of(sum_cols),
      ~ first(.x)
    ),
    
    .groups = "drop"
  )





```



## add columns for loss ratio and ROI (based on raw or estimated data)

```{r}

tpu_collapsed$loss_ratio_raw <- tpu_collapsed$Indemnity.Amount/tpu_collapsed$Total.Premium.Amount ## ROI


```


## add associated variables to tpu_all
### drought exposure 

### % grid-intervals with DM_max >= high drought
### mean USDM drought across all grid_intervals
### median USDM drought across all grid_intervals
### max USDM drought across all grid_intervals


```{r}

## load county areas file to identify grids in each county
load(file =  here("data/outputs/cpc_county_areas.rda"))
cpc_county_areas$County.Code <- as.numeric(cpc_county_areas$County.Code)


# Create empty vector to store results
DM_max_values    <- rep(NA_real_, nrow(tpu_collapsed))
DM_mean_values   <- rep(NA_real_, nrow(tpu_collapsed))
DM_median_values <- rep(NA_real_, nrow(tpu_collapsed))
DM_high_pct_values <- rep(NA_real_, nrow(tpu_collapsed))

# Loop over each row of tpu_collapsed
for (i in seq_len(nrow(tpu_collapsed))) {
  if(i %% 1000 == 0){
      print(i)
  }
  # Extract identifiers
  county <- tpu_collapsed$County.Code[i]
  state <- tpu_collapsed$State.Code[i]
  year <- tpu_collapsed$year[i]

  # Step 2: Filter cpc_county_areas 
  area_rows <- cpc_county_areas %>% 
    filter(State.Code == state) %>% 
    filter(County.Code == county)
  
  # Skip if no matching area rows
  if (nrow(area_rows) == 0) {
    print("no rows available")
    print(state)
    print(paste0("county, ", county))
    DM_max_values[i] <- NA
    DM_mean_values[i] <- NA
    DM_median_values[i] <- NA
    DM_high_pct_values[i] <- NA
    next
  }

  # Step 2b: Extract unique CPC grid codes
  grid_codes <- unique(area_rows$CPC.Grid.Code)

  # Step 3: Filter grids
  county_rows <- grids[grids$GRIDCODE %in% grid_codes, ]
  
  pattern_max <- paste0("^DM_max_", year, "_6(2[5-9]|3[0-5])$")
  pattern_median <- paste0("^DM_median_", year, "_6(2[5-9]|3[0-5])$")
  pattern_mean <- paste0("^DM_mean_", year, "_6(2[5-9]|3[0-5])$")

  # Step 4: Get values for DM indices
  DM_max_values[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_max)) %>%
    summarise(across(everything(), ~ max(.x, na.rm = TRUE))) %>%
    unlist() %>%
    max(na.rm = TRUE)
  
  DM_mean_values[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_mean)) %>%
    summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
    unlist() %>%
    mean(na.rm = TRUE)
  
  DM_median_values[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_median)) %>%
    summarise(across(everything(), ~ median(.x, na.rm = TRUE))) %>%
    unlist() %>%
    median(na.rm = TRUE)
  
  DM_high_pct_values[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_max)) %>%
    summarise(across(everything(), ~ mean(.x >= high_drought_threshold , na.rm = TRUE))) %>%
    unlist() %>%
    mean(na.rm = TRUE)
}

# Add columns to tpu_collapsed
tpu_collapsed$DM_max <- DM_max_values
tpu_collapsed$DM_mean <- DM_mean_values
tpu_collapsed$DM_median <- DM_median_values
tpu_collapsed$DM_high_pct <- DM_high_pct_values


save(tpu_collapsed, file = here("data/outputs/tpu_collapsed.rda"))




```


### payout variables
#### cost ratio variables are already calculated
#### need to calculate % payout frequency

```{r}


# Create empty vector to store results
RI_payout_pct <- numeric(nrow(tpu_collapsed))

# Loop over each row of tpu_collapsed
for (i in seq_len(nrow(tpu_collapsed))) {
  if(i %% 1000 == 0){
      print(i)
  }
  # Extract identifiers
  county <- tpu_collapsed$County.Code[i]
  state <- tpu_collapsed$State.Code[i]
  year <- tpu_collapsed$year[i]

  # Step 2: Filter cpc_county_areas 
  area_rows <- cpc_county_areas %>% 
    filter(State.Code == state) %>% 
    filter(County.Code == county)
  
  # Skip if no matching area rows
  if (nrow(area_rows) == 0) {
    print("no rows available")
    print(state)
    print(paste0("county, ", county))
    RI_payout_pct[i] <- NA
    next
  }

  # Step 2b: Extract unique CPC grid codes
  grid_codes <- unique(area_rows$CPC.Grid.Code)

  # Step 3: Filter grids
  county_rows <- grids[grids$GRIDCODE %in% grid_codes, ]
  
  pattern_RI <- paste0("^cpc_index_", year, "_6(2[5-9]|3[0-5])$")

  
  # Step 4: Get values for DM indices
  RI_payout_pct[i] <- county_rows %>% st_drop_geometry() %>% 
    select(matches(pattern_RI)) %>%
    summarise(across(everything(), ~ mean(.x <= ri_threshold, na.rm = TRUE))) %>%
    unlist() %>%
    mean(na.rm = TRUE)
  
}

# Add columns to tpu_collapsed
tpu_collapsed$RI_payout_pct <- RI_payout_pct

save(tpu_collapsed, file = here("data/outputs/tpu_collapsed.rda"))




```





### join relevant SOB columns (Policies Earning Premium Count, Units Earning Premium Count)
### very slow

```{r}

load(file = here("data/outputs/tpu_collapsed.rda"))

total_policies_sold_vct <- rep(NA_real_, nrow(tpu_collapsed))
total_policies_earning_vct <- rep(NA_real_, nrow(tpu_collapsed))
total_units_earning_vct <- rep(NA_real_, nrow(tpu_collapsed))



# Loop over each row of tpu_collapsed
for (i in seq_len(nrow(tpu_collapsed))) {
  if(i %% 100 == 0){
      print(i)
  }
  # Extract identifiers
  county <- tpu_collapsed$County.Code[i]
  state <- tpu_collapsed$State.Code[i]
  year <- tpu_collapsed$year[i]
  
  # Get column names
  data_w_names <- read.csv(here("data/PRF_sob/sobcov_2024_names.csv"))
  column_names <- colnames(data_w_names)
  
  # Read the data file for this year
  year_2 <- substr(as.character(year), 3, 4)
  sob_data <- read.csv(here(paste0("data/PRF_sob/raw_files/sobcov", year_2, ".txt")),
                   sep = '|')
  names(sob_data) <- column_names
  
    # Filter for the specific commodity and insurance plan
  data_prf <- sob_data %>%
    filter(Commodity.Code == 88) %>% 
    filter(Insurance.Plan.Code == 13) %>% 
    filter(Location.State.Code == state) %>% 
    filter(Location.County.Code == county) %>% 
    select(Policies.Sold.Count, Policies.Earning.Premium.Count, Units.Earning.Premium.Count) 
  
  n_obs <- nrow(data_prf)
  if(n_obs == 0){
     total_policies_sold_vct[i] <- 0
     total_policies_earning_vct[i] <- 0
     total_units_earning_vct[i] <- 0
     next
  }

  total_policies_sold_vct[i] <- sum(data_prf$Policies.Sold.Count)
  total_policies_earning_vct[i] <- sum(data_prf$Policies.Earning.Premium.Count)
  total_units_earning_vct[i] <- sum(data_prf$Units.Earning.Premium.Count)

}

tpu_collapsed$total_policies_sold <- total_policies_sold_vct
tpu_collapsed$total_policies_earning <- total_policies_earning_vct
tpu_collapsed$total_units_earning <- total_units_earning_vct

save(tpu_collapsed, file = here("data/outputs/tpu_collapsed.rda"))

```








