---
title: "Untitled"
output: html_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load required libraries

```{r}
source(here::here("_setup.R"))
```


## initial plots

## summary stats
###mean RI by year

```{r}
load(here("data/outputs/grids_usdm_stats.rda"))

grids <- grids %>% filter(!is.na(cpc_index_2010_625)) %>% filter(!(is.na(DM_max_2010_625)))

  


## NOTE: this script intentionally spans the full enrollment history (2007+),
## overriding the analysis-window start_year (2010) from _setup.R.
start_year <- 2007
```


## SOB enrollment by year - policies


```{r}

### plot for policies by state

state_policy_enrollment <- lapply(start_year:end_year, function(year){
  print(year)
  data_w_names <- read.csv(here("data/PRF_sob/sobcov_2024_names.csv"))
  column_names <- colnames(data_w_names)
  
  year_2 <- substr(as.character(year), 3, 4)
  data <- read.csv(here(paste0("data/PRF_sob/raw_files/sobcov", year_2, ".txt")),
                   sep = '|')
  names(data) <- column_names
  
  data_prf <- data %>%
    filter(Commodity.Code ==  88) %>% 
    filter(Insurance.Plan.Code == 13)
  
  print(unique(data_prf$Quantity.Type))
  
  state_sums <- data_prf %>%
    group_by(Location.State.Abbreviation) %>%
    summarize(total = sum(Policies.Sold.Count, na.rm = TRUE),
              total_acres = sum(Net.Reported.Quantity, na.rm = TRUE))
  
  state_sums$year <- year
  
  return(state_sums)
  
}) %>% do.call("rbind", .)


all_states <- unique(state_policy_enrollment$Location.State.Abbreviation)
years <- 2010:2025

# Expand the data to ensure every state-year combination is present
state_policy_enrollment <- state_policy_enrollment %>%
  complete(Location.State.Abbreviation = all_states, year = years, fill = list(total = 0))

state_policy_enrollment <- state_policy_enrollment %>% rename(State = Location.State.Abbreviation,
                                                              Policies = total)

state_policy_enrollment[["year"]] <- as.Date(paste0(state_policy_enrollment[["year"]], "-01-01"))


#---------------------------
# Label dataset
#---------------------------
label_states <- aoi_states

state_labels <- state_policy_enrollment %>%
  filter(State %in% label_states) %>%
  group_by(State) %>%
  filter(year == max(year)) %>%
  ungroup()

#---------------------------
# Plot
#---------------------------
p <- ggplot(state_policy_enrollment,
            aes(x = year,
                y = Policies/1000,
                group = State)) +

  # Lines for all states except Texas
  geom_line(data = filter(state_policy_enrollment, State != "TX"),
            aes(color = "Other"),
            linewidth = 1,
            alpha = 0.5) +

  # Texas line
  geom_line(data = filter(state_policy_enrollment, State == "TX"),
            aes(color = "TX"),
            linewidth = 1.2) +

  #---------------------------
  # Non-overlapping labels
  #---------------------------
  geom_text_repel(
    data = state_labels,
    aes(label = State),
    size = 5,
    color = "black",
    direction = "y",          # repel vertically (best for time series)
    hjust = 0,
    nudge_x = 0.3,            # push labels right of endpoints
    segment.color = "grey40",
    segment.size = 0.4,
    box.padding = 0.3,
    point.padding = 0.2,
    max.overlaps = Inf
  ) +

  # Colors
  scale_color_manual(values = c("TX" = "black",
                                "Other" = "grey")) +

  # Labels
  labs(
    x = "Data Source: USDA State-County-Crop Summary of Business",
    y = "Policies (k)",
    color = "State"
  ) +

  # X scale
  scale_x_date(
    breaks = seq(min(state_policy_enrollment$year),
                 max(state_policy_enrollment$year),
                 by = "2 years"),
    date_labels = "%Y",
    expand = c(0.05, 0)
  ) +

  # Prevent clipping of repelled labels
  coord_cartesian(clip = "off") +

  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 14),
    plot.title = element_text(size = 14),
    axis.title.x = element_text(hjust = 0.02, vjust = 0.2),
    plot.margin = margin(5.5, 40, 5.5, 5.5)  # room for labels
  )

p


```

## SOB enrollment by year - acres

```{r}

#---------------------------
# Plot for ACRES by state
#---------------------------
label_states <- aoi_states

state_labels <- state_policy_enrollment %>%
  filter(State %in% label_states) %>%
  group_by(State) %>%
  filter(year == max(year)) %>%
  ungroup()

p_acres <- ggplot(state_policy_enrollment,
                  aes(x = year,
                      y = total_acres / 1e6,
                      group = State)) +
  geom_line(data = filter(state_policy_enrollment, State != "TX"),
            aes(color = "Other"),
            linewidth = 1,
            alpha = 0.5) +
  geom_line(data = filter(state_policy_enrollment, State == "TX"),
            aes(color = "TX"),
            linewidth = 1.2) +
  geom_text_repel(
    data = state_labels,
    aes(label = State),
    size = 5,
    color = "black",
    direction = "y",
    hjust = 0,
    nudge_x = 0.3,
    segment.color = "grey40",
    segment.size = 0.4,
    box.padding = 0.3,
    point.padding = 0.2,
    max.overlaps = Inf
  ) +
  scale_color_manual(values = c("TX" = "black", "Other" = "grey")) +
  labs(
    x = "Data Source: USDA State-County-Crop Summary of Business",
    y = "Acres (millions)",
    color = "State"
  ) +
  scale_x_date(
    breaks = seq(min(state_policy_enrollment$year),
                 max(state_policy_enrollment$year),
                 by = "2 years"),
    date_labels = "%Y",
    expand = c(0.05, 0)
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 14),
    plot.title = element_text(size = 14),
    axis.title.x = element_text(hjust = 0.02, vjust = 0.2),
    plot.margin = margin(5.5, 40, 5.5, 5.5)
  )

p_acres

#---------------------------
# Table: % of national acres by state, 2010-2025
#---------------------------
acre_table <- state_policy_enrollment %>%
  mutate(year = as.integer(format(year, "%Y"))) %>%
  filter(year %in% 2010:2025) %>%
  group_by(year) %>%
  mutate(national_total = sum(total_acres, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(State) %>%
  summarise(
    total_acres_all_years  = sum(total_acres, na.rm = TRUE),
    national_total         = sum(national_total, na.rm = TRUE),  # sum of annual totals * n_years, so divide below
    .groups = "drop"
  ) %>%
  mutate(
    pct_national  = total_acres_all_years / sum(total_acres_all_years) * 100,
    in_aoi_region = State %in% aoi_states,
    rank          = rank(-total_acres_all_years, ties.method = "min")
  ) %>%
  arrange(rank)

# Region-level summary row
region_row <- acre_table %>%
  filter(in_aoi_region) %>%
  summarise(
    State                 = "9-State Region",
    total_acres_all_years = sum(total_acres_all_years),
    pct_national          = sum(pct_national),
    in_aoi_region         = TRUE,
    rank                  = NA_integer_
  )

rank_2025 <- state_policy_enrollment %>%
  mutate(year = as.integer(format(year, "%Y"))) %>%
  filter(year == 2025) %>%
  mutate(rank_2025 = rank(-total_acres, ties.method = "min")) %>%
  select(State, rank_2025)


acre_table_final <- bind_rows(acre_table, region_row) %>%
  select(State, rank, total_acres_all_years, pct_national, in_aoi_region)

acre_table_final <- acre_table_final %>%
  left_join(rank_2025, by = "State")

print(acre_table_final)

acre_table_aoi <- acre_table_final %>% filter(in_aoi_region == TRUE)
print(acre_table_aoi)


```

```{r}

acre_table_export <- acre_table_aoi %>%
  arrange(rank) %>%
  mutate(
    total_acres_millions = round(total_acres_all_years / 1e6),
    pct_national         = paste0(round(pct_national), "%")  # plain %, let escape_latex handle it
  ) %>%
  select(
    State,
    `Enrolled Acres (M)`     = total_acres_millions,
    `Share of US Acreage` = pct_national,
    `Rank (2010-2025)`    = rank
  ) %>%
  mutate(across(everything(), as.character)) %>%
  mutate(
    `Rank (2010-2025)` = ifelse(`Rank (2010-2025)` == "NA", "---", `Rank (2010-2025)`)
  )

# Replace NA in other columns
acre_table_export[acre_table_export == "NA"] <- "---"

export_df_to_latex(
  df        = acre_table_export,
  file_path = here("data/outputs/tables/acre_table_aoi.tex"),
  caption = "PRF-RI enrolled acres by state, summed across 2010--2025 (acre-years). Share of US acreage and rank are computed on the same cumulative basis."
)
```


```{r}
#---------------------------
# Plot for ACRES % change by state (AOI states only)
#---------------------------


# Calculate % change from previous year

state_pct_change <- state_policy_enrollment %>%
  filter(State %in% aoi_states) %>%
  arrange(State, year) %>%
  group_by(State) %>%
  mutate(pct_change = (total_acres - lag(total_acres)) / lag(total_acres) * 100) %>%
  filter(!is.na(pct_change)) %>%
  ungroup()

# Labels at last year
state_labels_pct <- state_pct_change %>%
  group_by(State) %>%
  filter(year == max(year)) %>%
  ungroup()

p_acres_pct <- ggplot(state_pct_change,
                      aes(x = year, y = pct_change, group = State)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_line(linewidth = 1, color = "black") +
  geom_point(size = 1.5, color = "black") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    x = "Data Source: USDA State-County-Crop Summary of Business",
    y = "Year-over-Year Change in Acres (%)",
  ) +
  scale_x_date(
    breaks = seq(min(state_pct_change$year),
                 max(state_pct_change$year),
                 by = "4 years"),
    date_labels = "%Y",
    expand = c(0.05, 0)
  ) +
  coord_cartesian(clip = "off", ylim = c(NA, 1000)) +
  facet_wrap(~ State, scales = "free_y") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14),
    axis.title.x = element_text(hjust = 0.02, vjust = 0.2),
    strip.text = element_text(size = 13, face = "bold"),
    plot.margin = margin(5.5, 10, 5.5, 5.5)
  )

p_acres_pct
```


```{r}
state_policy_enrollment_aoi <- state_policy_enrollment %>% filter(State %in% aoi_states)


```

## Number of states enrolled by year
```{r}
for (current_year in 2007:2025){
  print(current_year)
  print(state_policy_enrollment %>% filter(lubridate::year(year) ==  current_year) %>% filter (total_acres >0) %>% pull(State) %>% unique() %>% length())
}
```


## acreage and liaability $ represented by PRF by year

```{r}

### plot for liability and acreage by year
year_prf_enrollment <- lapply(start_year:end_year, function(year){
  print(year)
  data_w_names <- read.csv(here("data/PRF_sob/sobcov_2024_names.csv"))
  column_names <- colnames(data_w_names)
  
  year_2 <- substr(as.character(year), 3, 4)
  data <- read.csv(here(paste0("data/PRF_sob/raw_files/sobcov", year_2, ".txt")),
                   sep = '|')
  names(data) <- column_names
  
  data_prf <- data %>%
    filter(Commodity.Code == 88) %>%
    filter(Insurance.Plan.Code == 13)
  
  # Total liability (all PRF rows)
  total_liability <- sum(data_prf[[grep("Liability.Amount", colnames(data_prf), value = TRUE)[1]]], na.rm = TRUE)
  
  # Total liability across ALL commodities for % calculation
  total_liability_all <- sum(data[[grep("Liability.Amount", colnames(data), value = TRUE)[1]]], na.rm = TRUE)
  
  # Acreage: filter to Acres rows only (collapse whitespace)
  data_prf_acres <- data_prf %>%
    filter(trimws(Quantity.Type) == "Acres")
  
  total_acreage_prf <- sum(data_prf_acres$Net.Reported.Quantity, na.rm = TRUE)
  
  # Total acreage across ALL commodities
  total_acreage_all <- data %>%
    filter(trimws(Quantity.Type) == "Acres") %>%
    pull(Net.Reported.Quantity) %>%
    sum(na.rm = TRUE)
  
  data.frame(
    year = year,
    acreage_prf = total_acreage_prf,
    acreage_all = total_acreage_all,
    pct_liability = total_liability / total_liability_all,
    pct_acreage   = total_acreage_prf / total_acreage_all
  )
  
}) %>% do.call("rbind", .)

year_prf_enrollment[["year"]] <- as.Date(paste0(year_prf_enrollment[["year"]], "-01-01"))

```

## SOB total PRF subsidies

```{r}
### PRF subsidies by state and year — all US states, 2010–2025

subsidy_years <- 2010:2025

prf_subsidy_by_state_year <- lapply(subsidy_years, function(year){
  print(year)

  # Canonical column names (same source used elsewhere in this file)
  data_w_names <- read.csv(here("data/PRF_sob/sobcov_2024_names.csv"))
  column_names <- colnames(data_w_names)

  year_2 <- substr(as.character(year), 3, 4)
  data <- read.csv(here(paste0("data/PRF_sob/raw_files/sobcov", year_2, ".txt")),
                   sep = '|')
  names(data) <- column_names

  # PRF program only: Commodity Code 88, Insurance Plan Code 13
  data_prf <- data %>%
    filter(Commodity.Code == 88) %>%
    filter(Insurance.Plan.Code == 13)

  # Locate the "Subsidy Amount ($)" column robustly.
  # read.csv() sanitizes the header: "Subsidy Amount ($)" -> "Subsidy.Amount...."
  # Anchor with ^ so we don't grab "Additional Subsidy Amount".
  subsidy_col <- grep("^Subsidy\\.Amount", colnames(data_prf), value = TRUE)[1]

  state_sums <- data_prf %>%
    group_by(Location.State.Abbreviation) %>%
    summarize(subsidy = sum(.data[[subsidy_col]], na.rm = TRUE), .groups = "drop")

  state_sums$year <- year
  state_sums

}) %>% do.call("rbind", .)

# Ensure every state-year combination is present (fill missing with 0)
all_states_subsidy <- unique(prf_subsidy_by_state_year$Location.State.Abbreviation)

prf_subsidy_by_state_year <- prf_subsidy_by_state_year %>%
  complete(Location.State.Abbreviation = all_states_subsidy,
           year = subsidy_years,
           fill = list(subsidy = 0)) %>%
  rename(State = Location.State.Abbreviation)

# National annual total across all US states
prf_subsidy_by_year <- prf_subsidy_by_state_year %>%
  group_by(year) %>%
  summarize(total_subsidy = sum(subsidy, na.rm = TRUE), .groups = "drop")

print(prf_subsidy_by_year)

# Grand total across all states and all years, 2010–2025
total_prf_subsidy_2010_2025 <- sum(prf_subsidy_by_state_year$subsidy, na.rm = TRUE)
print(total_prf_subsidy_2010_2025)
```


## Percent grazing by year

```{r}

prf_grazing_pct_by_year <- function(years,
                                     states_filter = aoi_states) {
  
  results <- lapply(years, function(year) {
    
    year_2 <- substr(year, 3, 4)
    tpu_names <- read.csv(here("data/PRF_tpu/tpu_names.csv"))
    tpu <- read.csv(here(paste0("data/PRF_tpu/SOBSCCTPU", year_2, ".TXT")), sep = "|", header = F)
    names(tpu) <- names(tpu_names)
    
    tpu$Type.Name <- trimws(tpu$Type.Name)
    
    # Normalize type names for older years
    if (year <= 2012) {
      tpu[tpu$Type.Name == "Grazingland", "Type.Name"] <- "Grazing"
      tpu[tpu$Type.Name == "Hayland",     "Type.Name"] <- "Haying"
      tpu[tpu$Type.Name == "GRAZINGLAND", "Type.Name"] <- "Grazing"
      tpu[tpu$Type.Name == "HAYLAND",     "Type.Name"] <- "Haying"
    }
    
    tpu$Commodity.Code <- as.numeric(tpu$Commodity.Code)
    
    # Apply same base filters as payment_calcs (excluding use_type filter)
    tpu <- tpu %>%
      filter(State.Abbreviation %in% states_filter) %>%
      filter(Liability.Amount > 0) %>%
      filter(Commodity.Code == 88) %>%
      filter(Type.Name %in% c("Grazing", "Haying"))
    
    # Filter haying to non-organic non-irrigated, same as payment_calcs
    haying_rows <- tpu$Type.Name == "Haying" & year >= 2011
    if (any(haying_rows)) {
      tpu <- tpu %>%
        filter(Type.Name == "Grazing" | 
               (Type.Name == "Haying" & Practice.Code %in% c(525:535, 625:635)))
    }
    
    # Summarise acres by type
    acres_by_type <- tpu %>%
      group_by(Type.Name) %>%
      summarise(total_acres = sum(Net.Reporting.Level.Amount, na.rm = TRUE), .groups = "drop")
    
    total_acres <- sum(acres_by_type$total_acres)
    
    acres_by_type <- acres_by_type %>%
      mutate(
        year         = year,
        total_prf_acres = total_acres,
        pct_of_total = total_acres / total_acres * 100  # placeholder; see below
      )
    
    # Compute per-type percentage of total PRF acres
    acres_by_type$pct_of_total <- acres_by_type$total_acres / total_acres * 100
    
    acres_by_type
  })
  
  # Combine all years
  combined <- bind_rows(results)
  
  # Pivot to a wide summary with one row per year
  summary <- combined %>%
    select(year, Type.Name, total_acres, pct_of_total) %>%
    pivot_wider(
      names_from  = Type.Name,
      values_from = c(total_acres, pct_of_total),
      values_fill = 0
    ) %>%
    rename(
      grazing_acres     = total_acres_Grazing,
      haying_acres      = total_acres_Haying,
      grazing_pct       = pct_of_total_Grazing,
      haying_pct        = pct_of_total_Haying
    ) %>%
    arrange(year)
  
  return(summary)
}

years <- 2007:2025
prf_shares <- prf_grazing_pct_by_year(years)
print(prf_shares)

```

## Coverage level acreage share all years

```{r}

print(sum(prf_shares$grazing_acres)/(sum(prf_shares$grazing_acres) + sum(prf_shares$haying_acres)))


```


## coverage level by year


```{r}
load(here("data/outputs/tpu_all_years.rda"))

#---------------------------
# 1) States of interest
#---------------------------
states_keep <- aoi_states

#---------------------------
# 2) Filter + summarize acres
#---------------------------
acres_by_cov <- tpu_all_years %>%
  filter(State.Abbreviation %in% aoi_states) %>%
  group_by(year, Coverage.Level.Percent) %>%
  summarise(
    acres = sum(Net.Reporting.Level.Amount, na.rm = TRUE),
    .groups = "drop"
  )

#---------------------------
# 3) Compute share of acres
#---------------------------
acres_share <- acres_by_cov %>%
  group_by(year) %>%
  mutate(
    total_year_acres = sum(acres),
    share = acres / total_year_acres
  ) %>%
  ungroup()

#---------------------------
# 4) Plot — Acres by coverage level
#---------------------------
# 4) Plot — Acres by coverage level
#---------------------------
p_acres <- ggplot(acres_by_cov,
                  aes(x = year,
                      y = acres,
                      color = factor(Coverage.Level.Percent),
                      group = Coverage.Level.Percent)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 2.5) +
  scale_y_continuous(
    labels = label_number(scale = 1e-6, suffix = "M", accuracy = 1)
  ) +
  labs(
    title = "Insured Acres by Coverage Level",
    x = NULL,
    y = "Acres (millions)",
    color = "Coverage Level (%)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    axis.title      = element_text(size = 18),
    axis.text       = element_text(size = 16),
    plot.title      = element_text(size = 20, face = "bold"),
    legend.position = "top",
    legend.text     = element_text(size = 15),
    legend.title    = element_text(size = 16),
    legend.key.size = unit(1.4, "lines"),
    panel.grid.minor = element_blank()
  )
#---------------------------
# 5) Plot — Share of acres
#---------------------------
p_share <- ggplot(acres_share,
                  aes(x = year,
                      y = share,
                      fill = factor(Coverage.Level.Percent))) +
  geom_area(position = "stack", alpha = 0.9) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Share of Insured Acres by Coverage Level",
    subtitle = "Nine-State Region",
    x = NULL,
    y = "Share of Acres",
    fill = "Coverage Level (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 14)
  )

#---------------------------
# 6) Print plots
#---------------------------
p_acres
p_share

ggsave(
  filename = here("data/outputs/figures/coverage_level_year.png"),
  plot     = p_acres,
  width    = 9,
  height   = 6.5,
  dpi      = 300,
  bg       = "white"
)

```

## Coverage level acreage share all years

```{r}

acres_by_cov %>%
  group_by(Coverage.Level.Percent) %>%
  summarise(total = sum(acres, na.rm = TRUE)) %>% 
  mutate(percent = 100*(total / sum(total)) )


```

